---
title: "Nuclear cleaning & segmentation"
author: "Mira Sohn"
output:
    html_document:
        code_folding: hide
        df_print: paged
        toc: true
        toc_float: true
        toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE,
                      message=FALSE,
                      cache.lazy=FALSE)
```

Last run: `r date()`

This workflow is designed for exploratory analysis on CE at the single-cell level.


```{r libraries}
# Libraries
library(parallel)
library(tidyverse)
library(ggplot2)
library(reticulate)

# Additional options
set.seed(2570)

# Specify the number of available cores
# n_cores <- future::availableCores()
# options(mc.cores=n_cores)
```


```{r config}

# Path to output directory
out_dir <- "dapi"
if (! dir.exists(out_dir)) { dir.create(out_dir, recursive=TRUE) }

# Load configuration
config <- yaml::read_yaml("config/config.yaml")

# Input & output obj paths
input_obj <- '../../../../../squidpy_masking/scripts/snakemake/results/noperm/post_processing.zarr'
output_obj <- file.path(out_dir, "nuclear_seg.zarr")

# Specify the layers of ImageContainer obj for input and output arrays
sq_layer <- config[['sq_layer']]
out_layer <- config[['out_layer']]

# Specify channel
chan <- config[['channels']][['dapi']]

# Specify connectivity to determine which neighboring pixels are 
# considered connected when forming objects
connectivity <- config[['connectivity']]

# Specify parameters for cleaning
# Debris-control parameter
# (treat the smallest tail of objects as the best estimate of likely noise/speckle scale.)
min_component_quantile <- config[['nuc']][["min_component_quantile"]]
# small-hole filling parameter
# (The 25th percentile is more permissive than the 10th percentile,
# which makes sense because internal holes can be somewhat larger than isolated specks)
area_threshold_quantile <- config[['nuc']][["area_threshold_quantile"]]
# merged-object/artifact flagging parameter
# (isolates the extreme upper tail rather than the center of the size distribution,
# which is where you’d expect merged nuclei or image artifacts to appear)
large_object_quantile <- config[['nuc']][["large_object_quantile"]]
# Safety cap
# (Without it, one or two giant bizarre objects could push the “large object” 
# threshold so high that genuinely suspicious large nuclei would no longer be flagged.)
max_reasonable_area_multiplier <- config[['nuc']][["max_reasonable_area_multiplier"]]
# Sigma for Guanssian filter preceding watershed
dt_smoothing_sigma <- config[['nuc']][['dt_smoothing_sigma']]
# how close two peaks can be before they are treated as one
# (A larger footprint reduces over-splitting; a smaller one increases sensitivity
# to local bumps and may split one nucleus into multiple labels.)
peak_footprint <- config[['nuc']][['peak_footprint']]

# readability and speed of QC visualization
qc_overlay_max_objects <- config[['nuc']][['qc_overlay_max_objects']]

# Additional TRUE/FALSE options
fill_holes <- config[['nuc']][['fill_holes']]
border_clear <- config[['nuc']][['border_clear']]
watershed_for_splitting <- config[['nuc']][['watershed_for_splitting']]

# Load helper functions
source(config[['helper']][['R']])

# Specify conda env
use_condaenv(config[['conda_env']])
source_python(config[['helper']][['py']])
```

```{python ppackages}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from scipy import ndimage as ndi
from skimage import io, measure, morphology, segmentation, feature, color, util
from matplotlib.colors import ListedColormap
from skimage.segmentation import watershed
import dask.array as da
import squidpy as sq
import xarray as xr
import subprocess
import os
```

This workflow performs **nuclear cleaning and segmentation** from 
a **binarized DAPI mask**.

Assumptions:

- the input is already binary or can be safely coerced to binary
- the current file focuses on **nucleus-only segmentation**
- morphology cleanup thresholds should be **estimated from the image**, 
  not fixed arbitrarily
- segmentation outputs will later feed cytoplasmic compartment construction 
  and downstream TDP43 localization analysis

# Loading input `ImageContainer`

```{python load_input}

# ------------------------------------------------------------------------------
# This chunk imports input image saved as zarr in the previous step
# ------------------------------------------------------------------------------

# Load the Zarr store into an ImageContainer
# Set lazy=True to enable Dask-backed lazy loading, which is memory-efficient for large images.
# You can also specify chunks for Dask if needed.
img = sq.im.ImageContainer.load(r.input_obj, lazy=True)

print("Input loaded:\n", r.input_obj)
print("ImageContainer:\n", img)
```

# Summarizing initial input summary

We aim to give a first-pass description of object sizes/shapes in the input

```{python input_summary}

# ------------------------------------------------------------------------------
# This chunk inspects the input image
# ------------------------------------------------------------------------------

# Duplicate a layer in the ImageContainer obj
cleaned_img = img[r.sq_layer]
img.add_img(cleaned_img, layer=r.out_layer)

# Retrieve the Dask array with nuclear staining ([y, x, z, channels])
binary_mask = img[r.out_layer].data[:, :, 0, r.chan]

# Compute input sizes and shapes
initial_labels, initial_df = summarize_components(
    binary_mask,
    connectivity=r.connectivity)
```

Metrics:

- Label: A unique integer assigned to each segmented object so individual nuclei 
  can be identified and tracked separately.
- Area: The number of foreground pixels belonging to a segmented object, 
  representing its size in pixel units.
- Bounding box: The smallest axis-aligned rectangle that fully contains 
  the segmented object.
- Eccentricity: A shape descriptor measuring how elongated an object is, 
  where 0 is circle-like and values closer to 1 are more ellipse-like or stretched.
- Solidity: The fraction of the object’s area relative to the area of its convex hull,
  indicating how solid versus indented or irregular the shape is.
- Extent: The fraction of the bounding box occupied by the object, reflecting how fully 
 the object fills its enclosing rectangle.

```{r input_summary_df}
print(head(py$initial_df))
print(tail(py$initial_df))
```

# Estimating cleanup parameters adaptively {.tabset}

We examine the distribution of connected-component areas from `initial_df`, as summarized below:

```{python cleanup_parameters}

# ------------------------------------------------------------------------------
# This chunk calculates estimate parameters
# ------------------------------------------------------------------------------

# reticulate::repl_python()
param_est = estimate_cleanup_parameters( 
    initial_df,
    min_component_quantile=r.min_component_quantile,
    area_threshold_quantile=r.area_threshold_quantile,
    large_object_quantile=r.large_object_quantile,
    max_reasonable_area_multiplier=r.max_reasonable_area_multiplier)
```

- `min_size`: Lower bound for removing tiny disconnected debris objects
- `area_threshold`: Maximum hole size that may be filled if `fill_holes = TRUE`
- `large_object_cutoff`: Upper-tail threshold used to flag unusually large objects for review
- `median_area`: Useful descriptive summary of the connected-component distribution

Note that the red dashed vertical lines indicate median area. The “after” histogram is shown 
only to make the distribution easier to read after excluding extreme large objects from plotting;
it does not mean those large objects are removed from segmentation.

```{r parameters_df, results='asis'}

# ------------------------------------------------------------------------------
# This chunk visualizes area distribution and median area
# ------------------------------------------------------------------------------

# Print estimated parameters
df <- as.data.frame(py$param_est)
knitr::kable(df)

# Remove objects that are greater than large object cutoff
filtered_df <- py$initial_df %>%
  dplyr::filter(area <= df[['large_object_cutoff']])

df_list <- list(before=py$initial_df, after=filtered_df)

# Visualize data
for (name in names(df_list)) {
    cat('##', name, 'filtering at the large obj cutoff\n\n')
    print(plot_histo(df_list[[name]]))
    cat('\n\n')
}
```

Note that the red dashed vertical lines indicate median area. 


# Cleaning mask

This step performs the following:

- removes very small foreground objects using min_size
- fills small holes using area_threshold if enabled
- optionally removes border-touching objects if `border_clear = TRUE`

```{python cleanup_mask}

# ------------------------------------------------------------------------------
# This chunk cleans the binary mask
# ------------------------------------------------------------------------------

# reticulate::repl_python()
cleaned_mask = clean_input_mask(
    binary_mask,
    min_size=int(param_est["min_size"]),
    area_threshold=int(param_est["area_threshold"]),
    connectivity=r.connectivity,
    fill_holes=bool(r.fill_holes),
    border_clear=bool(r.border_clear)
    )
```

# Segmenting nuclei

This step runs the following if watershed is enabled:

- computes a distance transform on the cleaned mask
- smooths it
- finds local maxima as candidate nucleus centers
- uses marker-controlled watershed to split touching nuclei

We label connected components directly, otherwise.

```{python nuc_seg}

# ------------------------------------------------------------------------------
# This chunk runs segmentation optionally
# ------------------------------------------------------------------------------

if bool(r.watershed_for_splitting):
    nuclear_labels = split_merged_nuclei(
        cleaned_mask,
        sigma=r.dt_smoothing_sigma,
        peak_footprint=r.peak_footprint,
        connectivity=r.connectivity)
else:
    nuclear_labels = measure.label(
        cleaned_mask,
        connectivity=r.connectivity
        )
```

# Building the final feature table

This step computes per-nucleus measurements from the final labeled image. Refer to the following table
for detailed information:

- label
- area
- centroid
- bounding box
- eccentricity
- solidity
- extent
- major/minor axis lengths
- equivalent diameter
- perimeter

The `is_large_object_flag` column is included for QC only. It can help identify large connected 
regions that may represent biologically real extended neuronal structure, merged signal, or 
segmentation artifacts.

```{python feature_df}

# ------------------------------------------------------------------------------
# This chunk runs segmentation optionally
# ------------------------------------------------------------------------------

# Summarize image properties in a data frame
nuclei_df = regionprops_dataframe(nuclear_labels)

# Add QC flags
if nuclei_df.shape[0] > 0:
    large_cutoff = param_est["large_object_cutoff"]
    nuclei_df["is_large_object_flag"] = nuclei_df["area"] >= large_cutoff
else:
    nuclei_df["is_large_object_flag"] = pd.Series(dtype=bool)

print(nuclei_df.head())
print(nuclei_df.tail())

run_summary = {
    "input_shape": tuple(binary_mask.shape),
    "input_foreground_pixels": int(binary_mask.sum()),
    "cleaned_foreground_pixels": int(cleaned_mask.sum()),
    "n_initial_components": int(initial_df.shape[0]),
    "n_final_nuclei": int(nuclei_df.shape[0]),
    "estimated_min_size": int(param_est["min_size"]),
    "estimated_area_threshold": int(param_est["area_threshold"]),
    "estimated_large_object_cutoff": int(param_est["large_object_cutoff"])
}

for key,value in run_summary.items():
    print(key, "=", value)
```

```{python add_to_img}

# ------------------------------------------------------------------------------
# This chunk adds the cleaned masked array back to the ImageContainer obj
# ------------------------------------------------------------------------------

# reticulate::repl_python()
# Add the cleaned mask array to the ImageContainer obj
img[r.out_layer].data[:, :, 0, r.chan] = cleaned_mask

# Delete unnecessary layers
layers_del = [l for l in list(img) if l not in ['image', r.sq_layer, r.out_layer]]
print("The following layers are deleted: ", layers_del)
for l in layers_del:
    del img[l]
```

# Visualizing cleaned nuclear segmentation {.tabset}

- Input: `r sq_layer`
- Output: `r out_layer`

Unique pixels at each array are visualized using the following pseudocolors:

```{python vis_prep}

# ------------------------------------------------------------------------------
# This chunk is used to prepare visualizing processed images
# ----------------------------------------------------------------------------

# Specify a colormap used for unique pixels
color_map = {r.sq_layer: 'dodgerblue', r.out_layer: 'orange'}
for key, value in color_map.items(): 
    print(key, "=", value)

# Specify a function to visualize unique pixels
def display_merge(layer_list):

    # Stack binary images
    for l in layer_list:
        ar = img[l].data[:, :, 0, r.chan]
        plt.imshow(ar,
                   cmap=ListedColormap([(1, 1, 1, 0), color_map[l]]),
                   alpha=0.5)
    plt.xticks([])
    plt.yticks([])
    plt.savefig(f"{r.out_dir}/unique_pixels.png", dpi=600)
    plt.show()
```

```{r vis_images, results='asis'}

# ------------------------------------------------------------------------------
# This chunk visualizes input and output masking
# ------------------------------------------------------------------------------

# Visualize binary images across the channels
for (layer in c(sq_layer, out_layer)) {
    cat("## Channel", chan, "&", layer, "Layer\n\n")
    # Command to show image
    t_deparsed_1 <- paste0("img.show('", layer, "', channel=", chan, ", cmap='gray')")
    t_deparsed_2 <- paste0("plt.savefig('", out_dir, "/", layer, "_channel_", chan,  ".png', dpi=600)")
    t_deparsed_3 <- "plt.close()"
    t_deparsed <- c(t_deparsed_1, t_deparsed_2, t_deparsed_3)
    subchunkify(name=paste0(layer, "_", chan), t_deparsed=t_deparsed)

    link_output(paste0(out_dir, "/", layer, "_channel_", chan, ".png"))
    cat('\n\n')
}

# Overlay images before and after cleaning
cat("## Unique pixels\n\n")
t_deparsed <- paste0("display_merge(list(color_map.keys()))\n")
subchunkify(name="unique_pixels", t_deparsed=t_deparsed)
link_output(paste0(out_dir, "/unique_pixels.png"))
```

# Summary tables

Excel-readable tables are linked in the following table:

- input summary: object statistics before cleanup
- nuclear summary: object statistics after cleanup and connected-component labeling
- parameter summary: adaptive thresholds estimated from the image

```{r summary_tables, results='asis'}

# ------------------------------------------------------------------------------
# This chunk saves and links output summary tables
# ------------------------------------------------------------------------------

df_list <- list(
    input_summary=py$initial_df,
    nuclear_summary=py$nuclei_df,
    parameter_summary=py$param_est
)

path_df <- data.frame(table=names(df_list), link="")

for (name in names(df_list)) {
    tsv <- file.path(out_dir, paste0(name, ".tsv"))
    write.table(df_list[[name]],
                tsv,
                quote=FALSE,
                row.names=FALSE,
                col.names=TRUE)
    path_df[path_df$table == name, 'link'] <- tsv
}

knitr::kable(
    path_df %>%
        mutate(link=paste0("[", link, "](", link, ")"))
)
```

# Saving the `ImageContainer` object

Output images are saved as the following object:

```{python save_zarr}

# ------------------------------------------------------------------------------
# This chunk saves updated ImageContainer as Zarr
# ------------------------------------------------------------------------------

img.save(r.output_obj)
print(r.output_obj)
```



# Session info

```{r session_info, collapse=FALSE}
sessionInfo()
```



