---
title: "Nuclear cleaning & segmentation"
author: "Mira Sohn"
output:
    html_document:
        code_folding: hide
        df_print: paged
        toc: true
        toc_float: true
        toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE,
                      message=FALSE,
                      cache.lazy=FALSE)
```

Last run: `r date()`

```{r libraries}
# Libraries
library(parallel)
library(tidyverse)
library(ggplot2)
library(reticulate)

# Additional options
set.seed(2570)

# Specify the number of available cores
# n_cores <- future::availableCores()
# options(mc.cores=n_cores)
```


```{r config}

# ------------------------------------------------------------------
# Aim of this config section:
# - centralize all workflow settings
# - support future multi-sample use
# - allow scale-aware behavior when image metadata are available
# ------------------------------------------------------------------

# Path to output directory
out_dir <- "dapi"
if (! dir.exists(out_dir)) { dir.create(out_dir, recursive=TRUE) }

# Load configuration
config <- yaml::read_yaml("config/config.yaml")

# Input & output obj paths
input_obj <- '../../../../../squidpy_masking/scripts/snakemake/results/noperm/post_processing.zarr'
output_obj <- file.path(out_dir, "nuclear_seg.zarr")

# Specify the layers of ImageContainer obj for input and output arrays
sq_layer <- config[['sq_layer']]
out_layer <- config[['out_layer']]

# Specify DAPI channel
chan <- config[['channels']][['dapi']]

# Sample / image metadata
# Aim:
# - carry identity and scale metadata into exported tables
# - allow later pooling across many samples
sample_id <- config[['sample']][['sample_id']]
image_id <- config[['sample']][['image_id']]
sample_group <- config[['sample']][['group']]
pixel_size_um <- config[['sample']][['pixel_size_um']]
magnification <- config[['sample']][['magnification']]

# Specify connectivity to determine which neighboring pixels are
# considered connected when forming objects
connectivity <- config[['connectivity']]

# ------------------------------------------------------------------
# Adaptive cleanup parameters
# Aim:
# - estimate cleanup thresholds from the image itself
# - add safety bounds to improve robustness on unusual images
# ------------------------------------------------------------------
min_component_quantile <- config[['nuc']][["min_component_quantile"]]
area_threshold_quantile <- config[['nuc']][["area_threshold_quantile"]]
large_object_quantile <- config[['nuc']][["large_object_quantile"]]
max_reasonable_area_multiplier <- config[['nuc']][["max_reasonable_area_multiplier"]]

min_plausible_nucleus_area_px <- config[['nuc']][["min_plausible_nucleus_area_px"]]
max_plausible_nucleus_area_px <- config[['nuc']][["max_plausible_nucleus_area_px"]]
min_min_size_px <- config[['nuc']][["min_min_size_px"]]
max_min_size_px <- config[['nuc']][["max_min_size_px"]]
min_hole_area_px <- config[['nuc']][["min_hole_area_px"]]
max_hole_area_px <- config[['nuc']][["max_hole_area_px"]]

# ------------------------------------------------------------------
# Watershed splitting parameters
# Aim:
# - allow either manual pixel-based settings or image-adaptive estimation
# ------------------------------------------------------------------
watershed_for_splitting <- config[['nuc']][['watershed_for_splitting']]
watershed_auto_compute <- config[['nuc']][['watershed']][['auto_compute']]

manual_dt_smoothing_sigma <- config[['nuc']][['watershed']][['dt_smoothing_sigma']]
manual_peak_footprint <- config[['nuc']][['watershed']][['peak_footprint']]

sigma_as_fraction_of_radius <- config[['nuc']][['watershed']][['sigma_as_fraction_of_radius']]
peak_footprint_as_fraction_of_diameter <- config[['nuc']][['watershed']][['peak_footprint_as_fraction_of_diameter']]

min_dt_smoothing_sigma <- config[['nuc']][['watershed']][['min_dt_smoothing_sigma']]
max_dt_smoothing_sigma <- config[['nuc']][['watershed']][['max_dt_smoothing_sigma']]
min_peak_footprint <- config[['nuc']][['watershed']][['min_peak_footprint']]
max_peak_footprint <- config[['nuc']][['watershed']][['max_peak_footprint']]

# Readability and speed of QC visualization
qc_overlay_max_objects <- config[['nuc']][['qc_overlay_max_objects']]

# Additional TRUE/FALSE options
fill_holes <- config[['nuc']][['fill_holes']]
border_clear <- config[['nuc']][['border_clear']]

# Load helper functions
source(config[['helper']][['R']])

# Specify conda env
use_condaenv(config[['conda_env']])
source_python(config[['helper']][['py']])
```

```{python ppackages}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from scipy import ndimage as ndi
from skimage import io, measure, morphology, segmentation, feature, color, util
from matplotlib.colors import ListedColormap
from skimage.segmentation import watershed
import dask.array as da
import squidpy as sq
import xarray as xr
import subprocess
import os
```

This workflow performs **nuclear cleaning and segmentation** from 
a **binarized DAPI mask**.

Assumptions:

- the input is already binary or can be safely coerced to binary
- the current file focuses on **nucleus-only segmentation**
- morphology cleanup thresholds should be **estimated from the image**, 
  not fixed arbitrarily
- segmentation outputs will later feed cytoplasmic compartment construction 
  and downstream TDP43 localization analysis
- sample/image metadata may be missing for a demo image, so the workflow must
  still run when `pixel_size_um` is unknown

# Loading input `ImageContainer`

```{python load_input}

# ------------------------------------------------------------------------------
# This chunk imports input image saved as zarr in the previous step
# ------------------------------------------------------------------------------

# Load the Zarr store into an ImageContainer
# Set lazy=True to enable Dask-backed lazy loading, which is memory-efficient for large images.
# You can also specify chunks for Dask if needed.
img = sq.im.ImageContainer.load(r.input_obj, lazy=True)

print("Input loaded:\n", r.input_obj)
print("ImageContainer:\n", img)
```

# Summarizing initial input summary

We aim to give a first-pass description of object sizes/shapes in the input

```{python input_summary}

# ------------------------------------------------------------------------------
# Aim of this chunk:
# - inspect the incoming DAPI mask
# - summarize connected components before cleanup
# - record image-level context for later QC across samples
# ------------------------------------------------------------------------------

# Duplicate a layer in the ImageContainer obj
cleaned_img = img[r.sq_layer]
img.add_img(cleaned_img, layer=r.out_layer)

# Retrieve the Dask array with nuclear staining ([y, x, z, channels])
binary_mask = img[r.out_layer].data[:, :, 0, r.chan]

# Convert explicitly to boolean so downstream morphology behaves consistently
binary_mask = np.asarray(binary_mask).astype(bool)

# Record image-level summary values
image_shape = binary_mask.shape
image_area_px = int(binary_mask.shape[0] * binary_mask.shape[1])
input_foreground_pixels = int(binary_mask.sum())
input_foreground_fraction = (
    float(input_foreground_pixels / image_area_px) if image_area_px > 0 else np.nan
)

# Compute input sizes and shapes
initial_labels, initial_df = summarize_components(
    binary_mask,
    connectivity=r.connectivity
)

print("sample_id =", r.sample_id)
print("image_id =", r.image_id)
print("image_shape =", image_shape)
print("image_area_px =", image_area_px)
print("input_foreground_pixels =", input_foreground_pixels)
print("input_foreground_fraction =", input_foreground_fraction)
```

Metrics:

- Label: A unique integer assigned to each segmented object so individual nuclei 
  can be identified and tracked separately.
- Area: The number of foreground pixels belonging to a segmented object, 
  representing its size in pixel units.
- Bounding box: The smallest axis-aligned rectangle that fully contains 
  the segmented object.
- Eccentricity: A shape descriptor measuring how elongated an object is, 
  where 0 is circle-like and values closer to 1 are more ellipse-like or stretched.
- Solidity: The fraction of the object’s area relative to the area of its convex hull,
  indicating how solid versus indented or irregular the shape is.
- Extent: The fraction of the bounding box occupied by the object, reflecting how fully 
 the object fills its enclosing rectangle.

```{r input_summary_df}
print(head(py$initial_df))
print(tail(py$initial_df))
```

# Estimating cleanup parameters adaptively {.tabset}

The adaptive cleanup strategy is based on the connected-component size
distribution from the current image, but adds two safeguards:

- **plausibility filtering** so very tiny debris or giant artifacts do not dominate
  cleanup estimation
- **bounded auto-estimation** so cleanup and watershed parameters stay within a
  stable range across diverse images

In addition, watershed splitting parameters are now resolved in one of two ways:

- **auto mode**: derived from the estimated nucleus scale in the image
- **manual mode**: uses the explicit pixel-based settings in the config file

```{python cleanup_parameters}

# ------------------------------------------------------------------------------
# Aim of this chunk:
# - estimate cleanup thresholds from the connected-component distribution
# - add guardrails for unusual images
# - estimate scale-aware watershed parameters from the observed nucleus size
# ------------------------------------------------------------------------------

param_est = estimate_cleanup_parameters(
    initial_df,
    min_component_quantile=r.min_component_quantile,
    area_threshold_quantile=r.area_threshold_quantile,
    large_object_quantile=r.large_object_quantile,
    max_reasonable_area_multiplier=r.max_reasonable_area_multiplier,
    min_plausible_nucleus_area_px=r.min_plausible_nucleus_area_px,
    max_plausible_nucleus_area_px=r.max_plausible_nucleus_area_px,
    min_min_size_px=r.min_min_size_px,
    max_min_size_px=r.max_min_size_px,
    min_hole_area_px=r.min_hole_area_px,
    max_hole_area_px=r.max_hole_area_px
)

scale_est = estimate_nuclear_scale_parameters(
    initial_df,
    pixel_size_um=r.pixel_size_um,
    sigma_as_fraction_of_radius=r.sigma_as_fraction_of_radius,
    peak_footprint_as_fraction_of_diameter=r.peak_footprint_as_fraction_of_diameter,
    min_dt_smoothing_sigma=r.min_dt_smoothing_sigma,
    max_dt_smoothing_sigma=r.max_dt_smoothing_sigma,
    min_peak_footprint=r.min_peak_footprint,
    max_peak_footprint=r.max_peak_footprint,
    min_plausible_nucleus_area_px=r.min_plausible_nucleus_area_px,
    max_plausible_nucleus_area_px=r.max_plausible_nucleus_area_px
)

# Resolve the actual watershed parameters used in this run
# Background:
# - auto mode adapts to the image-derived nucleus scale
# - manual mode preserves explicit user control if needed
if bool(r.watershed_auto_compute):
    dt_smoothing_sigma = scale_est["dt_smoothing_sigma"]
    peak_footprint = scale_est["peak_footprint"]
else:
    dt_smoothing_sigma = r.manual_dt_smoothing_sigma
    peak_footprint = r.manual_peak_footprint

print("Resolved watershed sigma =", dt_smoothing_sigma)
print("Resolved peak footprint =", peak_footprint)
```

## Cleanup parameters {.tabset}

Metrics:

- `min_size`: Lower bound for removing tiny disconnected debris objects
- `area_threshold`: Maximum hole size that may be filled if `fill_holes = TRUE`
- `large_object_cutoff`: Upper-tail threshold used to flag unusually large objects for review
- `median_area`: Useful descriptive summary of the connected-component distribution
- `n_components`: Total number of positive connected components detected in the input mask
  before plausibility filtering
- `n_plausible_components`: Number of connected components whose areas fall within the configured 
  plausible nucleus-size range
- `fraction_plausible_components`: Proportion of all detected connected components that fall 
  within the configured plausible nucleus-size range

Note that the red dashed vertical lines indicate median area. The “after” histogram is shown 
only to make the distribution easier to read after excluding extreme large objects from plotting;
it does not mean those large objects are removed from segmentation.

```{r parameters_df, results='asis'}

# ------------------------------------------------------------------------------
# This chunk visualizes area distribution and median area for cleanup parameters
# ------------------------------------------------------------------------------

# Print estimated parameters
df <- as.data.frame(py$param_est) %>%
    gather('metric', 'value')
knitr::kable(df)

# Remove objects that are greater than large object cutoff
filtered_df <- py$initial_df %>%
  dplyr::filter(area <= df['large_object_cutoff', 'value'])

df_list <- list(before=py$initial_df, after=filtered_df)

# Visualize data
xintercept <- df[df$metric == 'median_area', 'value']
for (name in names(df_list)) {
    cat('###', name, 'filtering at the large obj cutoff\n\n')
    print(plot_histo(df_list[[name]], xintercept=xintercept))
    cat('\n\n')
}
```

Note that the red dashed vertical lines indicate median area. 

## Scale parameters

- `estimated_typical_nucleus_area_px`: Median area of plausible connected components,
  used as the workflow’s estimate of a typical nucleus size in pixels.
- `estimated_nucleus_radius_px`: Approximate nucleus radius inferred from `estimated_typical_nucleus_area_px`,
  assuming a roughly circular nucleus.
- `estimated_nucleus_diameter_px`: Approximate nucleus diameter inferred from the estimated 
  typical nucleus area.
- `dt_smoothing_sigma`: Gaussian smoothing width applied to the distance-transform image 
  before watershed peak detection; helps reduce over-splitting from small local bumps.
- `peak_footprint`: Size of the local neighborhood used when detecting distance-transform peaks 
  for watershed markers; larger values are more conservative, smaller values are more sensitive.

```{r scale_df, results='asis'}

# ------------------------------------------------------------------------------
# This chunk visualizes area distribution and median area
# ------------------------------------------------------------------------------

# Print estimated parameters
df <- data.frame(
    unlist(py$scale_est) %>%
        map(~ifelse(is.null(.x), "NULL", .x))
) %>%
gather('metric', 'value')
knitr::kable(df)
```

# Cleaning mask

This step performs the following:

- removes very small foreground objects using min_size
- fills small holes using area_threshold if enabled
- optionally removes border-touching objects if `border_clear = TRUE`

```{python cleanup_mask}

# ------------------------------------------------------------------------------
# Aim of this chunk:
# - apply conservative cleanup to the binary DAPI mask
# - remove tiny disconnected debris
# - fill small internal holes
# - preserve border-touching nuclei unless border_clear is explicitly enabled
# ------------------------------------------------------------------------------

cleaned_mask = clean_input_mask(
    binary_mask,
    min_size=int(param_est["min_size"]),
    area_threshold=int(param_est["area_threshold"]),
    connectivity=r.connectivity,
    fill_holes=bool(r.fill_holes),
    border_clear=bool(r.border_clear)
)
```

# Segmenting nuclei

This step runs the following if watershed is enabled:

- computes a distance transform on the cleaned mask
- smooths it
- finds local maxima as candidate nucleus centers
- uses marker-controlled watershed to split touching nuclei

We label connected components directly, otherwise.

```{python nuc_seg}

# ------------------------------------------------------------------------------
# Aim of this chunk:
# - split merged nuclei using watershed when enabled
# - use resolved scale-aware parameters in auto mode
# - otherwise fall back to plain connected components
# ------------------------------------------------------------------------------

if bool(r.watershed_for_splitting):
    nuclear_labels = split_merged_nuclei(
        cleaned_mask,
        sigma=dt_smoothing_sigma,
        peak_footprint=peak_footprint,
        connectivity=r.connectivity
    )
else:
    nuclear_labels = measure.label(
        cleaned_mask,
        connectivity=r.connectivity
    )
```

# Building the final feature table

This step computes per-nucleus measurements from the final labeled image. Refer to the following table
for detailed information:

- label
- area
- centroid
- bounding box
- eccentricity
- solidity
- extent
- major/minor axis lengths
- equivalent diameter
- perimeter

The `is_large_object_flag` column is included for QC only. It can help identify large connected 
regions that may represent biologically real extended neuronal structure, merged signal, or 
segmentation artifacts.

```{python feature_df}

# ------------------------------------------------------------------------------
# Aim of this chunk:
# - build the per-nucleus feature table
# - add QC flags and metadata for future multi-sample comparison
# - export an image-level summary table for run diagnostics
# ------------------------------------------------------------------------------

# Summarize image properties in a data frame
nuclei_df = regionprops_dataframe(nuclear_labels)

# Add border-touching flag so edge nuclei can be reviewed later without
# forcing their removal up front
nuclei_df = add_border_flag(nuclei_df, nuclear_labels)

# Add QC flags
if nuclei_df.shape[0] > 0:
    large_cutoff = param_est["large_object_cutoff"]
    nuclei_df["is_large_object_flag"] = nuclei_df["area"] >= large_cutoff
else:
    nuclei_df["is_large_object_flag"] = pd.Series(dtype=bool)

# Add sample/image metadata to support future pooled analyses
nuclei_df["sample_id"] = r.sample_id
nuclei_df["image_id"] = r.image_id
nuclei_df["group"] = r.sample_group
nuclei_df["pixel_size_um"] = r.pixel_size_um
nuclei_df["magnification"] = r.magnification

# Record the resolved cleanup / splitting settings used for this image
nuclei_df["estimated_min_size"] = int(param_est["min_size"])
nuclei_df["estimated_area_threshold"] = int(param_est["area_threshold"])
nuclei_df["estimated_large_object_cutoff"] = int(param_est["large_object_cutoff"])
nuclei_df["resolved_dt_smoothing_sigma"] = float(dt_smoothing_sigma)
nuclei_df["resolved_peak_footprint"] = int(peak_footprint)

print(nuclei_df.head())
print(nuclei_df.tail())

run_summary = {
    "sample_id": r.sample_id,
    "image_id": r.image_id,
    "group": r.sample_group,
    "pixel_size_um": r.pixel_size_um,
    "magnification": r.magnification,
    "input_shape": tuple(binary_mask.shape),
    "image_area_px": int(binary_mask.shape[0] * binary_mask.shape[1]),
    "input_foreground_pixels": int(binary_mask.sum()),
    "cleaned_foreground_pixels": int(cleaned_mask.sum()),
    "input_foreground_fraction": float(binary_mask.sum() / (binary_mask.shape[0] * binary_mask.shape[1])),
    "cleaned_foreground_fraction": float(cleaned_mask.sum() / (binary_mask.shape[0] * binary_mask.shape[1])),
    "n_initial_components": int(initial_df.shape[0]),
    "n_final_nuclei": int(nuclei_df.shape[0]),
    "estimated_min_size": int(param_est["min_size"]),
    "estimated_area_threshold": int(param_est["area_threshold"]),
    "estimated_large_object_cutoff": int(param_est["large_object_cutoff"]),
    "estimated_typical_nucleus_area_px": scale_est["estimated_typical_nucleus_area_px"],
    "estimated_nucleus_radius_px": scale_est["estimated_nucleus_radius_px"],
    "estimated_nucleus_diameter_px": scale_est["estimated_nucleus_diameter_px"],
    "resolved_dt_smoothing_sigma": float(dt_smoothing_sigma),
    "resolved_peak_footprint": int(peak_footprint),
    "watershed_auto_compute": bool(r.watershed_auto_compute),
    "border_clear": bool(r.border_clear),
    "fill_holes": bool(r.fill_holes)
}

for key, value in run_summary.items():
    print(key, "=", value)

image_qc_df = pd.DataFrame([run_summary])
scale_parameter_df = pd.DataFrame([scale_est])
```

```{python add_to_img}

# ------------------------------------------------------------------------------
# This chunk adds the cleaned masked array back to the ImageContainer obj
# ------------------------------------------------------------------------------

# reticulate::repl_python()
# Add the cleaned mask array to the ImageContainer obj
img[r.out_layer].data[:, :, 0, r.chan] = cleaned_mask

# Delete unnecessary layers
layers_del = [l for l in list(img) if l not in ['image', r.sq_layer, r.out_layer]]
print("The following layers are deleted: ", layers_del)
for l in layers_del:
    del img[l]
```

# Visualizing cleaned nuclear segmentation {.tabset}

- Input: `r sq_layer`
- Output: `r out_layer`

Unique pixels at each array are visualized using the following pseudocolors:

```{python vis_prep}

# ------------------------------------------------------------------------------
# This chunk is used to prepare visualizing processed images
# ----------------------------------------------------------------------------

# Specify a colormap used for unique pixels
color_map = {r.sq_layer: 'dodgerblue', r.out_layer: 'orange'}
for key, value in color_map.items(): 
    print(key, "=", value)

# Specify a function to visualize unique pixels
def display_merge(layer_list):

    # Stack binary images
    for l in layer_list:
        ar = img[l].data[:, :, 0, r.chan]
        plt.imshow(ar,
                   cmap=ListedColormap([(1, 1, 1, 0), color_map[l]]),
                   alpha=0.5)
    plt.xticks([])
    plt.yticks([])
    plt.savefig(f"{r.out_dir}/unique_pixels.png", dpi=600)
    plt.show()
```

```{r vis_images, results='asis'}

# ------------------------------------------------------------------------------
# This chunk visualizes input and output masking
# ------------------------------------------------------------------------------

# Visualize binary images across the channels
for (layer in c(sq_layer, out_layer)) {
    cat("## Channel", chan, "&", layer, "Layer\n\n")
    # Command to show image
    t_deparsed_1 <- paste0("img.show('", layer, "', channel=", chan, ", cmap='gray')")
    t_deparsed_2 <- paste0("plt.savefig('", out_dir, "/", layer, "_channel_", chan,  ".png', dpi=600)")
    t_deparsed_3 <- "plt.close()"
    t_deparsed <- c(t_deparsed_1, t_deparsed_2, t_deparsed_3)
    subchunkify(name=paste0(layer, "_", chan), t_deparsed=t_deparsed)

    link_output(paste0(out_dir, "/", layer, "_channel_", chan, ".png"))
    cat('\n\n')
}

# Overlay images before and after cleaning
cat("## Unique pixels\n\n")
t_deparsed <- paste0("display_merge(list(color_map.keys()))\n")
subchunkify(name="unique_pixels", t_deparsed=t_deparsed)
link_output(paste0(out_dir, "/unique_pixels.png"))
```

# Summary tables

Excel-readable tables are linked in the following table:

- input summary: object statistics before cleanup
- nuclear summary: object statistics after cleanup and connected-component labeling
- parameter summary: adaptive thresholds estimated from the image

```{r summary_tables, results='asis'}

# ------------------------------------------------------------------------------
# Aim of this chunk:
# - save all object-level and image-level summary tables
# - keep outputs easy to combine across multiple images later
# ------------------------------------------------------------------------------

df_list <- list(
    input_summary = py$initial_df,
    nuclear_summary = py$nuclei_df,
    parameter_summary = as.data.frame(py$param_est),
    image_qc_summary = py$image_qc_df,
    scale_parameter_summary = py$scale_parameter_df
)

path_df <- data.frame(table = names(df_list), link = "")

for (name in names(df_list)) {
    tsv <- file.path(out_dir, paste0(name, ".tsv"))
    df <- df_list[[name]]
    tryCatch({
        write.table(
            df,
            tsv,
            quote = FALSE,
            row.names = FALSE,
            col.names = TRUE,
            sep = "\t"
        )
    }, error = function(e) {
        c_length <- apply(df, 2, class)
        for (c in colnames(df)) {
            if (is.list(df[[c]])) {
                df[[c]] <- df[[c]] %>%
                    paste(sep=", ")
            }
        }
        write.table(
            df,
            tsv,
            quote = FALSE,
            row.names = FALSE,
            col.names = TRUE,
            sep = "\t"
        )
    })

    path_df[path_df$table == name, 'link'] <- tsv
}

knitr::kable(
    path_df %>%
        mutate(link = paste0("[", link, "](", link, ")"))
)
```

# Saving the `ImageContainer` object

Output images are saved as the following object:

```{python save_zarr}

# ------------------------------------------------------------------------------
# This chunk saves updated ImageContainer as Zarr
# ------------------------------------------------------------------------------

img.save(r.output_obj)
print(r.output_obj)
```

# Session info

```{r session_info, collapse=FALSE}
sessionInfo()
```



