---
title: "Nuclear cleaning & segmentation"
author: "Mira Sohn"
output:
    html_document:
        code_folding: hide
        df_print: paged
        toc: true
        toc_float: true
        toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE,
                      message=FALSE,
                      cache.lazy=FALSE)
```

Last run: `r date()`

This workflow is designed for exploratory analysis on CE at the single-cell level.


```{r libraries}
# Libraries
library(parallel)
library(tidyverse)
library(ggplot2)
library(reticulate)

# Additional options
set.seed(2570)

# Specify the number of available cores
# n_cores <- future::availableCores()
# options(mc.cores=n_cores)
```


```{r config}

# Path to output directory
out_dir <- "dapi"
if (! dir.exists(out_dir)) { dir.create(out_dir, recursive=TRUE) }

# Load configuration
config <- yaml::read_yaml("config/config.yaml")

# Load helper functions
source(config[['helper']][['R']])
source_python(config[['helper']][['py']])

# Specify conda env
use_condaenv(config[['conda_env']])
```

```{python ppackages}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from scipy import ndimage as ndi
from skimage import io, measure, morphology, segmentation, feature, color, util
from skimage.segmentation import watershed
import dask.array as da
import squidpy as sq
import xarray as xr
import subprocess
import os
```

This workflow performs **nuclear cleaning and segmentation** from 
a **binarized DAPI mask**.

Assumptions:

- the input is already binary or can be safely coerced to binary
- the current file focuses on **nucleus-only segmentation**
- morphology cleanup thresholds should be **estimated from the image**, 
  not fixed arbitrarily
- segmentation outputs will later feed cytoplasmic compartment construction 
  and downstream TDP43 localization analysis

# Loading input `ImageContainer`

```{python load_input}
# reticulate::repl_python()
stop()
# ------------------------------------------------------------------------------
# This chunk imports input image saved as zarr in the previous step
# ------------------------------------------------------------------------------

# Load the Zarr store into an ImageContainer
# Set lazy=True to enable Dask-backed lazy loading, which is memory-efficient for large images.
# You can also specify chunks for Dask if needed.
img = sq.im.ImageContainer.load(r.config['input_file'], lazy=True)
img = sq.im.ImageContainer.load(r.config['input_file'], lazy=True, chunks=chunksize)

print(f"The following file has been loaded: {input_obj}")
print(img)
```

# Session info

```{r session_info, collapse=FALSE}
sessionInfo()
```



