---
title: "Cytoplasmic cleaning & segmentation"
author: "Mira Sohn"
output:
    html_document:
        code_folding: hide
        df_print: paged
        toc: true
        toc_float: true
        toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE,
                      message=FALSE,
                      cache.lazy=FALSE)
```

Last run: `r date()`

```{r libraries}
# Libraries
library(parallel)
library(tidyverse)
library(ggplot2)
library(reticulate)

# Additional options
set.seed(2570)

# Specify the number of available cores
# n_cores <- future::availableCores()
# options(mc.cores=n_cores)
```

```{r import r_config}
# Load configuration
config <- yaml::read_yaml("config/config.yaml")

# Specify conda env
use_condaenv(config[['conda_env']])
```

```{python ppackages}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from scipy import ndimage as ndi
from skimage import io, measure, morphology, segmentation
from matplotlib.colors import ListedColormap
import dask.array as da
import squidpy as sq
import xarray as xr
import subprocess
import os
```

```{r config}

# Path to output directory
out_dir <- "map2"
if (! dir.exists(out_dir)) { dir.create(out_dir, recursive=TRUE) }

# Input & output obj paths
input_obj <- "dapi/nuclear_seg.zarr"
output_obj <- file.path(out_dir, "cytoplasmic_seg.zarr")

# Specify the layers of ImageContainer obj for input and output arrays
sq_layer <- config[['sq_layer']]
out_layer <- config[['out_layer']]

# Specify channel
cyt_chan <- config[['channels']][['map2']]

# Specify connectivity to determine which neighboring pixels are 
# considered connected when forming objects
connectivity <- config[['connectivity']]

# Sample/image metadata propagated to outputs
sample_id <- config[["sample"]][["sample_id"]]
image_id <- config[["sample"]][["image_id"]]
group <- config[["sample"]][["group"]]
pixel_size_um <- config[["sample"]][["pixel_size_um"]]
magnification <- config[["sample"]][["magnification"]]

# Cytoplasmic adaptive cleanup settings
min_component_quantile <- config[["cyt"]][["min_component_quantile"]]
area_threshold_quantile <- config[["cyt"]][["area_threshold_quantile"]]
large_object_quantile <- config[["cyt"]][["large_object_quantile"]]
max_reasonable_area_multiplier <- config[["cyt"]][["max_reasonable_area_multiplier"]]

# Cytoplasmic plausibility guardrails
min_plausible_cyt_area_px <- config[["cyt"]][["min_plausible_cyt_area_px"]]
max_plausible_cyt_area_px <- config[["cyt"]][["max_plausible_cyt_area_px"]]
min_min_size_px <- config[["cyt"]][["min_min_size_px"]]
max_min_size_px <- config[["cyt"]][["max_min_size_px"]]
min_hole_area_px <- config[["cyt"]][["min_hole_area_px"]]
max_hole_area_px <- config[["cyt"]][["max_hole_area_px"]]

# Optional cleanup behaviors
fill_holes <- config[["cyt"]][["fill_holes"]]
border_clear <- config[["cyt"]][["border_clear"]]
apply_opening <- config[["cyt"]][["apply_opening"]]
opening_radius <- config[["cyt"]][["opening_radius"]]

# QC / output behavior
qc_overlay_max_objects <- config[["cyt"]][["qc_overlay_max_objects"]]
save_intermediates <- config[["output"]][["save_intermediates"]]
save_qc_plots <- config[["output"]][["save_qc_plots"]]

# Load python helper functions
source(config[['helper']][['R']])
source_python(config[['helper']][['py']])
```

This workflow performs **cytoplasmic cleaning and segmentation** from a **binarized MAP2 mask**.

Assumptions:

- the input is already binary or can be safely coerced to binary
- the current file focuses on **cytoplasmic segmentation** using MAP2 immunofluorescence (IF) staining
- morphology cleanup thresholds should be **estimated from the image**, 
  not fixed arbitrarily
- segmentation outputs will later feed downstream TDP43 localization analysis

# Loading input `ImageContainer`

```{python load_input}

# ------------------------------------------------------------------------------
# This chunk imports input image saved as zarr in the previous step
# ------------------------------------------------------------------------------

# reticulate::repl_python()

# Load the Zarr store into an ImageContainer
# Set lazy=True to enable Dask-backed lazy loading, which is memory-efficient for large images.
# You can also specify chunks for Dask if needed.
img = sq.im.ImageContainer.load(r.input_obj, lazy=True)

print("Input loaded:\n", r.input_obj)
print("ImageContainer:\n", img)
```

```{python input_summary}

# ------------------------------------------------------------------------------
# This chunk inspects the input image
# ------------------------------------------------------------------------------

# Retrieve the Dask arrays for channels of interest ([y, x, z, channels])
cyt = img[r.out_layer].data[:, :, 0, r.cyt_chan]
cyt = np.asarray(cyt > 0).astype(bool)
```

Metrics:

- `image_area_px`: total number of pixels in the 2D image plane.
- `input_foreground_pixels`: number of MAP2-positive pixels before cleanup.
- `input_foreground_fraction`: fraction of the image occupied by MAP2-positive signal before cleanup.
- `label`: unique integer identifier for each connected object.
- `area`: number of pixels in an object.
- `bbox`: smallest axis-aligned rectangle enclosing the object.
- `centroid`: object center of mass in pixel coordinates.
- `eccentricity`: elongation measure; values near 0 are rounder, values near 1 are more elongated.
- `solidity`: object area divided by convex hull area; lower values suggest more irregular or indented shapes.
- `extent`: fraction of the bounding box occupied by the object.
- `major_axis_length / minor_axis_length`: ellipse-based axis lengths describing object size and elongation.
- `equivalent_diameter_area`: diameter of a circle with the same area as the object.
- `perimeter`: approximate object boundary length in pixels.
- `touches_border`: whether the object touches the image edge.
- `is_large_object_flag`: whether the object area exceeds the image-adaptive large-object QC cutoff.


# Summarizing initial input summary

We aim to give a first-pass description of object sizes/shapes in the input

```{python processed_input_summary}

# ------------------------------------------------------------------------------
# This chunk quantify the size and shape distribution of MAP2-positive objects before cleanup
# and create the empirical basis for adaptive threshold estimation
# ------------------------------------------------------------------------------

image_shape = tuple(cyt.shape)
image_area_px = int(cyt.size)
input_foreground_pixels = int(cyt.sum())
input_foreground_fraction = float(input_foreground_pixels / image_area_px) if image_area_px > 0 else np.nan

print("image_shape =", image_shape)
print("image_area_px =", image_area_px)
print("input_foreground_pixels =", input_foreground_pixels)
print("input_foreground_fraction =", input_foreground_fraction)

initial_labels, initial_df = summarize_components(
    cyt,
    connectivity=r.connectivity
)

# Append the border-touching QC flag
initial_df = add_border_flag(initial_df, initial_labels)
initial_df = append_sample_metadata(
    initial_df,
    sample_id=r.sample_id,
    image_id=r.image_id,
    group=r.group,
    pixel_size_um=r.pixel_size_um,
    magnification=r.magnification
)
```

```{r input_summary_df}
print(head(py$initial_df))
print(tail(py$initial_df))
```

# Estimating cleanup parameters adaptively {.tabset}

We examine the distribution of connected-component areas from `initial_df`, as summarized below:

```{python cleanup_parameters}

# ------------------------------------------------------------------------------
# This chunk estimates conservative cleanup thresholds from the current image and keeps
# thresholds adaptive while preventing extreme sparse/artifact-heavy images from destabilizing them
# ------------------------------------------------------------------------------

# reticulate::repl_python()
param_est = estimate_cleanup_parameters(
    initial_df,
    min_component_quantile=r.min_component_quantile,
    area_threshold_quantile=r.area_threshold_quantile,
    large_object_quantile=r.large_object_quantile,
    max_reasonable_area_multiplier=r.max_reasonable_area_multiplier,
    min_plausible_object_area_px=r.min_plausible_cyt_area_px,
    max_plausible_object_area_px=r.max_plausible_cyt_area_px,
    min_min_size_px=r.min_min_size_px,
    max_min_size_px=r.max_min_size_px,
    min_hole_area_px=r.min_hole_area_px,
    max_hole_area_px=r.max_hole_area_px
)

param_df = pd.DataFrame([param_est])
param_df = append_sample_metadata(
    param_df,
    sample_id=r.sample_id,
    image_id=r.image_id,
    group=r.group,
    pixel_size_um=r.pixel_size_um,
    magnification=r.magnification
)
```

Parameters:

- `min_size`: smallest object size retained during debris removal.
- `area_threshold`: largest internal hole size that will be filled when hole filling is enabled.
- `large_object_cutoff`: image-adaptive QC threshold used to flag unusually large MAP2 objects.
- `median_area`: median area of the plausibility-filtered object distribution used for stable thresholding.
- `n_components`: number of connected components in the original mask.
- `n_plausible_components`: number of objects falling inside the configured plausible size range.
- `fraction_plausible_components`: fraction of all objects considered plausible during adaptive estimation.

The distribution of object area is explored using histograms, before and after removing large objects.

```{r parameters_df, results='asis'}

# ------------------------------------------------------------------------------
# This chunk visualizes area distribution and median area
# ------------------------------------------------------------------------------

# Print estimated parameters
df <- as.data.frame(py$param_est) %>%
    gather('metric', 'value')
knitr::kable(df)

# Remove objects that are greater than large object cutoff
filtered_df <- py$initial_df %>%
  dplyr::filter(area <= df[df$metric == 'large_object_cutoff', 'value'])

df_list <- list(before=py$initial_df, after=filtered_df)

# Visualize data
xintercept <- df[df$metric == 'median_area', 'value']
for (name in names(df_list)) {
    cat('###', name, 'filtering at the large obj cutoff\n\n')
    print(plot_histo(df_list[[name]], xintercept=xintercept))
    cat('\n\n')
}
```

Note that the red dashed vertical lines indicate median area. The “after” histogram is shown 
only to make the distribution easier to read after excluding extreme large objects from plotting;
it does not mean those large objects are removed from segmentation.
