---
title: "Cytoplasmic cleaning & segmentation"
author: "Mira Sohn"
output:
    html_document:
        code_folding: hide
        df_print: paged
        toc: true
        toc_float: true
        toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE,
                      message=FALSE,
                      cache.lazy=FALSE)
```

Last run: `r date()`

This workflow is designed for exploratory analysis on CE at the single-cell level.


```{r libraries}
# Libraries
library(parallel)
library(tidyverse)
library(ggplot2)
library(reticulate)

# Additional options
set.seed(2570)

# Specify the number of available cores
# n_cores <- future::availableCores()
# options(mc.cores=n_cores)
```


```{r config}

# Path to output directory
out_dir <- "map2"
if (! dir.exists(out_dir)) { dir.create(out_dir, recursive=TRUE) }

# Load configuration
config <- yaml::read_yaml("config/config.yaml")

# Input & output obj paths
input_obj <- "dapi/nuclear_seg.zarr"
output_obj <- file.path(out_dir, "cytoplasmic_seg.zarr")

# Specify the layers of ImageContainer obj for input and output arrays
sq_layer <- config[['sq_layer']]
out_layer <- config[['out_layer']]

# Specify channel
cyt_chan <- config[['channels']][['map2']]

# Specify connectivity to determine which neighboring pixels are 
# considered connected when forming objects
connectivity <- config[['connectivity']]

# Specify parameters for cleaning
# Debris-control parameter
min_component_quantile <- config[['cyt']][['min_component_quantile']]

# Small-hole filling parameter
area_threshold_quantile <- config[['cyt']][['area_threshold_quantile']]

# Upper-tail cutoff to flag excessively large objects
large_object_quantile <- config[['cyt']][['large_object_quantile']]

# Safety cap for large-object threshold
max_reasonable_area_multiplier <- config[['cyt']][['max_reasonable_area_multiplier']]

# Readability and speed of QC visualization
qc_overlay_max_objects <- config[['cyto']][['qc_overlay_max_objects']]

# Additional TRUE/FALSE options
fill_holes <- config[['cyt']][['fill_holes']]
border_clear <- config[['cyt']][['border_clear']]

# Load helper functions
source(config[['helper']][['R']])

# Specify conda env
use_condaenv(config[['conda_env']])
source_python(config[['helper']][['py']])
```

```{python ppackages}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from scipy import ndimage as ndi
from skimage import io, measure, morphology, segmentation, feature, color, util
from matplotlib.colors import ListedColormap
from skimage.segmentation import watershed
import dask.array as da
import squidpy as sq
import xarray as xr
import subprocess
import os
```

This workflow performs **cytoplasmic cleaning and segmentation** from 
a **binarized MAP2 mask**.

Assumptions:

- the input is already binary or can be safely coerced to binary
- the current file focuses on **cytoplasmic segmentation** using MAP2 immunofluorescence (IF) staining
- morphology cleanup thresholds should be **estimated from the image**, 
  not fixed arbitrarily
- segmentation outputs will later feed downstream TDP43 localization analysis

# Loading input `ImageContainer`

```{python load_input}

# ------------------------------------------------------------------------------
# This chunk imports input image saved as zarr in the previous step
# ------------------------------------------------------------------------------

# reticulate::repl_python()

# Load the Zarr store into an ImageContainer
# Set lazy=True to enable Dask-backed lazy loading, which is memory-efficient for large images.
# You can also specify chunks for Dask if needed.
img = sq.im.ImageContainer.load(r.input_obj, lazy=True)

print("Input loaded:\n", r.input_obj)
print("ImageContainer:\n", img)
```

# Summarizing initial input summary

We aim to give a first-pass description of object sizes/shapes in the input

```{python input_summary}

# ------------------------------------------------------------------------------
# This chunk inspects the input image
# ------------------------------------------------------------------------------

# Retrieve the Dask arrays for channels of interest ([y, x, z, channels])
cyt = img[r.out_layer].data[:, :, 0, r.cyt_chan]
cyt = cyt > 0

# Compute input sizes and shapes
initial_labels, initial_df = summarize_components(
    cyt,
    connectivity=r.connectivity)
```

Metrics:

- Label: A unique integer assigned to each segmented object so individual nuclei 
  can be identified and tracked separately.
- Area: The number of foreground pixels belonging to a segmented object, 
  representing its size in pixel units.
- Bounding box: The smallest axis-aligned rectangle that fully contains 
  the segmented object.
- Eccentricity: A shape descriptor measuring how elongated an object is, 
  where 0 is circle-like and values closer to 1 are more ellipse-like or stretched.
- Solidity: The fraction of the object’s area relative to the area of its convex hull,
  indicating how solid versus indented or irregular the shape is.
- Extent: The fraction of the bounding box occupied by the object, reflecting how fully 
 the object fills its enclosing rectangle.

```{r input_summary_df}
print(head(py$initial_df))
print(tail(py$initial_df))
```

# Estimating cleanup parameters adaptively {.tabset}

We examine the distribution of connected-component areas from `initial_df`, as summarized below:

```{python cleanup_parameters}

# ------------------------------------------------------------------------------
# This chunk calculates estimate parameters
# ------------------------------------------------------------------------------

# reticulate::repl_python()
param_est = estimate_cleanup_parameters(
    initial_df,
    min_component_quantile=r.min_component_quantile,
    area_threshold_quantile=r.area_threshold_quantile,
    large_object_quantile=r.large_object_quantile,
    max_reasonable_area_multiplier=r.max_reasonable_area_multiplier)
```

- `min_size`: Lower bound for removing tiny disconnected debris objects
- `area_threshold`: Maximum hole size that may be filled if `fill_holes = TRUE`
- `large_object_cutoff`: Upper-tail threshold used to flag unusually large objects for review
- `median_area`: Useful descriptive summary of the connected-component distribution

The distribution of object area is explored using histograms, before and after removing large objects.

```{r parameters_df, results='asis'}

# ------------------------------------------------------------------------------
# This chunk visualizes area distribution and median area
# ------------------------------------------------------------------------------

# Print estimated parameters
df <- as.data.frame(py$param_est)
knitr::kable(df)

# Remove objects that are greater than large object cutoff
filtered_df <- py$initial_df %>%
  dplyr::filter(area <= df[['large_object_cutoff']])

df_list <- list(before=py$initial_df, after=filtered_df)

# Visualize data
for (name in names(df_list)) {
    cat('##', name, 'filtering at the large obj cutoff\n\n')
    print(plot_histo(df_list[[name]]))
    cat('\n\n')
}
```

Note that the red dashed vertical lines indicate median area. The “after” histogram is shown 
only to make the distribution easier to read after excluding extreme large objects from plotting;
it does not mean those large objects are removed from segmentation.

# Cleaning mask

This step performs the following:

- removes very small foreground objects using min_size
- fills small holes using area_threshold if enabled
- optionally removes border-touching objects if `border_clear = TRUE`
- optionally applies morphological opening to reduce thin protrusions
- optionally suppresses very thin branch-like foreground signal

In the current workflow, these cleanup steps are intended to reduce non-somatic signal 
while preserving larger compact regions.


```{python clean_mask}

# ------------------------------------------------------------------------------
# This chunk cleans the binary mask
# ------------------------------------------------------------------------------

cleaned_mask = clean_input_mask(
    cyt,
    min_size=int(param_est["min_size"]),
    area_threshold=int(param_est["area_threshold"]),
    connectivity=r.connectivity,
    fill_holes=bool(r.fill_holes),
    border_clear=bool(r.border_clear)
)
```

# Segmenting cytoplasmic regions

This step labels connected MAP2-positive objects after cleanup.

```{python label_cyto}

# ------------------------------------------------------------------------------
# This chunk labels cleaned cytoplasmic objects
# ------------------------------------------------------------------------------

cyto_labels = measure.label(
    cleaned_mask,
    connectivity=r.connectivity
)

print(cyto_labels.head())
print(cyto_labels.tail())
```

# Building the final feature table

This step computes region-level measurements from the final labeled image. Refer to the following table
for detailed information:

- label
- area
- centroid
- bounding box
- eccentricity
- solidity
- extent
- major/minor axis lengths
- equivalent diameter
- perimeter

The `is_large_object_flag` column is included for QC only. It can help identify large connected 
regions that may represent biologically real extended neuronal structure, merged signal, or 
segmentation artifacts.

```{python feature_df}

# ------------------------------------------------------------------------------
# This chunk summarizes final MAP2-positive segmented objects
# ------------------------------------------------------------------------------

cytoplasm_df = regionprops_dataframe(cyto_labels)

if cytoplasm_df.shape[0] > 0:
    large_cutoff = param_est["large_object_cutoff"]
    cytoplasm_df["is_large_object_flag"] = cytoplasm_df["area"] >= large_cutoff
else:
    cytoplasm_df["is_large_object_flag"] = pd.Series(dtype=bool)

print(cytoplasm_df.head())
print(cytoplasm_df.tail())

run_summary = {
    "input_shape": tuple(cyt.shape),
    "input_foreground_pixels": int(cyt.sum()),
    "cleaned_foreground_pixels": int(cleaned_mask.sum()),
    "n_initial_components": int(initial_df.shape[0]),
    "n_final_cyto_components": int(cytoplasm_df.shape[0]),
    "estimated_min_size": int(param_est["min_size"]),
    "estimated_area_threshold": int(param_est["area_threshold"]),
    "estimated_large_object_cutoff": int(param_est["large_object_cutoff"])
}

for key,value in run_summary.items():
    print(key, "=", value)
```

```{python add_to_img}

# ------------------------------------------------------------------------------
# This chunk adds the cleaned masked array back to the ImageContainer obj
# ------------------------------------------------------------------------------

# reticulate::repl_python()
# Add the cleaned mask array to the ImageContainer obj
img[r.out_layer].data[:, :, 0, r.cyt_chan] = cleaned_mask

# Delete unnecessary layers
layers_del = [l for l in list(img) if l not in ['image', r.sq_layer, r.out_layer]]
print("The following layers are deleted: ", layers_del)
for l in layers_del:
    del img[l]
```

# Visualizing cleaned cytoplasmic segmentation {.tabset}

- Input: `r sq_layer`
- Output: `r out_layer`

Unique pixels at each array are visualized using the following pseudocolors:

```{python vis_prep}

# ------------------------------------------------------------------------------
# This chunk is used to prepare visualizing processed images
# ----------------------------------------------------------------------------

# Specify a colormap used for unique pixels
color_map = {r.sq_layer: 'dodgerblue', r.out_layer: 'orange'}
for key, value in color_map.items(): 
    print(key, "=", value)

# Specify a function to visualize unique pixels
def display_merge(layer_list):

    # Stack binary images
    for l in layer_list:
        ar = img[l].data[:, :, 0, r.cyt_chan]
        plt.imshow(ar,
                   cmap=ListedColormap([(1, 1, 1, 0), color_map[l]]),
                   alpha=0.5)
    plt.xticks([])
    plt.yticks([])
    plt.savefig(f"{r.out_dir}/unique_pixels.png", dpi=600)
    plt.show()
```

```{r vis_images, results='asis'}

# ------------------------------------------------------------------------------
# This chunk visualizes input and output masking
# ------------------------------------------------------------------------------

# Visualize binary images across the channels
for (layer in c(sq_layer, out_layer)) {
    cat("## Channel", cyt_chan, "&", layer, "Layer\n\n")
    # Command to show image
    t_deparsed_1 <- paste0("img.show('", layer, "', channel=", cyt_chan, ", cmap='gray')")
    t_deparsed_2 <- paste0("plt.savefig('", out_dir, "/", layer, "_channel_", cyt_chan,  ".png', dpi=600)")
    t_deparsed_3 <- "plt.close()"
    t_deparsed <- c(t_deparsed_1, t_deparsed_2, t_deparsed_3)
    subchunkify(name=paste0(layer, "_", cyt_chan), t_deparsed=t_deparsed)

    link_output(paste0(out_dir, "/", layer, "_channel_", cyt_chan, ".png"))
    cat('\n\n')
}

# Overlay images before and after cleaning
cat("## Unique pixels\n\n")
t_deparsed <- paste0("display_merge(list(color_map.keys()))\n")
subchunkify(name="unique_pixels", t_deparsed=t_deparsed)
link_output(paste0(out_dir, "/unique_pixels.png"))
```


# Summary tables

Excel-readable tables are linked in the following table:

- input summary: object statistics before cleanup
- cytoplasmic summary: object statistics after cleanup and connected-component labeling
- parameter summary: adaptive thresholds estimated from the image

```{r summary_tables, results='asis'}

# ------------------------------------------------------------------------------
# This chunk saves and links output summary tables
# ------------------------------------------------------------------------------

df_list <- list(
    input_summary=py$initial_df,
    cytoplasmic_summary=py$cytoplasmic_df,
    parameter_summary=py$param_est
)

path_df <- data.frame(table=names(df_list), link="")

for (name in names(df_list)) {
    tsv <- file.path(out_dir, paste0(name, ".tsv"))
    write.table(df_list[[name]],
                tsv,
                quote=FALSE,
                row.names=FALSE,
                col.names=TRUE)
    path_df[path_df$table == name, 'link'] <- tsv
}

knitr::kable(
    path_df %>%
        mutate(link=paste0("[", link, "](", link, ")"))
)
```

# Saving the `ImageContainer` object

Output images are saved as the following object:

```{python save_zarr}

# ------------------------------------------------------------------------------
# This chunk saves updated ImageContainer as Zarr
# ------------------------------------------------------------------------------

img.save(r.output_obj)
print(r.output_obj)
```



# Session info

```{r session_info, collapse=FALSE}
sessionInfo()
```



