---
title: "Cytoplasmic cleaning & segmentation"
author: "Mira Sohn"
output:
    html_document:
        code_folding: hide
        df_print: paged
        toc: true
        toc_float: true
        toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE,
                      message=FALSE,
                      cache.lazy=FALSE)
```

Last run: `r date()`

```{r libraries}
# Libraries
library(parallel)
library(tidyverse)
library(ggplot2)
library(reticulate)

# Additional options
set.seed(2570)

# Specify the number of available cores
# n_cores <- future::availableCores()
# options(mc.cores=n_cores)
```

```{r import r_config}
# Load configuration
config <- yaml::read_yaml("config/config.yaml")

# Specify conda env
use_condaenv(config[['conda_env']])
```

```{python ppackages}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from scipy import ndimage as ndi
from skimage import io, measure, morphology, segmentation
from matplotlib.colors import ListedColormap
import dask.array as da
import squidpy as sq
import xarray as xr
import subprocess
import os
```

```{r config}

# Path to output directory
out_dir <- "map2"
if (! dir.exists(out_dir)) { dir.create(out_dir, recursive=TRUE) }

# Input & output obj paths
input_obj <- "dapi/nuclear_seg.zarr"
output_obj <- file.path(out_dir, "cytoplasmic_seg.zarr")

# Specify the layers of ImageContainer obj for input and output arrays
sq_layer <- config[['sq_layer']]
out_layer <- config[['out_layer']]

# Specify channel
cyt_chan <- config[['channels']][['map2']]

# Specify connectivity to determine which neighboring pixels are 
# considered connected when forming objects
connectivity <- config[['connectivity']]

# Sample/image metadata propagated to outputs
sample_id <- config[["sample"]][["sample_id"]]
image_id <- config[["sample"]][["image_id"]]
group <- config[["sample"]][["group"]]
pixel_size_um <- config[["sample"]][["pixel_size_um"]]
magnification <- config[["sample"]][["magnification"]]

# Cytoplasmic adaptive cleanup settings
min_component_quantile <- config[["cyt"]][["min_component_quantile"]]
area_threshold_quantile <- config[["cyt"]][["area_threshold_quantile"]]
large_object_quantile <- config[["cyt"]][["large_object_quantile"]]
max_reasonable_area_multiplier <- config[["cyt"]][["max_reasonable_area_multiplier"]]

# Cytoplasmic plausibility guardrails
min_plausible_cyt_area_px <- config[["cyt"]][["min_plausible_cyt_area_px"]]
max_plausible_cyt_area_px <- config[["cyt"]][["max_plausible_cyt_area_px"]]
min_min_size_px <- config[["cyt"]][["min_min_size_px"]]
max_min_size_px <- config[["cyt"]][["max_min_size_px"]]
min_hole_area_px <- config[["cyt"]][["min_hole_area_px"]]
max_hole_area_px <- config[["cyt"]][["max_hole_area_px"]]

# Optional cleanup behaviors
fill_holes <- config[["cyt"]][["fill_holes"]]
border_clear <- config[["cyt"]][["border_clear"]]
apply_opening <- config[["cyt"]][["apply_opening"]]
opening_radius <- config[["cyt"]][["opening_radius"]]

# QC / output behavior
qc_overlay_max_objects <- config[["cyt"]][["qc_overlay_max_objects"]]
save_intermediates <- config[["output"]][["save_intermediates"]]
save_qc_plots <- config[["output"]][["save_qc_plots"]]

# Load python helper functions
source(config[['helper']][['R']])
source_python(config[['helper']][['py']])
```

This workflow performs **cytoplasmic cleaning and segmentation** from a **binarized MAP2 mask**.

Assumptions:

- the input is already binary or can be safely coerced to binary
- the current file focuses on **cytoplasmic segmentation** using MAP2 immunofluorescence (IF) staining
- morphology cleanup thresholds should be **estimated from the image**, 
  not fixed arbitrarily
- segmentation outputs will later feed downstream TDP43 localization analysis

# Loading input `ImageContainer`

```{python load_input}

# ------------------------------------------------------------------------------
# This chunk imports input image saved as zarr in the previous step
# ------------------------------------------------------------------------------

# reticulate::repl_python()

# Load the Zarr store into an ImageContainer
# Set lazy=True to enable Dask-backed lazy loading, which is memory-efficient for large images.
# You can also specify chunks for Dask if needed.
img = sq.im.ImageContainer.load(r.input_obj, lazy=True)

print("Input loaded:\n", r.input_obj)
print("ImageContainer:\n", img)
```

```{python input_summary}

# ------------------------------------------------------------------------------
# This chunk inspects the input image
# ------------------------------------------------------------------------------

# Retrieve the Dask arrays for channels of interest ([y, x, z, channels])
cyt = img[r.out_layer].data[:, :, 0, r.cyt_chan]
cyt = np.asarray(cyt > 0).astype(bool)
```

Metrics:

- `image_area_px`: total number of pixels in the 2D image plane.
- `input_foreground_pixels`: number of MAP2-positive pixels before cleanup.
- `input_foreground_fraction`: fraction of the image occupied by MAP2-positive signal before cleanup.
- `label`: unique integer identifier for each connected object.
- `area`: number of pixels in an object.
- `bbox`: smallest axis-aligned rectangle enclosing the object.
- `centroid`: object center of mass in pixel coordinates.
- `eccentricity`: elongation measure; values near 0 are rounder, values near 1 are more elongated.
- `solidity`: object area divided by convex hull area; lower values suggest more irregular or indented shapes.
- `extent`: fraction of the bounding box occupied by the object.
- `major_axis_length / minor_axis_length`: ellipse-based axis lengths describing object size and elongation.
- `equivalent_diameter_area`: diameter of a circle with the same area as the object.
- `perimeter`: approximate object boundary length in pixels.
- `touches_border`: whether the object touches the image edge.
- `is_large_object_flag`: whether the object area exceeds the image-adaptive large-object QC cutoff.


# Summarizing initial input summary

This chunk measures MAP2-positive objects before cleanup so the size distribution can be used 
to estimate adaptive thresholds and document starting image quality.

```{python processed_input_summary}

# ------------------------------------------------------------------------------
# This chunk quantify the size and shape distribution of MAP2-positive objects before cleanup
# and create the empirical basis for adaptive threshold estimation
# ------------------------------------------------------------------------------

image_shape = tuple(cyt.shape)
image_area_px = int(cyt.size)
input_foreground_pixels = int(cyt.sum())
input_foreground_fraction = float(input_foreground_pixels / image_area_px) if image_area_px > 0 else np.nan

print("image_shape =", image_shape)
print("image_area_px =", image_area_px)
print("input_foreground_pixels =", input_foreground_pixels)
print("input_foreground_fraction =", input_foreground_fraction)

initial_labels, initial_df = summarize_components(
    cyt,
    connectivity=r.connectivity
)

# Append the border-touching QC flag
initial_df = add_border_flag(initial_df, initial_labels)
initial_df = append_sample_metadata(
    initial_df,
    sample_id=r.sample_id,
    image_id=r.image_id,
    group=r.group,
    pixel_size_um=r.pixel_size_um,
    magnification=r.magnification
)
```

```{r input_summary_df}
print(head(py$initial_df))
print(tail(py$initial_df))
```

# Estimating cleanup parameters adaptively {.tabset}

We examine the distribution of connected-component areas from `initial_df`, as summarized below:

```{python cleanup_parameters}

# ------------------------------------------------------------------------------
# This chunk estimates conservative cleanup thresholds from the current image and keeps
# thresholds adaptive while preventing extreme sparse/artifact-heavy images from destabilizing them
# ------------------------------------------------------------------------------

# reticulate::repl_python()
param_est = estimate_cleanup_parameters(
    initial_df,
    min_component_quantile=r.min_component_quantile,
    area_threshold_quantile=r.area_threshold_quantile,
    large_object_quantile=r.large_object_quantile,
    max_reasonable_area_multiplier=r.max_reasonable_area_multiplier,
    min_plausible_object_area_px=r.min_plausible_cyt_area_px,
    max_plausible_object_area_px=r.max_plausible_cyt_area_px,
    min_min_size_px=r.min_min_size_px,
    max_min_size_px=r.max_min_size_px,
    min_hole_area_px=r.min_hole_area_px,
    max_hole_area_px=r.max_hole_area_px
)

param_df = pd.DataFrame([param_est])
param_df = append_sample_metadata(
    param_df,
    sample_id=r.sample_id,
    image_id=r.image_id,
    group=r.group,
    pixel_size_um=r.pixel_size_um,
    magnification=r.magnification
)
```

Parameters:

- `min_size`: smallest object size retained during debris removal.
- `area_threshold`: largest internal hole size that will be filled when hole filling is enabled.
- `large_object_cutoff`: image-adaptive QC threshold used to flag unusually large MAP2 objects.
- `median_area`: median area of the plausibility-filtered object distribution used for stable thresholding.
- `n_components`: number of connected components in the original mask.
- `n_plausible_components`: number of objects falling inside the configured plausible size range.
- `fraction_plausible_components`: fraction of all objects considered plausible during adaptive estimation.

The distribution of object area is explored using histograms, before and after removing large objects.

```{r parameters_df, results='asis'}

# ------------------------------------------------------------------------------
# This chunk visualizes area distribution and median area
# ------------------------------------------------------------------------------

# Print estimated parameters
df <- as.data.frame(py$param_est) %>%
    gather('metric', 'value')
knitr::kable(df)

# Remove objects that are greater than large object cutoff
filtered_df <- py$initial_df %>%
  dplyr::filter(area <= df[df$metric == 'large_object_cutoff', 'value'])

df_list <- list(before=py$initial_df, after=filtered_df)

# Visualize data
xintercept <- df[df$metric == 'median_area', 'value']
for (name in names(df_list)) {
    cat('##', name, 'filtering at the large obj cutoff\n\n')
    print(plot_histo(df_list[[name]], xintercept=xintercept))
    cat('\n\n')
}
```

Note that the red dashed vertical lines indicate median area. The “after” histogram is shown 
only to make the distribution easier to read after excluding extreme large objects from plotting;
it does not mean those large objects are removed from segmentation.

# Cleaning the MAP2 mask

This section removes very small disconnected debris, optionally fills small holes, and optionally
applies a light morphology opening for conservative cleanup.

```{python clean_mask}

# ------------------------------------------------------------------------------
# This chunk removes tiny disconnected debris, optionally fills small holes,
# and preserves the main MAP2-positive territories while keeping cleanup conservative
# ------------------------------------------------------------------------------

cleaned_mask = clean_input_mask(
    cyt,
    min_size=int(param_est["min_size"]),
    area_threshold=int(param_est["area_threshold"]),
    connectivity=r.connectivity,
    fill_holes=bool(r.fill_holes),
    border_clear=bool(r.border_clear)
)

if bool(r.apply_opening) and int(r.opening_radius) > 0:
    cleaned_mask = morphology.binary_opening(
        cleaned_mask,
        footprint=morphology.disk(int(r.opening_radius))
    )

cleaned_foreground_pixels = int(cleaned_mask.sum())
cleaned_foreground_fraction = float(cleaned_foreground_pixels / image_area_px) if image_area_px > 0 else np.nan

print("cleaned_foreground_pixels =", cleaned_foreground_pixels)
print("cleaned_foreground_fraction =", cleaned_foreground_fraction)
```

# Segmenting cytoplasmic regions

This step labels connected MAP2-positive objects after cleanup.


```{python label_cyto}

# ------------------------------------------------------------------------------
# This chunk labels cleaned cytoplasmic objects
# ------------------------------------------------------------------------------

# reticulate::repl_python()
# Label cytoplasmic objects
cyto_labels = measure.label(cleaned_mask, connectivity=r.connectivity)
```

# Building the final feature table

This step computes region-level measurements from the final labeled image. Refer to the following table
for detailed information:

- label
- area
- centroid
- bounding box
- eccentricity
- solidity
- extent
- major/minor axis lengths
- equivalent diameter
- perimeter

The `is_large_object_flag` column is included for QC only. It can help identify large connected 
regions that may represent biologically real extended neuronal structure, merged signal, or 
segmentation artifacts.

```{python feature_df}

# ------------------------------------------------------------------------------
# This chunk summarizes final MAP2-positive segmented objects
# ------------------------------------------------------------------------------

cytoplasm_df = regionprops_dataframe(cyto_labels)
cytoplasm_df = add_border_flag(cytoplasm_df, cyto_labels)

if cytoplasm_df.shape[0] > 0:
    large_cutoff = param_est["large_object_cutoff"]
    cytoplasm_df["is_large_object_flag"] = cytoplasm_df["area"] >= large_cutoff
else:
    cytoplasm_df["is_large_object_flag"] = pd.Series(dtype=bool)

cytoplasm_df = append_sample_metadata(
    cytoplasm_df,
    sample_id=r.sample_id,
    image_id=r.image_id,
    group=r.group,
    pixel_size_um=r.pixel_size_um,
    magnification=r.magnification
)

print(cytoplasm_df.head())
print(cytoplasm_df.tail())
```

```{python add_to_img}

# ------------------------------------------------------------------------------
# this chunk adds the cleaned masked array back to the imagecontainer obj
# ------------------------------------------------------------------------------

# reticulate::repl_python()
# Add the cleaned mask array to the ImageContainer obj
img[r.out_layer].data[:, :, 0, r.cyt_chan] = cleaned_mask

# Delete unnecessary layers
layers_del = [l for l in list(img) if l not in ['image', r.sq_layer, r.out_layer]]
print("The following layers are deleted: ", layers_del)
for l in layers_del:
    del img[l]
```

# Building run-level QC summary

Metrics:

- `input_foreground_fraction`: fraction of image occupied by MAP2 signal before cleanup.
- `cleaned_foreground_fraction`: fraction of image occupied by MAP2 signal after cleanup.
- `n_initial_components`: number of connected components before cleanup.
- `n_final_components`: number of connected components after cleanup.
- `fraction_foreground_retained`: fraction of MAP2-positive pixels retained after cleanup.
- `fraction_components_retained`: fraction of connected objects retained after cleanup.
- `n_large_objects_final`: number of final objects exceeding the adaptive large-object QC threshold.
- `n_border_objects_final`: number of final objects touching the image edge.
- `flag_low_signal_image`: warns that the image contains very little MAP2-positive foreground.
- `flag_component_collapse`: warns that many initial objects collapsed into far fewer final objects.

```{python run_qc}

# ------------------------------------------------------------------------------
# this chunk summaizes the processing outcome at the image/run level
# ------------------------------------------------------------------------------

run_summary = build_mask_run_summary(
    input_mask=cyt,
    cleaned_mask=cleaned_mask,
    initial_df=initial_df,
    final_df=cytoplasm_df,
    sample_id=r.sample_id,
    image_id=r.image_id,
    group=r.group,
    pixel_size_um=r.pixel_size_um,
    magnification=r.magnification
)

run_summary["estimated_min_size"] = int(param_est["min_size"])
run_summary["estimated_area_threshold"] = int(param_est["area_threshold"])
run_summary["estimated_large_object_cutoff"] = int(param_est["large_object_cutoff"])
run_summary["n_large_objects_final"] = int(cytoplasm_df["is_large_object_flag"].sum()) if cytoplasm_df.shape[0] > 0 else 0
run_summary["n_border_objects_final"] = int(cytoplasm_df["touches_border"].sum()) if cytoplasm_df.shape[0] > 0 else 0

artifact_flags = flag_sparse_or_artifact_image(
    run_summary["input_foreground_fraction"],
    run_summary["cleaned_foreground_fraction"],
    run_summary["n_initial_components"],
    run_summary["n_final_components"]
)

run_summary_df = pd.DataFrame([run_summary])
```

```{r print_run_qc, results='asis'}
py$run_summary_df %>%
    t() %>%
    as.data.frame() %>%
    rownames_to_column('metric') %>%
    set_names(c('metric', 'value')) %>%
    knitr::kable()
```

# Visualizing cleaned nuclear segmentation {.tabset}

- Input: `r sq_layer`
- Output: `r out_layer`

Unique pixels at each array are visualized using the following pseudocolors:

```{python vis_prep}

# ------------------------------------------------------------------------------
# This chunk is used to prepare visualizing processed images
# ----------------------------------------------------------------------------

# Specify a colormap used for unique pixels
color_map = {r.sq_layer: 'dodgerblue', r.out_layer: 'orange'}
for key, value in color_map.items(): 
    print(key, "=", value)

# Specify a function to visualize unique pixels
def display_merge(layer_list):

    # Stack binary images
    for l in layer_list:
        ar = img[l].data[:, :, 0, r.cyt_chan]
        plt.imshow(ar,
                   cmap=ListedColormap([(1, 1, 1, 0), color_map[l]]),
                   alpha=0.5)
    plt.xticks([])
    plt.yticks([])
    plt.savefig(f"{r.out_dir}/unique_pixels.png", dpi=600)
    plt.show()
```

```{r vis_images, results='asis'}

# ------------------------------------------------------------------------------
# This chunk visualizes input and output masking
# ------------------------------------------------------------------------------

# Visualize binary images across the channels
for (layer in c(sq_layer, out_layer)) {
    cat("## Channel", cyt_chan, "&", layer, "Layer\n\n")
    # Command to show image
    t_deparsed_1 <- paste0("img.show('", layer, "', channel=", cyt_chan, ", cmap='gray')")
    t_deparsed_2 <- paste0("plt.savefig('", out_dir, "/", layer, "_channel_", cyt_chan,  ".png', dpi=600)")
    t_deparsed_3 <- "plt.close()"
    t_deparsed <- c(t_deparsed_1, t_deparsed_2, t_deparsed_3)
    subchunkify(name=paste0(layer, "_", cyt_chan), t_deparsed=t_deparsed)

    link_output(paste0(out_dir, "/", layer, "_channel_", cyt_chan, ".png"))
    cat('\n\n')
}

# Overlay images before and after cleaning
cat("## Unique pixels\n\n")
t_deparsed <- paste0("display_merge(list(color_map.keys()))\n")
subchunkify(name="unique_pixels", t_deparsed=t_deparsed)
link_output(paste0(out_dir, "/unique_pixels.png"))
```

# Summary tables

Excel-readable tables are linked in the following table:

- input summary: object statistics before cleanup
- nuclear summary: object statistics after cleanup and connected-component labeling
- parameter summary: adaptive thresholds estimated from the image
- image QC summary: image-level QC summary for the run

```{r summary_tables, results='asis'}

# ------------------------------------------------------------------------------
# Aim of this chunk:
# - save all object-level and image-level summary tables
# - keep outputs easy to combine across multiple images later
# ------------------------------------------------------------------------------

df_list <- list(
    input_summary = py$initial_df,
    parameter_summary = as.data.frame(py$param_est),
    image_qc_summary = py$run_summary_df
)

path_df <- data.frame(table = names(df_list), link = "")

for (name in names(df_list)) {
    tsv <- file.path(out_dir, paste0(name, ".tsv"))
    df <- df_list[[name]]
    tryCatch({
        write.table(
            df,
            tsv,
            quote = FALSE,
            row.names = FALSE,
            col.names = TRUE,
            sep = "\t"
        )
    }, error = function(e) {
        c_length <- apply(df, 2, class)
        for (c in colnames(df)) {
            if (is.list(df[[c]])) {
                df[[c]] <- df[[c]] %>%
                    paste(sep=", ")
            }
        }
        write.table(
            df,
            tsv,
            quote = FALSE,
            row.names = FALSE,
            col.names = TRUE,
            sep = "\t"
        )
    })

    path_df[path_df$table == name, 'link'] <- tsv
}

knitr::kable(
    path_df %>%
        mutate(link = paste0("[", link, "](", link, ")"))
)
```

# Saving the `ImageContainer` object

Output images are saved as the following object:

```{python save_zarr}

# ------------------------------------------------------------------------------
# This chunk saves updated ImageContainer as Zarr
# ------------------------------------------------------------------------------

img.save(r.output_obj)
print(r.output_obj)
```

# Session info

```{r session_info, collapse=FALSE}
sessionInfo()
```
