---
title: "Compartment integration & TDP43 overlap quantification"
author: "Mira Sohn"
output:
    html_document:
        code_folding: hide
        df_print: paged
        toc: true
        toc_float: true
        toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE,
                      message=FALSE,
                      cache.lazy=FALSE)
```

Last run: `r date()`

It integrates previously cleaned/segmented DAPI and MAP2 masks and quantifies binarized TDP43
occupancy across nuclear and extra-nuclear compartments.

```{r libraries}
# Libraries
library(parallel)
library(tidyverse)
library(ggplot2)
library(reticulate)

# Additional options
set.seed(2570)

# Specify the number of available cores
# n_cores <- future::availableCores()
# options(mc.cores=n_cores)
```


```{r config}

# Path to output directory
out_dir <- "tdp43"
if (! dir.exists(out_dir)) { dir.create(out_dir, recursive=TRUE) }

# Load configuration
config <- yaml::read_yaml("config/config.yaml")

# Input & output obj paths
input_obj <- "map2/cytoplasmic_seg.zarr"
output_obj <- file.path(out_dir, "tdp32_quant.zarr")

# Specify the layers of ImageContainer obj for input and output arrays
sq_layer <- config[['sq_layer']]
out_layer <- config[['out_layer']]

# Specify channels
channels <- map(
    names(config[['channels']]),
    ~config[['channels']][[.x]]) %>%
    set_names(names(config[['channels']]))

# Specify channel names
chan_names <- list(nuc="dapi", cyt="map2", target="tdp43")

# Specify connectivity to determine which neighboring pixels are 
# considered connected when forming objects
connectivity <- config[['connectivity']]

# Conservative cleanup settings for binarized TDP43
target_min_component_quantile <- config[['target']][['min_component_quantile']]
target_area_threshold_quantile <- config[['target']][['area_threshold_quantile']]
target_large_object_quantile <- config[['target']][['large_object_quantile']]
target_max_reasonable_area_multiplier <- config[['target']][['max_reasonable_area_multiplier']]

# TRUE/FALSE options for TDP43 handling
target_apply_cleanup <- config[['target']][['apply_cleanup']]
target_fill_holes <- config[['target']][['fill_holes']]
target_border_clear <- config[['target']][['border_clear']]

# Optional nucleus-to-MAP2 assignment settings
assignment_overlap_min_pixels <- config[['integration']][['assignment_overlap_min_pixels']]
assignment_max_distance <- config[['integration']][['assignment_max_distance']]

# Load helper functions
source(config[['helper']][['R']])

# Specify conda env
use_condaenv(config[['conda_env']])
source_python(config[['helper']][['py']])
```

```{python ppackages}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from scipy import ndimage as ndi
from skimage import io, measure, morphology, segmentation, feature, color, util
from matplotlib.colors import ListedColormap
import dask.array as da
import squidpy as sq
import xarray as xr
import subprocess
import os
```

This workflow performs compartment integration and overlap-based TDP43 quantification 
from previously cleaned binary masks.

Assumptions:

- DAPI has already been cleaned and segmented into nuclear objects
- MAP2 has already been cleaned into a usable MAP2-positive mask
- TDP43 is already binarized
- TDP43 should be treated as an occupancy mask, not an intensity image
- extra-nuclear compartment is defined initially as MAP2-positive signal minus nucleus
- any additional TDP43 cleanup should be minimal and conservative

# Loading input `ImageContainer`

```{python load_input}

# ------------------------------------------------------------------------------
# This chunk imports input image saved as zarr in the previous step
# ------------------------------------------------------------------------------

# reticulate::repl_python()

# Load the Zarr store into an ImageContainer
# Set lazy=True to enable Dask-backed lazy loading, which is memory-efficient for large images.
# You can also specify chunks for Dask if needed.
img = sq.im.ImageContainer.load(r.input_obj, lazy=True)

print("Input loaded:\n", r.input_obj)
print("ImageContainer:\n", img)
```

# Extracting masks used for integration

This step retrieves:

- DAPI cleaned mask
- MAP2 cleaned mask
- TDP43 binary mask

```{python input_summary}

# ------------------------------------------------------------------------------
# This chunk inspects the input image
# ------------------------------------------------------------------------------

# Retrieve the Dask arrays for channels of interest ([y, x, z, channels])
arr_dic = {}
for key, value in r.channels.items(): 
    arr = img[r.out_layer].data[:, :, 0, value]
    arr_dic[key] = arr > 0

for key, value in arr_dic.items():
    print(key, "foreground pixels =", int(value.sum()))
```

# Summarizing initial TDP43 input

We first inspect the connected-component structure of the binarized TDP43 mask, mainly to decide 
whether minimal cleanup is needed.

```{python inspect_tdp}

# ------------------------------------------------------------------------------
# This chunk summarizes the input TDP43 mask
# ------------------------------------------------------------------------------
# reticulate::repl_python()
target_initial_labels, target_initial_df = summarize_components(
    arr_dic[r.chan_names['target']],
    connectivity=r.connectivity)
```

Metrics:

- Label: A unique integer assigned to each segmented object so individual nuclei 
  can be identified and tracked separately.
- Area: The number of foreground pixels belonging to a segmented object, 
  representing its size in pixel units.
- Bounding box: The smallest axis-aligned rectangle that fully contains 
  the segmented object.
- Eccentricity: A shape descriptor measuring how elongated an object is, 
  where 0 is circle-like and values closer to 1 are more ellipse-like or stretched.
- Solidity: The fraction of the object’s area relative to the area of its convex hull,
  indicating how solid versus indented or irregular the shape is.
- Extent: The fraction of the bounding box occupied by the object, reflecting how fully 
 the object fills its enclosing rectangle.

```{r input_summary_df}
print(head(py$target_initial_df))
print(tail(py$target_initial_df))
```

# Estimating conservative TDP43 cleanup parameters {.tabset}

These parameters are used only if `tdp43_apply_cleanup = TRUE`.

```{python cleanup_parameters}

# ------------------------------------------------------------------------------
# This chunk estimates conservative cleanup parameters for TDP43
# ------------------------------------------------------------------------------

target_param_est = estimate_cleanup_parameters(
    target_initial_df,
    min_component_quantile=r.target_min_component_quantile,
    area_threshold_quantile=r.target_area_threshold_quantile,
    large_object_quantile=r.target_large_object_quantile,
    max_reasonable_area_multiplier=r.target_max_reasonable_area_multiplier)
```

- `min_size`: Lower bound for removing tiny disconnected debris objects
- `area_threshold`: Maximum hole size that may be filled if `fill_holes = TRUE`
- `large_object_cutoff`: Upper-tail threshold used to flag unusually large objects for review
- `median_area`: Useful descriptive summary of the connected-component distribution

The distribution of object area is explored using histograms, before and after removing large objects.

```{r parameters_df, results='asis'}

# ------------------------------------------------------------------------------
# This chunk visualizes area distribution and median area
# ------------------------------------------------------------------------------

df <- as.data.frame(py$target_param_est)
knitr::kable(df)

filtered_df <- py$target_initial_df %>%
  dplyr::filter(area <= df[['large_object_cutoff']])

df_list <- list(before=py$target_initial_df, after=filtered_df)

for (name in names(df_list)) {
    cat('##', name, 'filtering at the large obj cutoff\n\n')
    print(plot_histo(df_list[[name]]))
    cat('\n\n')
}
```

Note that the red dashed vertical lines indicate median area. The “after” histogram is shown 
only to make the distribution easier to read after excluding extreme large objects from plotting;
it does not mean those large objects are removed from segmentation.

# Cleaning TDP43 mask conservatively

This step is optional. If enabled, it should only:

- remove tiny disconnected speckles
- fill very small holes if requested
- avoid aggressive morphology that could distort localization

```{python cleanup_tdp}

# ------------------------------------------------------------------------------
# This chunk applies optional conservative cleanup to TDP43
# ------------------------------------------------------------------------------

if bool(r.target_apply_cleanup):
    tdp_clean = clean_input_mask(
        arr_dic[r.chan_names['target']],
        min_size=int(target_param_est["min_size"]),
        area_threshold=int(target_param_est["area_threshold"]),
        connectivity=r.connectivity,
        fill_holes=bool(r.target_fill_holes),
        border_clear=bool(r.target_border_clear)
    )
else:
    tdp_clean = tdp.copy()

print("TDP43 cleaned foreground pixels =", int(tdp_clean.sum()))
```

# Building compartment masks

We define:

- `nuclear_mask` = DAPI-positive nuclear mask
- `cytoplasmic_mask` = MAP2-positive extra-nuclear mask
- `target_nuclear_mask` = TDP43 overlapping nucleus
- `target_cytoplasmic_mask` = TDP43 overlapping extra-nuclear MAP2 region
- `target_other_mask` = TDP43 outside both primary compartments

```{python compartment_masks}

# ------------------------------------------------------------------------------
# This chunk constructs compartment masks
# ------------------------------------------------------------------------------

# reticulate::repl_python()
nuclear_mask = arr_dic[r.chan_names['nuc']]
cytoplasmic_mask = np.logical_and(
    arr_dic[r.chan_names['cyt']],
    np.logical_not(nuclear_mask))

target_nuclear_mask = np.logical_and(target_clean, nuclear_mask)
target_cytoplasmic_mask = np.logical_and(target_clean, cytoplasmic_mask)
target_other_mask = np.logical_and(
    target_clean,
    np.logical_not(np.logical_or(nuclear_mask, cytoplasmic_mask))
)

print("Nuclear pixels =", int(nuclear_mask.sum()))
print("Cytoplasmic pixels =", int(cytoplasmic_mask.sum()))
print("TDP43 nuclear pixels =", int(target_nuclear_mask.sum()))
print("TDP43 cytoplasmic pixels =", int(target_cytoplasmic_mask.sum()))
print("TDP43 other pixels =", int(target_other_mask.sum()))
```

# Session info

```{r session_info, collapse=FALSE}
sessionInfo()
```



