---
title: "Compartment integration & TDP43 overlap quantification"
author: "Mira Sohn"
output:
    html_document:
        code_folding: hide
        df_print: paged
        toc: true
        toc_float: true
        toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE,
                      message=FALSE,
                      cache.lazy=FALSE)
```

Last run: `r date()`

It integrates previously cleaned/segmented DAPI and MAP2 masks and quantifies binarized TDP43
occupancy across nuclear and extra-nuclear compartments.

```{r libraries}
# Libraries
library(parallel)
library(tidyverse)
library(ggplot2)
library(reticulate)

# Additional options
set.seed(2570)

# Specify the number of available cores
# n_cores <- future::availableCores()
# options(mc.cores=n_cores)
```

```{r config}

# ------------------------------------------------------------------------------
# This chunk configures file paths and variables
# ------------------------------------------------------------------------------

# Path to output directory
out_dir <- "tdp43"
if (! dir.exists(out_dir)) { dir.create(out_dir, recursive=TRUE) }

# Load configuration
config <- yaml::read_yaml("config/config.yaml")

# Input & output obj paths
input_obj <- "map2/cytoplasmic_seg.zarr"
output_obj <- file.path(out_dir, "tdp43_quant.zarr")

# Specify the layers of ImageContainer obj for input and output arrays
sq_layer <- config[['sq_layer']]
out_layer <- config[['out_layer']]

# Specify channels
channels <- map(
    names(config[['channels']]),
    ~config[['channels']][[.x]]) %>%
    set_names(names(config[['channels']]))

# Specify channel names
chan_names <- list(nuc="dapi", cyt="map2", target="tdp43")

# Specify connectivity to determine which neighboring pixels are 
# considered connected when forming objects
connectivity <- config[['connectivity']]

# Conservative cleanup settings for binarized TDP43
target_min_component_quantile <- config[['target']][['min_component_quantile']]
target_area_threshold_quantile <- config[['target']][['area_threshold_quantile']]
target_large_object_quantile <- config[['target']][['large_object_quantile']]
target_max_reasonable_area_multiplier <- config[['target']][['max_reasonable_area_multiplier']]

# TRUE/FALSE options for TDP43 handling
target_apply_cleanup <- config[['target']][['apply_cleanup']]
target_fill_holes <- config[['target']][['fill_holes']]
target_border_clear <- config[['target']][['border_clear']]

# Optional nucleus-to-MAP2 assignment settings
assignment_overlap_min_pixels <- config[['integration']][['assignment_overlap_min_pixels']]
assignment_max_distance <- config[['integration']][['assignment_max_distance']]

# Load helper functions
source(config[['helper']][['R']])

# Specify conda env
use_condaenv(config[['conda_env']])
source_python(config[['helper']][['py']])
```

```{python ppackages}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from scipy import ndimage as ndi
from scipy.spatial import cKDTree
from skimage import io, measure, morphology, segmentation, feature, color, util
from matplotlib.colors import ListedColormap
import dask.array as da
import squidpy as sq
import xarray as xr
import subprocess
import os
```

This workflow performs compartment integration and overlap-based TDP43 quantification 
from previously cleaned binary masks.

Assumptions:

- DAPI has already been cleaned and segmented into nuclear objects
- MAP2 has already been cleaned into a usable MAP2-positive mask
- TDP43 is already binarized
- TDP43 should be treated as an occupancy mask, not an intensity image
- extra-nuclear compartment is defined initially as MAP2-positive signal minus nucleus
- any additional TDP43 cleanup should be minimal and conservative

# Loading input `ImageContainer`

```{python load_input}

# ------------------------------------------------------------------------------
# This chunk imports input image saved as zarr in the previous step
# ------------------------------------------------------------------------------

# reticulate::repl_python()

# Load the Zarr store into an ImageContainer
# Set lazy=True to enable Dask-backed lazy loading, which is memory-efficient for large images.
# You can also specify chunks for Dask if needed.
img = sq.im.ImageContainer.load(r.input_obj, lazy=True)

print("Input loaded:\n", r.input_obj)
print("ImageContainer:\n", img)
```

# Extracting masks used for integration

This step retrieves:

- DAPI cleaned mask
- MAP2 cleaned mask
- TDP43 binary mask

```{python input_summary}

# ------------------------------------------------------------------------------
# This chunk inspects the input image
# ------------------------------------------------------------------------------

# Retrieve the Dask arrays for channels of interest ([y, x, z, channels])
arr_dic = {}
for key, value in r.channels.items(): 
    arr = img[r.out_layer].data[:, :, 0, value]
    arr_dic[key] = arr > 0

for key, value in arr_dic.items():
    print(key, "foreground pixels =", int(value.sum()))
```

# Summarizing initial TDP43 input

This section inspects the connected-component structure of the binarized TDP43 mask before any 
optional cleanup is applied.

The purpose is to understand whether the current TDP43 mask contains many tiny speckles, 
fragmented objects, or unusually large connected components that might justify conservative cleanup. 
This is an inspection step, not yet a transformation step.

```{python inspect_tdp}

# ------------------------------------------------------------------------------
# This chunk summarizes the input TDP43 mask
# ------------------------------------------------------------------------------
# reticulate::repl_python()

# Compute connected-component labels and a per-object summary table for the target protein
target_initial_labels, target_initial_df = summarize_components(
    arr_dic[r.chan_names['target']],
    connectivity=r.connectivity)
```

Metrics shown below describe the connected components in the initial TDP43 mask:

- Label: A unique integer assigned to each segmented object so individual nuclei 
  can be identified and tracked separately.
- Area: The number of foreground pixels belonging to a segmented object, 
  representing its size in pixel units.
- Bounding box: The smallest axis-aligned rectangle that fully contains 
  the segmented object.
- Eccentricity: A shape descriptor measuring how elongated an object is, 
  where 0 is circle-like and values closer to 1 are more ellipse-like or stretched.
- Solidity: The fraction of the object’s area relative to the area of its convex hull,
  indicating how solid versus indented or irregular the shape is.
- Extent: The fraction of the bounding box occupied by the object, reflecting how fully 
 the object fills its enclosing rectangle.

```{r input_summary_df}
print(head(py$target_initial_df))
print(tail(py$target_initial_df))
```

# Estimating conservative TDP43 cleanup parameters {.tabset}

This section computes adaptive cleanup thresholds from the observed TDP43 component distribution.

These thresholds are used only if cleanup is enabled in the config file. The goal is to support minimal,
data-driven cleanup rather than hard-coded morphology settings.

```{python cleanup_parameters}

# ------------------------------------------------------------------------------
# This chunk estimates conservative cleanup parameters for TDP43
# ------------------------------------------------------------------------------

target_param_est = estimate_cleanup_parameters(
    target_initial_df,
    min_component_quantile=r.target_min_component_quantile,
    area_threshold_quantile=r.target_area_threshold_quantile,
    large_object_quantile=r.target_large_object_quantile,
    max_reasonable_area_multiplier=r.target_max_reasonable_area_multiplier)
```

In particular, this step estimates:

- `min_size`: Lower bound for removing tiny disconnected debris objects
- `area_threshold`: Maximum hole size that may be filled if `fill_holes = TRUE`
- `large_object_cutoff`: Upper-tail threshold used to flag unusually large objects for review
- `median_area`: Useful descriptive summary of the connected-component distribution

The distribution of object area is explored using histograms, before and after removing large objects.

```{r parameters_df, results='asis'}

# ------------------------------------------------------------------------------
# This chunk visualizes area distribution and median area
# ------------------------------------------------------------------------------

# Convert the parameter summary into an R data frame for display
df <- as.data.frame(py$target_param_est)
knitr::kable(df)

# Exclude very large objects only for improved plotting readability
filtered_df <- py$target_initial_df %>%
  dplyr::filter(area <= df[['large_object_cutoff']])

df_list <- list(before=py$target_initial_df, after=filtered_df)

# Render area histograms for both views
for (name in names(df_list)) {
    cat('##', name, 'filtering at the large obj cutoff\n\n')
    print(plot_histo(df_list[[name]]))
    cat('\n\n')
}
```

Note that the red dashed vertical lines indicate median area. The “after” histogram is shown 
only to make the distribution easier to read after excluding extreme large objects from plotting;
it does not mean those large objects are removed from segmentation.

# Cleaning TDP43 mask conservatively

This section applies optional cleanup to the binarized TDP43 mask. Because TDP43 is the signal being 
quantified rather than the compartment-defining channel, cleanup should remain conservative. 

If enabled, it should only:

- remove tiny disconnected speckles
- fill very small holes if requested
- avoid aggressive morphology that could distort localization

```{python cleanup_tdp}

# ------------------------------------------------------------------------------
# This chunk applies optional conservative cleanup to TDP43
# ------------------------------------------------------------------------------

# Apply cleanup only if enabled in the config file
if bool(r.target_apply_cleanup):
    target_clean = clean_input_mask(
        arr_dic[r.chan_names['target']],
        min_size=int(target_param_est["min_size"]),
        area_threshold=int(target_param_est["area_threshold"]),
        connectivity=r.connectivity,
        fill_holes=bool(r.target_fill_holes),
        border_clear=bool(r.target_border_clear)
    )
else:
    # Otherwise keep the original TDP43 binary mask unchanged
    target_clean = arr_dic[r.chan_names['target']].copy()

print("TDP43 cleaned foreground pixels =", int(target_clean.sum()))
```

# Building compartment masks

This section defines the core biological compartments used for all downstream measurements. The current 
first-pass compartment logic is:

- `nuclear_mask` = DAPI-positive nuclear mask
- `cytoplasmic_mask` = MAP2-positive extra-nuclear mask
- `target_nuclear_mask` = TDP43 overlapping nucleus
- `target_cytoplasmic_mask` = TDP43 overlapping extra-nuclear MAP2 region
- `target_other_mask` = TDP43 outside both primary compartments

```{python compartment_masks}

# ------------------------------------------------------------------------------
# This chunk constructs compartment masks
# ------------------------------------------------------------------------------

# reticulate::repl_python()

# Use the DAPI mask as the nuclear compartment
nuclear_mask = arr_dic[r.chan_names['nuc']]

# Define extra-nuclear MAP2-positive territory as cytoplasmic mask
cytoplasmic_mask = np.logical_and(
    arr_dic[r.chan_names['cyt']],
    np.logical_not(nuclear_mask))

# Intersect cleaned TDP43 with each compartment definition
target_nuclear_mask = np.logical_and(target_clean, nuclear_mask)
target_cytoplasmic_mask = np.logical_and(target_clean, cytoplasmic_mask)

# Define residual TDP43 outside both main compartments
target_other_mask = np.logical_and(
    target_clean,
    np.logical_not(np.logical_or(nuclear_mask, cytoplasmic_mask))
)

# Print basic pixel-count summaries for each mask
print("Nuclear pixels =", int(nuclear_mask.sum()))
print("Cytoplasmic pixels =", int(cytoplasmic_mask.sum()))
print("TDP43 nuclear pixels =", int(target_nuclear_mask.sum()))
print("TDP43 cytoplasmic pixels =", int(target_cytoplasmic_mask.sum()))
print("TDP43 other pixels =", int(target_other_mask.sum()))
```

# Labeling nuclei and cytoplasm-positive objects

This section converts the binary nuclear and cytoplasmic masks into connected-component label 
images. Label images make it possible to summarize individual nuclei and cytoplasmic objects 
without repeatedly creating expensive full-image masks in later steps.

These labels are used downstream for:

- nucleus-to-MAP2 assignment
- per-nucleus summaries
- cytoplasmic object summaries

```{python compartmental_labels}

# ------------------------------------------------------------------------------
# This chunk labels nuclei and MAP2-positive objects
# ------------------------------------------------------------------------------

# Label connected components in the nuclear and cytoplasmic masks
nuclear_labels = measure.label(nuclear_mask, connectivity=r.connectivity)
cyto_labels = measure.label(cytoplasmic_mask, connectivity=r.connectivity)

# Print the number of labeled objects in each compartment
print("Number of nuclei =", int(nuclear_labels.max()))
print("Number of cytoplasmic objects =", int(cyto_labels.max()))
```

# Building nucleus-to-MAP2 assignment table

This section builds a first-pass mapping between nuclei and MAP2-positive objects. To reduce runtime, 
the assignment uses a faster strategy than the original per-nucleus masking loop:

- build nucleus and cytoplasmic object tables once
- assign nuclei by overlap first
- only if enabled, assign remaining unassigned nuclei by nearest centroid
- store the final assignment table for downstream summaries

This avoids repeated full-image mask creation for each nucleus and is much more scalable.

```{python assigment_table}

# ------------------------------------------------------------------------------
# This chunk build summary tables for nuclei and cytoplasmic objects
# ------------------------------------------------------------------------------

# Measure nucleus labels, centroids, and areas
nuc_props = measure.regionprops_table(
    nuclear_labels,
    properties=["label", "centroid", "area"]
)

# Summarize nuclear metrics in a data frame
nuclei_df = pd.DataFrame(nuc_props).rename(columns={
    "label": "nucleus_label",
    "centroid-0": "nuc_y",
    "centroid-1": "nuc_x",
    "area": "nucleus_area"
})

# Measure cytoplasmic object labels, centroids, and areas
cyto_props = measure.regionprops_table(
    cyto_labels,
    properties=["label", "centroid", "area"]
)

# Summarize cytoplasmic metrics in a data frame
cytoplasm_df = pd.DataFrame(cyto_props).rename(columns={
    "label": "cyto_label",
    "centroid-0": "cyto_y",
    "centroid-1": "cyto_x",
    "area": "cyto_area"
})
```

```{python update_assignment_table}

# ------------------------------------------------------------------
# This chunk adds overlap-based nucleus-to-MAP2 assignment
# ------------------------------------------------------------------

# reticulate::repl_python()

# Find pixels where a labeled nucleus and labeled cytoplasmic object coexist
overlap_mask = (nuclear_labels > 0) & (cyto_labels > 0)

# If any overlap exists, count overlap pixels for each nucleus-cytoplasm pair
if overlap_mask.sum() > 0:
    overlap_pairs = pd.DataFrame({
        "nucleus_label": nuclear_labels[overlap_mask].ravel(),
        "cyto_label": cyto_labels[overlap_mask].ravel()
    })

    overlap_counts_df = (
        overlap_pairs
        .groupby(["nucleus_label", "cyto_label"], as_index=False)
        .size()
        .rename(columns={"size": "overlap_pixels"})
        .sort_values(["nucleus_label", "overlap_pixels"], ascending=[True, False])
    )

    # Keep only the strongest overlapping MAP2 object for each nucleus
    best_overlap_df = overlap_counts_df.drop_duplicates(
        subset=["nucleus_label"],
        keep="first"
    )

    # Mark overlap-based assignments
    best_overlap_df["assignment_method"] = "overlap"
    best_overlap_df["centroid_distance"] = np.nan

    # Enforce a minimum overlap requirement before accepting an assignment
    best_overlap_df.loc[
        best_overlap_df["overlap_pixels"] < int(r.assignment_overlap_min_pixels),
        ["cyto_label", "assignment_method"]
    ] = [np.nan, "unassigned"]
else:
    # Create an empty overlap table if no overlaps were detected
    best_overlap_df = pd.DataFrame(columns=[
        "nucleus_label", "cyto_label", "overlap_pixels",
        "assignment_method", "centroid_distance"
    ])

# Merge overlap results back onto the full nucleus table
assignment_df = nuclei_df.merge(
    best_overlap_df[["nucleus_label",
        "cyto_label",
        "overlap_pixels",
        "assignment_method",
        "centroid_distance"]],
    on="nucleus_label",
    how="left"
)

# Fill missing values for nuclei that received no overlap-based assignment
assignment_df["overlap_pixels"] = assignment_df["overlap_pixels"].fillna(0)
assignment_df["assignment_method"] = assignment_df["assignment_method"].fillna("unassigned")
```

```{python cleanup}

# ------------------------------------------------------------------
# This chunk optionally assigns remaining nuclei by nearest cytoplasmic centroid
# ------------------------------------------------------------------

# Proceed only if at least one cytoplasmic object exists
if cytoplasm_df.shape[0] > 0:
    unassigned_mask = assignment_df["assignment_method"] == "unassigned"

    # Use nearest-neighbor assignment only if enabled and needed
    if bool(r.assignment_max_distance) and unassigned_mask.sum() > 0:
        cyto_tree = cKDTree(cytoplasm_df[["cyto_y", "cyto_x"]].to_numpy())

        distances, indices = cyto_tree.query(
            assignment_df.loc[unassigned_mask, ["nuc_y", "nuc_x"]].to_numpy(),
            k=1
        )

        nearest_labels = cytoplasm_df.iloc[indices]["cyto_label"].to_numpy()

        # Store nearest candidate assignments and distances
        assignment_df.loc[unassigned_mask, "nearest_cyto_label"] = nearest_labels
        assignment_df.loc[unassigned_mask, "centroid_distance"] = distances

        # Accept nearest assignments only within the allowed maximum distance
        valid_mask = (
            unassigned_mask &
            (assignment_df["centroid_distance"] <= float(r.assignment_max_distance))
        )

        assignment_df.loc[valid_mask, "cyto_label"] = assignment_df.loc[valid_mask, "nearest_cyto_label"]
        assignment_df.loc[valid_mask, "assignment_method"] = "nearest"

# Finalize the canonical assigned cytoplasmic label column
assignment_df["assigned_cyto_label"] = assignment_df["cyto_label"]

# Drop the temporary column name to keep the final table cleaner
assignment_df = assignment_df.drop(columns=["cyto_label"], errors="ignore")

print(assignment_df.head())
print(assignment_df.tail())
```

# Building compartment-level summary metrics

This section computes whole-image summary metrics describing how TDP43-positive area is distributed 
across nuclear, cytoplasmic, and other regions. These summary values provide the main first-pass 
biological readout for this workflow and are also useful as quick QC measures before looking at 
per-nucleus results.

Mask-based metrics:

- `input_shape_y`, `input_shape_x`: height (y) and width (x) of image
- `nuclear_area`: number of pixels in the nuclear compartment (e.g. DAPI-mask)
- `cytoplasmic_area`: number of pixels in the cytoplasmic compartment (e.g. MAP2-mask)
- `target_total_area`: number of target-positive pixels (e.g. TDP43-positive pixels)
- `target_nuclear_area`: number of target-positive pixels that fall inside the nuclear compartment
  (e.g. `target_clean` AND `nuclear_mask`)
- `target_cytoplasmic_agrea`: number of target-positive pixels that fall inside the cytoplasmic compartment
  (e.g. `target_clean` AND `cytoplasmic_mask`)
- `target_other_area`: number of target-positive pixels that fall outside both the nuclear and 
  cytoplasmic compartments
  (e.g. signal outside segmented cells, segmentation mismatch, or TDP43-positive pixels 
  not covered by the current compartment definitions)

Fraction-based metrics:

- `nuclear_fraction_of_target`: `target_nuclear_area` / `target_total_area`
- `cytoplasmic_fraction_of_target`: `target_nuclear_area` / `target_total_area`
- `other_fraction_of_target`: `target_other_area` / `target_total_area`

Occupancy-based metrics:

- `nuclear_occupancy`: `target_nuclear_area` / `nuclear_area` (e.g. DAPI AND TDP43)
- `cytoplasmic_occupancy`: `target_cytoplasmic_area` / `nuclear_area` (e.g. MAP2 AND TDP43)
- `cytoplasmic_to_nuclear_target_area_ratio`:
  `target_cytoplasmic_area` / `target_nuclear_area`
  (> 1 if more TDP43 signal is greater in cytoplasm)

```{python compartment_metrics}

# ------------------------------------------------------------------------------
# This chunk computes whole-image compartment overlap and occupancy metrics
# ------------------------------------------------------------------------------

# reticulate::repl_python()
# Count total TDP43-positive area after cleanup
target_total_area = int(target_clean.sum())

# Count TDP43-positive area in each compartment
target_nuclear_area = int(target_nuclear_mask.sum())
target_cytoplasmic_area = int(target_cytoplasmic_mask.sum())
target_other_area = int(target_other_mask.sum())

# Count total available area for the nuclear and cytoplasmic compartments
nuclear_area = int(nuclear_mask.sum())
cytoplasmic_area = int(cytoplasmic_mask.sum())

# Build a one-row summary table containing whole-image compartment metrics
compartment_summary_df = pd.DataFrame([{
    "input_shape_y": int(nuclear_mask.shape[0]),
    "input_shape_x": int(nuclear_mask.shape[1]),
    "nuclear_area": nuclear_area,
    "cytoplasmic_area": cytoplasmic_area,
    "target_total_area": target_total_area,
    "target_nuclear_area": target_nuclear_area,
    "target_cytoplasmic_area": target_cytoplasmic_area,
    "target_other_area": target_other_area,
    "nuclear_fraction_of_target": (
        target_nuclear_area / target_total_area if target_total_area > 0 else np.nan
    ),
    "cytoplasmic_fraction_of_target": (
        target_cytoplasmic_area / target_total_area if target_total_area > 0 else np.nan
    ),
    "other_fraction_of_target": (
        target_other_area / target_total_area if target_total_area > 0 else np.nan
    ),
    "nuclear_occupancy": (
        target_nuclear_area / nuclear_area if nuclear_area > 0 else np.nan
    ),
    "cytoplasmic_occupancy": (
        target_cytoplasmic_area / cytoplasmic_area if cytoplasmic_area > 0 else np.nan
    ),
    "cytoplasmic_to_nuclear_target_area_ratio": (
        target_cytoplasmic_area / target_nuclear_area if target_nuclear_area > 0 else np.nan
    )
}])
```

```{r compartment_summary, results='asis'}
knitr::kable(t(py$compartment_summary_df) %>% as.data.frame %>% set_names('value'))
```

# Building per-nucleus summary metrics

This section computes per-nucleus TDP43 overlap metrics using a fast label-based aggregation approach.

Rather than generating a full binary mask for each nucleus, this method uses `np.bincount` 
on the label image. This is much faster and scales better to large images or many nuclei.

The resulting table contains:

- per-nucleus area
- TDP43-positive area inside each nucleus
- nuclear occupancy
- assigned cytoplasmic object
- assigned cytoplasmic area and TDP43 overlap
- cytoplasmic-to-nuclear TDP43 ratio

Assignment-related metrics:

- `assigned_cyto_label`: label ID of the cytoplasmic object assigned to that nucleus
- `assignment_method`: `overlap`, `nearest` (nearest centroid if no overlap assignment was available),
  `unassigned` (no cytoplasmic object was assigned)
- `overlap_pixels`: number of pixels of overlap between the nucleus and its best-matching cytoplasmic object
- `centroid_distance`: distance between the nucleus centroid and the assigned cytoplasmic object centroid

Compartment-related metrics:

- `assigned_cyto_area`: area of the assigned cytoplasmic object (e.g. assigned MAP2-positive objects)
- `assigned_cyto_target_area`: number of target-positive pixels inside the assigned cytoplasmic object
  (e.g. assigned TDP43 AND MAP2-positive objects)
- `assigned_cyto_occupancy`: `assigned_cyto_target_area` / `assigned_cyto_area`
- `cytoplasmic_to_nuclear_target_area_ratio`:
  `assigned_cyto_target_area` / `nuclear_target_area`

```{python per_nuc_metrics}

# ------------------------------------------------------------------------------
# Compute fast per-nucleus summary metrics using label-based aggregation
# ------------------------------------------------------------------------------

# Count total pixels belonging to each nucleus label
nucleus_area_counts = np.bincount(nuclear_labels.ravel())

# Count TDP43-positive pixels belonging to each nucleus label
nuclear_target_counts = np.bincount(
    nuclear_labels[target_clean].ravel(),
    minlength = nucleus_area_counts.shape[0]
)

# Build a nucleus-level summary table
per_nucleus_df = pd.DataFrame({
    "nucleus_label": np.arange(1, nucleus_area_counts.shape[0]),
    "nucleus_area": nucleus_area_counts[1:],
    "nuclear_target_area": nuclear_target_counts[1:]
})

# Compute fraction of each nucleus occupied by TDP43
per_nucleus_df["nuclear_occupancy"] = (
    per_nucleus_df["nuclear_target_area"] /
    per_nucleus_df["nucleus_area"].replace(0, np.nan)
)

# Count total pixels belonging to each cytoplasmic object label
cyto_area_counts = np.bincount(cyto_labels.ravel())

# Count TDP43-positive pixels belonging to each cytoplasmic object label
cyto_target_counts = np.bincount( cyto_labels[target_clean].ravel(), minlength = cyto_area_counts.shape[0])

# Build a cytoplasmic object summary table
cyto_summary_df = pd.DataFrame({
    "assigned_cyto_label": np.arange(1, cyto_area_counts.shape[0]),
    "assigned_cyto_area": cyto_area_counts[1:],
    "assigned_cyto_target_area": cyto_target_counts[1:]
})

# Compute fraction of each assigned cytoplasmic object occupied by TDP43
cyto_summary_df["assigned_cyto_occupancy"] = (
    cyto_summary_df["assigned_cyto_target_area"] /
    cyto_summary_df["assigned_cyto_area"].replace(0, np.nan)
)

# Merge nucleus-to-cytoplasm assignment results onto the per-nucleus table
per_nucleus_df = per_nucleus_df.merge(
    assignment_df[[
        "nucleus_label",
        "assigned_cyto_label",
        "assignment_method",
        "overlap_pixels",
        "centroid_distance"
    ]],
    on = "nucleus_label",
    how = "left"
)

# Merge cytoplasmic object summaries using the assigned cytoplasmic label
per_nucleus_df = per_nucleus_df.merge(
    cyto_summary_df,
    on = "assigned_cyto_label",
    how = "left"
)

# Compute the cytoplasmic-to-nuclear TDP43 area ratio per assigned unit
per_nucleus_df["cytoplasmic_to_nuclear_target_area_ratio"] = (
    per_nucleus_df["assigned_cyto_target_area"] /
    per_nucleus_df["nuclear_target_area"].replace(0, np.nan)
)

# Print the beginning and end of the per-nucleus summary table
print(per_nucleus_df.head())
print(per_nucleus_df.tail())
```

# Run summary

This section prints a concise run-level summary so that the most important counts are visible 
without opening the exported tables. These values are useful for quick QC and to confirm that 
the workflow behaved as expected.

Metrics:

- `input_shape`: (height, width)
- `n_nuclei`: number of labeled nuclei detected in the nuclear mask
- `n_cytoplasmic_objects`: number of labeled cytoplasmic objects detected in the cytoplasmic mask
- `n_assigned_nuclei`: number of nuclei that receive a cytoplasmic assignment by either overlap or
  nearest-neighbor logic (low values may indicate segmentation mismatch or overly strict assignment criteria)
- `target_total_area`: total target-positive area after cleanup
- `target_nuclear_area`: target-positive area inside nuclei
- `target_cytoplasmic_area`: target-positive area inside the cytoplasmic mask
- `target_other_area`: target-positive area outside both main compartments

```{python run_level_summary}

# ------------------------------------------------------------------------------
# Print concise run-level summary metrics
# ------------------------------------------------------------------------------

# Collect a compact set of top-level QC metrics
run_summary = {
    "input_shape": tuple(nuclear_mask.shape),
    "n_nuclei": int(nuclear_labels.max()),
    "n_cytoplasmic_objects": int(cyto_labels.max()),
    "n_assigned_nuclei": int((assignment_df["assignment_method"] != "unassigned").sum()),
    "target_total_area": int(target_clean.sum()),
    "target_nuclear_area": int(target_nuclear_mask.sum()),
    "target_cytoplasmic_area": int(target_cytoplasmic_mask.sum()),
    "target_other_area": int(target_other_mask.sum())
}

# Print each run-level metric
for key, value in run_summary.items():
    print(key, "=", value)
```

# Visualizing compartment overlap {.tabset}

This section generates QC figures for the key masks and overlap relationships.

These plots provide a visual check that:

- nuclear and cytoplasmic compartments look reasonable
- TDP43 cleanup did not distort the mask unexpectedly
- the overlap logic places TDP43 signal into the expected compartments

## Individual masks

```{python qc_plots}

# ------------------------------------------------------------------------------
# Generate and link QC figures produced by the Python helper functions
# ------------------------------------------------------------------------------

mask_dic = {'nuclear': nuclear_mask, 'cytoplasmic': cytoplasmic_mask, 'target': target_clean}
path_dic = {}

for key, value in mask_dic.items():
    out_path = f"{r.out_dir}/{key}_mask.png"
    plt.figure(figsize=(8, 8))
    plt.imshow(value, cmap='gray')
    plt.xticks([])
    plt.yticks([])
    plt.savefig(out_path, dpi=600, bbox_inches="tight")
    plt.close()
    path_dic[key] = out_path
```

```{r individual_qc_plots, results='asis'}
# Build a table of QC image links for easy browsing in the rendered document
data.frame(
    link = unlist(py$path_dic)
) %>%
rownames_to_column('mask') %>%
mutate(link = paste0("[", link, "](", link, ")")) %>%
knitr::kable()
```

## Mask overlay

Color map:

- nuclear mask: dodgerblue
- cytoplasmic mask: orange
- target mask: violet

```{python mask_overlay}
out_path = f"{r.out_dir}/mask_overlay.png"
overlay_plot(nuclear_mask, cytoplasmic_mask, target_clean, out_path)
```

```{r mask_overlay_link, results='asis'}
link_output(py$out_path)
```

## Target-compartment overlay

Color map:

- target inside the nuclear mask: dodgerblue
- target inside the cytoplasmic mask: orange
- target outside the nuclear and cytoplasmic mask: violet

```{python target_compartment_overlay, eval=FALSE}
out_path = f"{r.out_dir}/target_compartment_overlay.png"
overlay_plot(target_nuclear_mask, target_cytoplasmic_mask, target_other_mask, out_path)
```

```{r target_compartment_overlay_link, results='asis'}
link_output(py$out_path)
```

# Summary tables

This section exports the main result tables as tab-separated files so they can be inspected in Excel,
R, or Python later.

The exported tables capture:

- TDP43 input connected-component summary
- estimated TDP43 cleanup parameters
- whole-image compartment summary
- nucleus-to-MAP2 assignment table
- per-nucleus overlap summary

```{r save_output_tables, results='asis'}

# ------------------------------------------------------------------------------
# Write key summary tables to disk and render a link table
# ------------------------------------------------------------------------------

# Collect all tables that should be exported as TSV files
df_list <- list(
    target_input_summary = py$target_initial_df,
    target_parameter_summary = py$target_param_est,
    compartment_summary = py$compartment_summary_df,
    nucleus_assignment_summary = py$assignment_df,
    per_nucleus_summary = py$per_nucleus_df
) %>%
# Clean each column if necessary
mclapply(function(df) {
    df %>%
        as.data.frame() %>%
        mutate_if(is.list, unlist)

})

# Initialize a table that will store output file links
path_df <- data.frame(
    table = names(df_list),
    link = ""
)

# Write each table to disk and record its output path
for (name in names(df_list)) {
    tsv <- file.path(out_dir, paste0(name, ".tsv"))
    df <- df_list[[name]]

    # Remove unassigned objects
    if (name %in% c('nucleus_assignment_summary', 'per_nucleus_summary')) {
        df <- df %>%
            dplyr::filter(assignment_method != "unassigned")
    }

    # Save as TSV
    write.table(
        df,
        tsv,
        quote=FALSE,
        row.names=FALSE,
        col.names=TRUE,
        sep = "\t"
    )
    path_df[path_df$table == name, "link"] <- tsv
}

# Render the exported-table link summary
knitr::kable(
    path_df %>%
        mutate(link = paste0("[", link, "](", link, ")"))
)
```

# Saving the `ImageContainer` object

The final `ImageContainer` object contains the following layers:

- `nuclear_mask`: cleaned nuclear mask (e.g. DAPI)
- `cytoplasmic_mask`: cleaned cytoplasmic mask (e.g. MAP2)
- `target_clean`: cleaned target mask (e.g. TDP43)
- `target_nuclear`: target in nucleus (e.g. nuclear TDP43)
- `target_cytoplasmic`: target in cytoplasm (e.g. cytoplasmic TDP43)
- `target_other`: target unassigned

```{python add_to_imagecontainer}

# ------------------------------------------------------------------------------
# Add derived masks back into the ImageContainer
# ------------------------------------------------------------------------------

# reticulate::repl_python()
# Add the nuclear mask as a new layer
img_dic = {
    "nuclear_mask": nuclear_mask,
    "cytoplasmic_mask": cytoplasmic_mask,
    "target_clean": target_clean,
    "target_nuclear": target_nuclear_mask,
    "target_cytoplasmic": target_cytoplasmic_mask,
    "target_other": target_other_mask
}

for key, value in img_dic.items():
    img.add_img(value, layer=key)

# Delete unnecessary layers
layers_del = [l for l in list(img) if l not in list(img_dic.keys())]
print("The following layers are deleted: ", layers_del)
for l in layers_del:
    del img[l]

# Print the layer names now present in the output object
print("Derived layers added to ImageContainer")
print(img)

# Write the output ImageContainer to the configured path
img.save(r.output_obj)

# Print the saved output path
print(r.output_obj)
```

# Session info

```{r session_info, collapse=FALSE}
sessionInfo()
```



