---
title: "Visium spatial transcriptomics"
author: "Mira Sohn"
output:
    html_document:
        code_folding: hide
        df_print: paged
        toc: true
        toc_float: true
        toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE,
                      message=FALSE,
                      echo=FALSE,
                      cache.lazy=FALSE)
```

Last run: `r date()`

This workflow is designed for exploratory analysis on the Visium dataset from the human thalamus 
([Schulmann et al.](https://www.biorxiv.org/content/10.1101/2024.09.03.611112v1)).

```{r libraries}
# Libraries
library(parallel)
library(tidyverse)
library(ggplot2)
library(plotly)
library(Seurat)
library(patchwork)
library(sceasy)
library(reticulate)
library(spacexr)
library(SummarizedExperiment)
library(SpatialExperiment)
```

```{r additional_options}
# Additional options
set.seed(2570)
source('helpers.R')
use_condaenv("../../menv")

# Specify the number of available cores
n_cores <- future::availableCores()
options(future.globals.maxSize=500000 * 1024^2,   # 500G (1000 * 1024^2 = 1G)
        mc.cores=n_cores,
        future.rng.onMisuse="ignore")

# Import scanpy in R
sc <- import("scanpy")
```

```{python ppackages}
import anndata as ad
import scanpy as sc
import pandas as pd
import numpy as np
import random
import os

random.seed(1516)
```

```{r config}

# --------------------------------------------------------------------------------
# This chunk specifies input paths and variables in R
# --------------------------------------------------------------------------------

# Paths to input datasets
input_path <- "../../input/Schulmann_visium/Visium_thal_seurat.rds"
input_h5ad <- "../../input/Schulmann_visium/combined_reclustered.h5ad"

# Path to output directory
out_dir <- "visium-thalamus-allcells"


# Column names for cell subtypes
celltype_col <- "annotation_06012026"

# Filter neurons?
filter_neurons <- FALSE

# Column name for disease conditions
selection_col <- "general_disease"

# Filter specific barcodes based on the `selection_col` column in the metadata?
# Set to NULL to disable
barcode_selection <- "Control"

# Upper bound of cell number. Disable by setting it to NULL
ncell_upper_bound <- 12000

# Specify genes of interest
gene_set <- c(
    "SYT1", "GAD2", "SOX14", "SLC17A6", "RNF220", "LAMP5",
    "BCL11B", "TUBB3", "GRN", "ATP1B1", "HSP90AA1", "HSP90AB1",
    "NEFL", "CLU", "GAPDH", "PSAP", "SERINC1", "TUBA1B",
    "PRNP", "MEG3", "NRG3", "MALAT1", "KCNIP4", "NALF1",
    "PTPRD", "ROBO1", "MAGI2", "FTX", "LRP1B", "CALB1",
    "NR2F2", "ERBB4", "PVALB", "CALB2", "MEIS2",
    "NPY", "GAD1", "LHX6", "SST", "ADARB2",
    "VIP", "CXCL14", "TCF7L2", "FOXP2", "NRP1",
    "HTR2C", "IL1RAPL2"
)

# Specify colors for plots
plot_colors <- c(
    Astrocyte="#FFD700",
    BVC="#4DBBD5",
    `ExNeu RNF220-`="#F4F1DE",
    `ExNeu RNF220+`="#E64B35",
    `InNeu SOX14-`="#00A087",
    `InNeu SOX14+`="#8491B4",
    OL="#91D1C2",
    OPC="#7E6148",
    Microglia="#B09C85",
    PVM="#DC0000",
    CP="#7A21DD",
    Lymphocyte="#FFD400",
    Undetermined="#000000"
)

# Create directories if missing
for (p in c(out_dir)) {
    if (! dir.exists(p)) { dir.create(p, recursive=TRUE) }
}
```

# Loading input data

## Single-cell obj

```{r import_dataset}

# --------------------------------------------------------------------------------
# This chunk imports the input dataset
# --------------------------------------------------------------------------------

# Import visium dataset
seurat <- readRDS(input_path)

print(paste("Input file path:", input_path))
print(seurat)
```

## Staining images

```{r hne_images, results='asis'}
p <- SpatialFeaturePlot(
    seurat,
    features="nCount_Spatial",
    alpha=c(0, 0)) + 
    theme(legend.position="none")
hne_svg <- file.path(out_dir, "HnE_image.svg")
ggsave(hne_svg, p, device="svg")
print(p)
link_plot(hne_svg)
```

# Genes of interest {.tabset}

The following genes of interest are explored across the image set. Gene not present
in the current dataset are labeled as 'absent genes'.

```{r genes_of_interest}

# --------------------------------------------------------------------------------
# This chunk cleans genes of interest
# --------------------------------------------------------------------------------

# Retrieve the normalized count matrix
sct_mat <- GetAssayData(seurat, assay="SCT", layer="data")

# Retrieve all gene symbols
all_genes <- rownames(sct_mat)

# Remove genes of interest if absent and print
genes_absent <- gene_set[!gene_set %in% all_genes]
genes_present <- setdiff(gene_set, genes_absent)
print(paste0("- Genes of interest: ", paste0(gene_set, collapse=", ")))
print(paste0("- Genes present: ", paste0(genes_present, collapse=", ")))
print(paste0("- Genes absent: ", paste0(genes_absent, collapse=", ")))

# Clean gene_set to exclude absent genes
gene_set <- intersect(gene_set, genes_present)
```

Following the methodology of [Schulmann et al.](https://www.biorxiv.org/content/10.1101/2024.09.03.611112v1),
we utilize normalized expression data generated via the SCTransform method 
([Hafemeister et al.](https://link.springer.com/article/10.1186/s13059-019-1874-1)).

```{r spatial_featureplot, results='asis'}

# --------------------------------------------------------------------------------
# This chunk explores the spatial expression of genes of interest
# --------------------------------------------------------------------------------

for (gene in gene_set) {
    cat('##', gene, '\n\n')
    p <- SpatialFeaturePlot(
        seurat,
        features=gene,
        pt.size.factor=3.5)
    print(p)
    plot_path <- file.path(
        out_dir,
        paste0(gene, '_spatialfeature.svg'))
    ggsave(plot_path,
           p,
           device="svg")
    link_plot(plot_path)
    cat('\n\n')
}
```

# Deconvolution

We deconvolve spot-level cell subtypes using 
the [Robust Cell Type Decomposition (RCTD, Cable et al.)](https://www.nature.com/articles/s41587-021-00830-w).
This analysis is performed using raw counts from annotated datasets listed below:

## Load annotated single-cell datasets

```{python load_inhouse}

# --------------------------------------------------------------------------------
# This chunk loads annotated AnnData obj
# --------------------------------------------------------------------------------

# Import input annotated AnnData obj
adata = sc.read_h5ad(r.input_h5ad)

print("Input annotated dataset loaded:", r.input_h5ad)
print(adata)
```

All input cells:

```{r input_cell_counts, results='asis'}
py$adata[['obs']][[celltype_col]] %>%
    table %>%
    as.data.frame() %>%
    set_names(c('celltype', 'nCells')) %>%
    knitr::kable()
```

```{python filter_neurons}

# --------------------------------------------------------------------------------
# This chunk optionally filter neurons
# --------------------------------------------------------------------------------

# Filter neurons if specified
if r.filter_neurons:
    neuron_mask = adata.obs[r.celltype_col].str.contains('Neu', na=False)
    adata = adata[neuron_mask, :]
    print("Input cells are narrowed to neurons.")

print(adata)
```

```{python filter_by_condition}

# --------------------------------------------------------------------------------
# This chunk optionally narrow cells to those from conditions of interest
# --------------------------------------------------------------------------------

if r.barcode_selection is not None:
    barcodes = adata.obs[r.selection_col].isin([r.barcode_selection])
    adata = adata[barcodes, :]
    print("Barcodes aligning the following conditions are filtered:",
        r.barcode_selection,
        "in",
        r.selection_col)
    print(adata)
```

```{r barcode_cell_counts, results='asis'}
if (!is.null(barcode_selection)) {
    py$adata[['obs']][[celltype_col]] %>%
        table %>%
        as.data.frame() %>%
        set_names(c('celltype', 'nCells')) %>%
        knitr::kable() %>%
        print()
}
```

```{python apply_upperbound}

# --------------------------------------------------------------------------------
# This chunk optionally filter cells to retain a specified number of cells
# --------------------------------------------------------------------------------

# reticulate::repl_python()
# Retain randomly selected N cells the most, if specified
if r.ncell_upper_bound is not None:
    max_ncells = int(r.ncell_upper_bound)
    print("In each cell subtype, cells are randomly selected to retain only", max_ncells, "cells.")
    obs = adata.obs.copy()
    rng = np.random.default_rng(42)
    retained_cells = []
    for celltype, idx in obs.groupby(r.celltype_col).groups.items():
        idx = np.array(list(idx))
        if len(idx) > max_ncells:
            keep = rng.choice(idx, size=max_ncells, replace=False)
        else:
            keep = idx
        print(celltype, "---- from", len(idx), "to", len(keep))
        retained_cells.extend(keep)

    adata = adata[retained_cells]
```

```{r filtered_cell_counts, results='asis'}
if (filter_neurons) {
    py$adata[['obs']][[celltype_col]] %>%
        table %>%
        as.data.frame() %>%
        set_names(c('celltype', 'nCells')) %>%
        knitr::kable() %>%
        print()
}
```

To transfer annotations from the annotated dataset to the Visium dataset, the annotated dataset is converted
into a Seurat obj.

## Doublet mode {.tabset}

Annotation transfer with the doublet mode option computes in the following two cell subtypes:

- **first_type**: first predicted celltype
- **second_type**: second predicted celltype, if exists

```{python retreive_adata}
# Prep features and barcodes as a list for the next step
all_features = list(adata.var_names)
all_barcodes = list(adata.obs.index)
```

```{r adata_to_seurat, cache=TRUE}

# --------------------------------------------------------------------------------
# This chunk converts the AnnData obj to Seurat and normalize
# --------------------------------------------------------------------------------

# Prep an input count matrix by extracting from the concatenated AnnData
ad_mat <- t(py$adata[['layers']][['counts']])
rownames(ad_mat) <- py$all_features
colnames(ad_mat) <- py$all_barcodes

# Build a new Seurat obj using raw counts
ad_seurat <- CreateSeuratObject(
    counts=ad_mat,
    meta.data=py$adata[['obs']]
) %>%
# Filter nonzero-count cells
subset(nCount_RNA > 0)

print(ad_seurat)
```

```{r prep_rctd, cache=TRUE, dependson='adata_to_seurat'}

# --------------------------------------------------------------------------------
# This chunk preps input for RCTD
# --------------------------------------------------------------------------------

# Specify ID col
Idents(ad_seurat) <- celltype_col

# Prep raw counts from the annotated dataset
counts <- ad_seurat[['RNA']]$counts

# Prep annotation from the annotated dataset
cluster <- as.factor(ad_seurat@meta.data[[celltype_col]])
names(cluster) <- colnames(ad_seurat)

# Prep nUMI from the annotated dataset
nUMI <-ad_seurat@meta.data[['nCount_RNA']]
names(nUMI) <- colnames(ad_seurat)

# Create a new `Reference`
reference <- Reference(
    counts,
    cluster,
    nUMI,
    n_max_cells=ncell_upper_bound)

# Prep a list of query across images
query_list <- mclapply(1:length(Images(seurat)), function(i) {

    # Extract raw counts per image
    ct <- GetAssayData(
        seurat,
        assay='Spatial',
        layer=Layers(seurat[['Spatial']])[i])
    # Extract coordinates per image
    cd <- GetTissueCoordinates(
        seurat,
        image=Images(seurat)[i])
    # Clean column names for the coordinate table
    colnames(cd) <- c('x', 'y')
    cd[is.na(colnames(cd))] <- NULL
    # Create a new spatial obj
    obj <- SpatialRNA(
        coords=cd,
        counts=ct,
        colSums(ct))
    return(obj)
}) %>%
# Name the list
set_names(Images(seurat))
```

```{r run_rctd, cache=TRUE, dependson='prep_rctd'}

# --------------------------------------------------------------------------------
# This chunk runs RCTD
# --------------------------------------------------------------------------------

# Create a list of RCTD obj
rctd_list <- lapply(
    names(query_list), function(name) {
        # Create a new RCTD obj
        create.RCTD(query_list[[name]],
                    reference,
                    max_cores=n_cores) %>%
            # Run RCTD
            run.RCTD(doublet_mode="doublet")
    }) %>%
    set_names(names(query_list))
```

### Per-celltype {.tabset}

The top 2 cell subtypes are visualized separately.

```{r prep_doublet_separated}

# --------------------------------------------------------------------------------
# This chunk preps to visualize RCTD results from the doublet mode separately
# --------------------------------------------------------------------------------

# Add RCTD results to the metadata of Visium obj
rctd_df <- lapply(rctd_list, function(res) {
    df_names <- c('results_df', 'weights')
    results_df <- res@results[['results_df']]
    weights_df <- res@results[['weights_doublet']]
    colnames(weights_df) <- c('first_weight', 'second_weight')
    return(cbind(results_df, weights_df))
}) %>%
bind_rows()

# Add the transferred annotation to the metadata of Visium seurat obj
seurat <- AddMetaData(seurat, metadata=rctd_df)
```

```{r vis_doublet_separated, results='asis'}

# Plot and save
for (name in Images(seurat)) {
    cat('####', name, '\n\n')
    p_list <- mclapply(c('first_type', 'second_type'), function(anno) {
        SpatialDimPlot(seurat,
                       group.by=anno,
                       pt.size.factor=3.5,
                       cols=plot_colors,
                       images=name) +
        theme(legend.margin=margin(0, 0, 0, 0), 
              legend.title=element_text(size=8),
              legend.text=element_text(size=8))

    })
    p <- p_list[[1]] | p_list[[2]]
    print(p)
    p_path <- file.path(out_dir, paste0(name, '_doublet_sep.svg'))
    ggsave(p_path, p, device="svg")
    link_plot(p_path)
}
```

### Per-spot {.tabset}

Per-spot proportions are visualized.

```{r doublet_both, results='asis'}

# --------------------------------------------------------------------------------
# This chunk visualizes RCTD results from the doublet mode in a single plot
# --------------------------------------------------------------------------------

coords_list <- list()
for (name in Images(seurat)) {
    cat('####', name, '\n\n')

    # Prep input data frame
    coords <- GetTissueCoordinates(seurat, image=name) %>%
        set_names(c('x', 'y'))
    coords_list[[name]] <- coords
    # Reshape the data frame
    celltype_df <- seurat@meta.data[,
        c('first_type',
          'second_type',
          'first_weight',
          'second_weight')] %>%
        rownames_to_column('barcode') %>%
        pivot_wider(names_from='first_type',
                    values_from='first_weight',
                    values_fill=0,
                    names_repair="minimal",
                    values_fn=sum) %>%
        pivot_wider(names_from='second_type',
                    values_from='second_weight',
                    values_fill=0,
                    names_repair="minimal",
                    values_fn=sum)

    # Aggregate duplicated columns
    cell_df <- sapply(
        split.default(celltype_df[,-1], names(celltype_df)[-1]),
        rowSums)
    rownames(cell_df) <- celltype_df$barcode

    # Imput missing values with zero
    coords <- cbind(coords, cell_df[rownames(coords),])

    # Plot celltype proportions by coordinate
    p <- ggplot() +
        scatterpie::geom_scatterpie(
            data=coords,
            cols=colnames(coords)[-c(1:2)],
            color=NA,
            aes(x=x, y=y, r=1)) +
        coord_equal() +
        custom_theme() +
        coord_flip() + 
        scale_x_reverse() +
        scale_fill_manual(values=plot_colors)

    print(p)

    # Save and link
    p_path <- file.path(
        out_dir,
        paste0(name, '_doublet_both.svg'))
    ggsave(p_path, p, device="svg")
    link_plot(p_path)
}
```

## Multi mode {.tabset}

Since two cell subtypes do not adequately represent all transferred celltypes, we assess the proportion of 
cell subtypes across the annotated celltypes.

```{r run_rctd_multi, cache=TRUE, dependson='run_rctd'}

# --------------------------------------------------------------------------------
# This chunk runs multi-mode RCTD
# --------------------------------------------------------------------------------

# Create a list of RCTD obj
multi_list <- lapply(
    names(query_list), function(name) {
        # Create a new RCTD obj
        create.RCTD(query_list[[name]], reference, max_cores=n_cores) %>%
            # Run RCTD
            run.RCTD(doublet_mode="multi")
    }) %>%
    set_names(names(query_list))
```

```{r celltype_weights, cache=TRUE, dependson='run_rctd_multi'}

# --------------------------------------------------------------------------------
# This chunk extracts and cleans the multi-mode results
# --------------------------------------------------------------------------------

# Update the metadata of the Visium Seurat obj to include columns for cell subtypes
all_celltypes <- levels(reference@cell_types)

# Extract per-celltype weights across coordinates and images
weights_df <- lapply(multi_list, function(o) {

    # Retrieve coordinate barcodes
    coords_rctd <- colnames(o@spatialRNA@counts)

    # Create a data frame consisting of weights across cell subtypes
    weights_rctd <- mclapply(o@results, function(res) {

        # Clean results
        df_i <- as.data.frame(res$sub_weights) %>%
            rownames_to_column('celltype') %>%
            set_names(c('celltype', 'weights')) %>%
            pivot_wider(names_from='celltype', values_from='weights') %>%
            as.data.frame()

        # Add columns for celltypes that are missing in each barcode
        col_placeholder <- setdiff(all_celltypes, colnames(df_i))
        for (c in col_placeholder) {
            df_i[[c]] <- 0
        }

        # Reorder columns
        df_i <- df_i[, all_celltypes]
        return(df_i) 
        }) %>%
        bind_rows()

    rownames(weights_rctd) <- coords_rctd
    return(weights_rctd) 
    }) %>%
    bind_rows()

# Reorder rows to have them aligned with those in the Seurat obj
weights_df <- weights_df[colnames(seurat),]

# Update the Seurat obj to include deconvolution results
seurat@meta.data <- cbind(seurat@meta.data, weights_df)
```

### Proportion piescatter plot {.tabset}

```{r vis_rctd_multi, results='asis'}

# --------------------------------------------------------------------------------
# This chunk preps plots the proportion of cell subtypes computed using the multi-node RCTD
# --------------------------------------------------------------------------------

all_celltypes <- colnames(weights_df)

for (name in Images(seurat)) {
    cat('####', name, '\n\n')

    # Prep input data frame
    coords <- coords_list[[name]]
    coords <- cbind(
        coords,
        weights_df[rownames(coords), ])
    coords_list[[name]] <- coords

    # Plot celltype proportions by coordinate
    p <- ggplot() +
        scatterpie::geom_scatterpie(
            data=coords,
            cols=all_celltypes,
            color=NA,
            aes(x=x, y=y, r=1)) +
        coord_equal() +
        custom_theme() +
        coord_flip() + 
        scale_x_reverse() +
        scale_fill_manual(values=plot_colors)

    print(p)

    # Save and link
    p_path <- file.path(
        out_dir,
        paste0(name, '_name_proportion.svg'))
    ggsave(p_path, p, device="svg")
    link_plot(p_path)
}
```

### Density heatmap {.tabset}

Here, we visualize the proportion of individual cell subtypes ranging from 0 to 1. Note that the following
pseudocount is added to all proportions so we transform to a log2 scale. The pseudocount is calculated
as the minimum proportion greater than zero within all input cells.

```{r pseudocount}
# Compute the minimum proportion greater than zero
all_prop <- unlist(coords[, all_celltypes])
nonzero_min <- min(all_prop[all_prop > 0])
print(paste0("A pseudocount of ", nonzero_min, " is added..."))
```

```{r vis_rctd_individual, results='asis', fig.width=12}

# --------------------------------------------------------------------------------
# This chunk preps plots the proportion of cell subtypes computed using the multi-node RCTD
# --------------------------------------------------------------------------------

for (name in Images(seurat)) {

    cat('####', name, '{.tabset}\n\n')

    # Extract the coordinate data frame for each celltype add the pseudocount
    coords <- coords_list[[name]] %>%
        mutate_at(all_celltypes, function(x) x + nonzero_min)

    # Calculate log2-transformed min and max proportions
    l2_props <- coords[, all_celltypes] %>%
        unlist() %>%
        log2()
    l2_min_max <- c(min(l2_props), max(l2_props))

    for (celltype in all_celltypes) {

        # Prep input data frame
        cat('#####', celltype, '{.tabset}\n\n')
        # Filter a single celltype
        coords_i <- coords[, c('x', 'y', celltype)]
        colnames(coords_i)[3] <- 'proportion'

        # Is this celltype zero %?
        zero_prop <- ifelse(
            sum(coords_i[['proportion']] - nonzero_min) == 0,
            TRUE,
            FALSE)

        for (prop in c('fixed', 'adjusted')) {

            cat('######', prop, 'scale\n\n')
            # Plot celltype proportions by coordinate, with and without a log2-transformation
            p1 <- ggplot(coords_i, aes(x=x, y=y, color=proportion)) +
                geom_point(size=0.5) +
                ggtitle('Raw proportion')
            p2 <- ggplot(coords_i, aes(x=x, y=y, color=log2(proportion))) +
                geom_point(size=0.5) +
                ggtitle('Log2-transformed proportion')
            p_list <- list(raw=p1, l2=p2) %>%
                map(~.x +
                    custom_theme() + 
                    coord_flip() + 
                    scale_x_reverse() 
                )
            p_list <- lapply(names(p_list), function(pname) {
                p <- if (prop == "fixed" & pname == "raw") {
                    p_list[[pname]] + 
                        scale_color_gradient(
                            low='yellow',
                            high='red',
                            limits=c(0, 1))
                } else if (prop == "fixed" & pname == "l2") {
                    p_list[[pname]] + 
                        scale_color_gradient(
                            low='yellow',
                            high='red',
                            limits=c(l2_min_max[1], l2_min_max[2]))
                } else {
                    p_list[[pname]] + 
                        scale_color_gradient(low='yellow', high='red')
                }
                return(p)
                })

            # Override the dot color to yellow if the celltype computed as zero %
            if (zero_prop) {
                print(paste0("Computed proportion of ", name, " ", celltype, ": 0"))
                p_list <- map(p_list, ~.x + geom_point(color='yellow'))
            }

            p_final <- p_list[[1]] | p_list[[2]]
            print(p_final)
            cat('\n\n')

            # Save and link
            p_path <- file.path(
                out_dir,
                paste0(name, '_', celltype, '_', prop, '.svg'))
            ggsave(p_path, p_final, device="svg")
            link_plot(p_path)
        }
    }
}
```

# Updated Seurat obj

```{r save}

# --------------------------------------------------------------------------------
# This chunk saves the Visium Seurat obj with RCTD-transferred annotation
# --------------------------------------------------------------------------------

out_rds <- file.path(out_dir, "updated_seurat_visium.rds")
saveRDS(seurat, out_rds, compress=FALSE)

print(paste("Spatial obj saved:", out_rds))
```


# Session info

```{r session_info, collapse=FALSE}
sessionInfo()
```

