---
title: "Visium spatial transcriptomics"
author: "Mira Sohn"
output:
    html_document:
        code_folding: hide
        df_print: paged
        toc: true
        toc_float: true
        toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE,
                      message=FALSE,
                      echo=FALSE,
                      cache.lazy=FALSE)
```

Last run: `r date()`

This workflow is designed for exploratory analysis on the Visium dataset from the human thalamus 
([Schulmann et al.](https://www.biorxiv.org/content/10.1101/2024.09.03.611112v1)).

```{r libraries}
# Libraries
library(parallel)
library(tidyverse)
library(ggplot2)
library(plotly)
library(Seurat)
library(patchwork)
library(sceasy)
library(reticulate)
library(spacexr)
library(SummarizedExperiment)
library(SpatialExperiment)
```

```{r additional_options}
# Additional options
set.seed(2570)
source('helpers.R')
use_condaenv("../../menv")

# Specify the number of available cores
n_cores <- future::availableCores()
options(future.globals.maxSize=500000 * 1024^2,   # 500G (1000 * 1024^2 = 1G)
        mc.cores=n_cores,
        future.rng.onMisuse="ignore")

# Import scanpy in R
sc <- import("scanpy")
```

```{python ppackages}
import anndata as ad
import scanpy as sc
import pandas as pd
import numpy as np
import random
import os

random.seed(1516)
```

```{r config}

# --------------------------------------------------------------------------------
# This chunk specifies input paths and variables in R
# --------------------------------------------------------------------------------

# Paths to input datasets
input_path <- "../../input/Schulmann_visium/Visium_thal_seurat.rds"
h5ad_list <- list(
    InNeuSOX14neg="../../input/thalamus_neurons_reannotated/InNeuSOX14neg_adata.h5ad",
    InNeuSOX14pos="../../input/thalamus_neurons_reannotated/InNeuSOX14pos_adata.h5ad",
    ExNeuRNF220neg="../../input/thalamus_neurons_reannotated/ExNeuRNF220neg_adata.h5ad",
    ExNeuRNF220pos="../../input/thalamus_neurons_reannotated/ExNeuRNF220pos_adata.h5ad",
    AST="../../input/thalamus_glials_reannotated/Astrocyte_adata.h5ad",
    OPC="../../input/thalamus_glials_reannotated/OPC_adata.h5ad",
    OL="../../input/thalamus_glials_reannotated/OL_adata.h5ad",
    MG="../../input/thalamus_glials_reannotated/Microglia_adata.h5ad",
    PVM="../../input/thalamus_glials_reannotated/PVM_adata.h5ad",
    Lympho="../../input/thalamus_glials_reannotated/Lymphocyte_adata.h5ad",
    BVC="../../input/thalamus_glials_reannotated/BVC_adata.h5ad",
    CP="../../input/thalamus_glials_reannotated/CP_adata.h5ad"
)

# Path to output directory
out_dir <- "visium-thalamus-allcells"

# Column names for cell subtypes and disease conditions
celltype_col <- "annotation_04142026"
selection_col <- "general_disease"

# Filter specific barcodes based on the `selection_col` column in the metadata?
# Set to NULL to disable
barcode_selection <- NULL

# Subsampling proportion
# NOTE: specify 1 to use all barcodes
sample_prop <- 0.2

# Specify genes of interest
gene_list <- list(
    set1=c('SYT1', 'GAD2', 'SOX14'),
    set2=c('SYT1', 'SLC17A6', 'RNF220'),
    set3=c('LAMP5', 'BCL11B', 'TUBB3', 'GRN'),
    set4=c(
        'ATP1B1',
        'HSP90AA1',
        'HSP90AB1',
        'NEFL',
        'CLU',
        'GAPDH',
        'PSAP',
        'SERINC1',
        'TUBA1B',
        'PRNP',
        'MEG3',
        'MALAT1',
        'NRG3',
        'MALAT1',
        'NRG3',
        'KCNIP4',
        'NALF1',
        'PTPRD',
        'ROBO1',
        'MAGI2',
        'FTX',
        'LRP1B',
        'CALB1'
    )
)

# Annotation column names (specify to NULL if unnecessary)
celltype_col <- "annotation_04142026"
celltype_col_glial <- 'renamed_annotation_cell_type_post_outlier_removal'

# Specify colors for plots
# plot_colors <- c('#F4F1DE', '#E07A5F', '#3D405B', '#4ecdc4')
plot_colors <- c(
    "#3C5488",
    "#4DBBD5",
    "#F4F1DE",
    "#E64B35",
    "#00A087",
    "#8491B4",
    "#91D1C2",
    "#7E6148",
    "#B09C85",
    "#DC0000",
    "#7A21DD",
    "#FFD400"
)

# Collapse all glial celltypes into "Glia"?
collapse_glia <- FALSE

# Create directories if missing
for (p in c(out_dir)) {
    if (! dir.exists(p)) { dir.create(p, recursive=TRUE) }
}
```

# Loading input data

## Single-cell obj

```{r import_dataset}

# --------------------------------------------------------------------------------
# This chunk imports the input dataset
# --------------------------------------------------------------------------------

# Import visium dataset
seurat <- readRDS(input_path)

print(paste("Input file path:", input_path))
print(seurat)
```

## Staining images

```{r hne_images, results='asis'}
p <- SpatialFeaturePlot(
    seurat,
    features="nCount_Spatial",
    alpha=c(0, 0)) + 
    theme(legend.position="none")
hne_svg <- file.path(out_dir, "HnE_image.svg")
ggsave(hne_svg, p, device="svg")
print(p)
link_plot(hne_svg)
```

# Genes of interest {.tabset}

The following genes of interest are explored across the image set. Gene not present
in the current dataset are labeled as 'absent genes'.

```{r genes_of_interest}

# --------------------------------------------------------------------------------
# This chunk cleans genes of interest
# --------------------------------------------------------------------------------

# Retrieve the normalized count matrix
sct_mat <- GetAssayData(seurat, assay="SCT", layer="data")

# Retrieve all gene symbols
all_genes <- rownames(sct_mat)

# Remove genes of interest if absent and print
gene_set <- unique(unlist(gene_list))
genes_absent <- gene_set[!gene_set %in% all_genes]
genes_present <- setdiff(gene_set, genes_absent)
print(paste0("- Genes of interest: ", paste0(gene_set, collapse=", ")))
print(paste0("- Genes present: ", paste0(genes_present, collapse=", ")))
print(paste0("- Genes absent: ", paste0(genes_absent, collapse=", ")))

# Clean gene_set to exclude absent genes
gene_set <- intersect(gene_set, genes_present)
```

Following the methodology of [Schulmann et al.](https://www.biorxiv.org/content/10.1101/2024.09.03.611112v1),
we utilize normalized expression data generated via the SCTransform method 
([Hafemeister et al.](https://link.springer.com/article/10.1186/s13059-019-1874-1)).

```{r spatial_featureplot, results='asis'}

# --------------------------------------------------------------------------------
# This chunk explores the spatial expression of genes of interest
# --------------------------------------------------------------------------------

for (gene in gene_set) {
    cat('##', gene, '\n\n')
    p <- SpatialFeaturePlot(
        seurat,
        features=gene,
        pt.size.factor=3.5)
    print(p)
    plot_path <- file.path(
        out_dir,
        paste0(gene, '_spatialfeature.svg'))
    ggsave(plot_path,
           p,
           device="svg")
    link_plot(plot_path)
    cat('\n\n')
}
```

# Deconvolution

We deconvolve spot-level cell subtypes using 
the [Robust Cell Type Decomposition (RCTD, Cable et al.)](https://www.nature.com/articles/s41587-021-00830-w).
This analysis is performed using raw counts from annotated datasets listed below:

## Load annotated single-cell datasets

```{python load_inhouse}

# --------------------------------------------------------------------------------
# This chunk loads annotated AnnData obj
# --------------------------------------------------------------------------------

if r.sample_prop < 1:
    print("barcode subsampling specified:", r.sample_prop, "x nBarcodes")

adatas = {}

for name, path in r.h5ad_list.items():
    # Load input AnnData
    adata_i = sc.read_h5ad(path)
    # Subsample barcodes from each AnnData obj, if specified
    if r.sample_prop < 1:
        # Extract barcodes
        barcodes = list(adata_i.obs_names)
        # Subsample barcodes
        k = max(1, int(len(barcodes) * r.sample_prop))
        indices = random.sample(range(len(barcodes)), k)
        barcodes_sub = [barcodes[i] for i in indices]
        # Subset the AnnData
        adata_sub = adata_i[barcodes_sub].copy()
        # Add the subsetted AnnData to the `adatas` dictionary
        adatas[name] = adata_sub
    # Add the entire AnnData, otherwise
    else:
        adatas[name] = adata_i

print("Input annotated dataset loaded and concatenated:")
for name, path in r.h5ad_list.items():
    print(name, "-", path)

# Concatenate all obj
adata = ad.concat(
    list(adatas.values()),
    keys=list(adatas.keys()),
    index_unique="_",
    label="AnnData",
    join="outer",
    merge="same")

print(adata)

# Prep features and barcodes as a list for the next step
all_features = list(adata.var_names)
all_barcodes = list(adata.obs.index)
```

To transfer annotations from the annotated dataset to the Visium dataset, the annotated dataset is converted
into a Seurat obj.

## Doublet mode {.tabset}

Annotation transfer with the doublet mode option computes in the following two cell subtypes:

- **first_type**: first predicted celltype
- **second_type**: second predicted celltype, if exists

```{r adata_to_seurat, cache=TRUE}

# --------------------------------------------------------------------------------
# This chunk converts the AnnData obj to Seurat and normalize
# --------------------------------------------------------------------------------

# Prep an input count matrix by extracting from the concatenated AnnData
ad_mat <- t(py$adata[['layers']][['counts']])
rownames(ad_mat) <- py$all_features
colnames(ad_mat) <- py$all_barcodes

# Prep metadata
ad_meta <- py$adata[['obs']]
if (any(is.na(ad_meta[[celltype_col]])) & !is.null(celltype_col_glial)) {
    ad_meta[[celltype_col]] <- ifelse(
        is.na(ad_meta[[celltype_col]]),
        ad_meta[[celltype_col_glial]],
        ad_meta[[celltype_col]]
    )
}

# Build a new Seurat obj using raw counts
ad_seurat <- CreateSeuratObject(
    counts=ad_mat,
    meta.data=ad_meta
) %>%
# Filter nonzero-count cells
subset(nCount_RNA > 0)

print(ad_seurat)
```

```{r select_cells, cache=TRUE, dependson='adata_to_seurat'}

# --------------------------------------------------------------------------------
# This chunk optionally selects samples of interest
# --------------------------------------------------------------------------------

if (!is.null(barcode_selection)) {
    m_df <- ad_seurat@meta.data
    barcodes_selected <- m_df[m_df[[selection_col]] %in% barcode_selection,] %>%
        rownames()
    print(paste0(length(barcodes_selected), " cells are filtered to proceed with the next steps."))
    print(paste0("These are for the following cells: ", barcode_selection, " in ", selection_col))
    ad_seurat <- subset(ad_seurat, cells=barcodes_selected)
    print(ad_seurat)
}
```

```{r prep_rctd, cache=TRUE, dependson=c('adata_to_seurat', 'select_cells')}

# --------------------------------------------------------------------------------
# This chunk preps input for RCTD
# --------------------------------------------------------------------------------

# Specify ID col
Idents(ad_seurat) <- celltype_col

# Prep raw counts from the annotated dataset
counts <- ad_seurat[['RNA']]$counts

# Prep annotation from the annotated dataset
cluster <- as.factor(ad_seurat@meta.data[[celltype_col]])
names(cluster) <- colnames(ad_seurat)

# Prep nUMI from the annotated dataset
nUMI <-ad_seurat@meta.data[['nCount_RNA']]
names(nUMI) <- colnames(ad_seurat)

# Create a new `Reference`
reference <- Reference(counts, cluster, nUMI)

# Prep a list of query across images
query_list <- mclapply(1:length(Images(seurat)), function(i) {
    # Extract raw counts per image
    ct <- GetAssayData(
        seurat,
        assay='Spatial',
        layer=Layers(seurat[['Spatial']])[i])
    # Extract coordinates per image
    cd <- GetTissueCoordinates(
        seurat,
        image=Images(seurat)[i])
    # Clean column names for the coordinate table
    colnames(cd) <- c('x', 'y')
    cd[is.na(colnames(cd))] <- NULL
    # Create a new spatial obj
    obj <- SpatialRNA(
        coords=cd,
        counts=ct,
        colSums(ct))
    return(obj)
}) %>%
# Name the list
set_names(Images(seurat))
```

```{r run_rctd, cache=TRUE, dependson='prep_rctd'}

# --------------------------------------------------------------------------------
# This chunk runs RCTD
# --------------------------------------------------------------------------------

# Create a list of RCTD obj
rctd_list <- lapply(
    names(query_list), function(name) {
        # Create a new RCTD obj
        create.RCTD(query_list[[name]], reference, max_cores=n_cores) %>%
            # Run RCTD
            run.RCTD(doublet_mode="doublet")
    }) %>%
    set_names(names(query_list))
```

### Per-celltype {.tabset}

The top 2 cell subtypes are visualized separately.

```{r doublet_separated, results='asis'}

# --------------------------------------------------------------------------------
# This chunk visualizes RCTD results from the doublet mode separately
# --------------------------------------------------------------------------------

# Add RCTD results to the metadata of Visium obj
rctd_df <- map_df(rctd_list, ~.x@results$results_df)

# Convert individual glial cells into "Glia" if specified
anno_cols <- c('first_type', 'second_type')
if (collapse_glia) {
    for (name in anno_cols) {
        rctd_df[[name]] <- map_chr(
            as.character(rctd_df[[name]]),
            ~ifelse(str_detect(.x, "Neu"), .x, "Glia"))
    }
}

# Relevel cells
anno_levels <- as.character(unique(cluster))
anno_levels <- if (collapse_glia) {
    c(anno_levels[anno_levels != "Glia"], "Glia")
} else {
    anno_levels
}

for (anno in anno_cols) {
    rctd_df[[anno]] <- factor(rctd_df[[anno]], levels=anno_levels)
}

# If uncertain, replace the annotation of the second type with "Undetermined"
if (any(rctd_df[['spot_class']] == 'doublet_uncertain')) {
    anno_levels <- c(anno_levels, "Undetermined")
    rctd_df[['second_type']] <- as.character(rctd_df[['second_type']])
    rctd_df[rctd_df[['spot_class']] == "doublet_uncertain", 'second_type'] <- "Undetermined"
    rctd_df[['second_type']] <- factor(rctd_df[['second_type']], levels=anno_levels)
    # Specify colors for individual cell subtypes
    plot_cols <- c(plot_colors, "black")
} else {
    plot_cols <- plot_colors
}
names(plot_cols) <- anno_levels

# Add the transferred annotation to the metadata of Visium seurat obj
seurat <- AddMetaData(seurat, metadata=rctd_df)

# Plot and save
for (name in Images(seurat)) {
    cat('####', name, '\n\n')
    p_list <- mclapply(c('first_type', 'second_type'), function(anno) {
        SpatialDimPlot(seurat,
                       group.by=anno,
                       pt.size.factor=3.5,
                       cols=plot_cols,
                       images=name) +
        theme(legend.margin=margin(0, 0, 0, 0), 
              legend.title=element_text(size=8),
              legend.text=element_text(size=8))

    })
    p <- p_list[[1]] | p_list[[2]]
    print(p)
    p_path <- file.path(out_dir, paste0(name, '_doublet_sep.svg'))
    ggsave(p_path, p, device="svg")
    link_plot(p_path)
}
```

### Per-spot {.tabset}

The top 2 cell subtypes are visualized across spots with a uniform weight.

```{r doublet_both, results='asis'}

# --------------------------------------------------------------------------------
# This chunk visualizes RCTD results from the doublet mode in a single plot
# --------------------------------------------------------------------------------
# Specify colors for plots
plot_cols <- c(plot_colors, "black")

coords_list <- list()

for (name in Images(seurat)) {
    cat('####', name, '\n\n')

    # Prep input data frame
    coords <- GetTissueCoordinates(seurat, image=name) %>%
        set_names(c('x', 'y'))
    coords_list[[name]] <- coords
    celltype_df <- seurat@meta.data[, c('first_type', 'second_type')]
    coords <- cbind(coords, celltype_df[rownames(coords),])
    # Extract celltypes
    top_celltypes <- as.vector(celltype_df) %>%
        unlist() %>%
        unique() %>%
        levels()

    # Add celltype columns with an weight of 0.5 if they align with any of the top 2 celltypes
    for (c in top_celltypes) {
        coords[, c] <- ifelse(
            coords$first_type == c & coords$second_type == c, 1, 
            ifelse(coords$first_type != c & coords$second_type != c, 0, 0.5))
    }

    # Plot celltype proportions by coordinate
    p <- ggplot() +
        scatterpie::geom_scatterpie(
            data=coords,
            cols=top_celltypes,
            color=NA,
            aes(x=x, y=y, r=1)) +
        coord_equal() +
        custom_theme() +
        coord_flip() + 
        scale_x_reverse() +
        scale_fill_manual(values=plot_cols)

    print(p)

    # Save and link
    p_path <- file.path(
        out_dir,
        paste0(name, '_doublet_both.svg'))
    ggsave(p_path, p, device="svg")
    link_plot(p_path)
}
```

## Multi mode {.tabset}

Since two cell subtypes do not adequately represent all transferred celltypes, we assess the proportion of 
cell subtypes across the annotated celltypes.

```{r run_rctd_multi, cache=TRUE, dependson='run_rctd'}

# --------------------------------------------------------------------------------
# This chunk runs multi-mode RCTD
# --------------------------------------------------------------------------------

# Create a list of RCTD obj
multi_list <- lapply(
    names(query_list), function(name) {
        # Create a new RCTD obj
        create.RCTD(query_list[[name]], reference, max_cores=n_cores) %>%
            # Run RCTD
            run.RCTD(doublet_mode="multi")
    }) %>%
    set_names(names(query_list))
```

```{r celltype_weights, cache=TRUE, dependson='run_rctd_multi'}

# --------------------------------------------------------------------------------
# This chunk extracts and cleans the multi-mode results
# --------------------------------------------------------------------------------

# Update the metadata of the Visium Seurat obj to include columns for cell subtypes
all_celltypes <- levels(reference@cell_types)

# Extract per-celltype weights across coordinates and images
weights_df <- lapply(multi_list, function(o) {

    # Retrieve coordinate barcodes
    coords_rctd <- colnames(o@spatialRNA@counts)

    # Create a data frame consisting of weights across cell subtypes
    weights_rctd <- mclapply(o@results, function(res) {

        # Clean results
        df_i <- as.data.frame(res$sub_weights) %>%
            rownames_to_column('celltype') %>%
            set_names(c('celltype', 'weights')) %>%
            pivot_wider(names_from='celltype', values_from='weights') %>%
            as.data.frame()

        # Add columns for celltypes that are missing in each barcode
        col_placeholder <- setdiff(all_celltypes, colnames(df_i))
        for (c in col_placeholder) {
            df_i[[c]] <- 0
        }

        # Reorder columns
        df_i <- df_i[, all_celltypes]
        return(df_i) 
        }) %>%
        bind_rows()

    rownames(weights_rctd) <- coords_rctd
    return(weights_rctd) 
    }) %>%
    bind_rows()

# Reorder rows to have them aligned with those in the Seurat obj
weights_df <- weights_df[colnames(seurat),]

# Update the Seurat obj to include deconvolution results
seurat@meta.data <- cbind(seurat@meta.data, weights_df)
```

### Proportion piescatter plot {.tabset}

```{r vis_rctd_multi, results='asis'}

# --------------------------------------------------------------------------------
# This chunk preps plots the proportion of cell subtypes computed using the multi-node RCTD
# --------------------------------------------------------------------------------

for (name in Images(seurat)) {
    cat('####', name, '\n\n')

    # Prep input data frame for coordinates
    coords <- coords_list[[name]]

    # Extract barcodes for the current image
    barcode_set <- rownames(coords)

    # Sum all glial proportions if specified and subset barcodes
    glial_cols <- all_celltypes[!str_detect(all_celltypes, "Neu")]
    neuronal_cols <- all_celltypes[str_detect(all_celltypes, "Neu")]
    prop_df <- if (collapse_glia) {
        weights_df[barcode_set, 'Glia'] <- weights_df[barcode_set, glial_cols] %>%
            rowSums()
        weights_df[barcode_set, c(neuronal_cols, 'Glia')]
    } else {
        weights_df[barcode_set, ]
    }

    # Add proportions to the coordinate data frame
    coords <- cbind(coords, prop_df)
    coords_list[[name]] <- coords

    # Plot celltype proportions by coordinate
    p <- ggplot() +
        scatterpie::geom_scatterpie(
            data=coords,
            cols=colnames(prop_df),
            color=NA,
            aes(x=x, y=y, r=1)) +
        coord_equal() +
        custom_theme() +
        coord_flip() + 
        scale_fill_manual(values=plot_cols) +
        scale_x_reverse()

    print(p)

    # Save and link
    p_path <- file.path(
        out_dir,
        paste0(name, '_name_proportion.svg'))
    ggsave(p_path, p, device="svg")
    link_plot(p_path)
}
```

### Density heatmap {.tabset}

Here, we visualize the proportion of individual cell subtypes ranging from 0 to 1.

```{r vis_rctd_individual, results='asis'}

# --------------------------------------------------------------------------------
# This chunk preps plots the proportion of cell subtypes computed using the multi-node RCTD
# --------------------------------------------------------------------------------

for (name in Images(seurat)) {

    cat('####', name, '{.tabset}\n\n')

    # Extract the coordinate data frame for each celltype
    coords <- coords_list[[name]]

    for (celltype in all_celltypes) {

        # Prep input data frame
        cat('#####', celltype, '{.tabset}\n\n')
        # Filter a single celltype
        coords_i <- coords[, c('x', 'y', celltype)]
        colnames(coords_i)[3] <- 'proportion'

        for (prop in c('fixed', 'adjusted')) {

            cat('######', prop, 'scale\n\n')
            # Plot celltype proportions by coordinate
            p <- ggplot(coords_i, aes(x=x, y=y, color=proportion)) +
                geom_point(size=0.5) +
                custom_theme() + 
                coord_flip() + 
                scale_x_reverse() 
            p  <- if (prop == "fixed") {
                p + scale_color_gradient(low='yellow', high='red', limits=c(0, 1))
            } else {
                p + scale_color_gradient(low='yellow', high='red')
            }

            print(p)

            # Save and link
            p_path <- file.path(
                out_dir,
                paste0(name, '_', celltype, '_', prop, '.svg'))
            ggsave(p_path, p, device="svg")
            link_plot(p_path)
        }
    }
}
```


# Updated Seurat obj

```{r save}

# --------------------------------------------------------------------------------
# This chunk saves the Visium Seurat obj with RCTD-transferred annotation
# --------------------------------------------------------------------------------

out_rds <- file.path(out_dir, "updated_seurat_visium.rds")
saveRDS(seurat, out_rds, compress=FALSE)

print(paste("Spatial obj saved:", out_rds))
```


# Session info

```{r session_info, collapse=FALSE}
sessionInfo()
```

