---
title: "Visium spatial transcriptomics"
author: "Mira Sohn"
output:
    html_document:
        code_folding: hide
        df_print: paged
        toc: true
        toc_float: true
        toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE,
                      message=FALSE,
                      cache.lazy=FALSE)
```

Last run: `r date()`

This workflow is designed for exploratory analysis on the Visium dataset from the human thalamus 
([Schulmann et al.](https://www.biorxiv.org/content/10.1101/2024.09.03.611112v1)).

```{r libraries}
# Libraries
library(parallel)
library(tidyverse)
library(ggplot2)
library(plotly)
library(Seurat)
library(patchwork)
library(sceasy)
library(reticulate)
```

```{r additional_options}
# Additional options
set.seed(2570)
source('helpers.R')
use_condaenv("../../menv")

# Specify the number of available cores
n_cores <- future::availableCores()
options(future.globals.maxSize=500000 * 1024^2,   # 500G (1000 * 1024^2 = 1G)
        mc.cores=n_cores,
        future.rng.onMisuse="ignore")

# Import scanpy in R
sc <- import("scanpy")
```

```{python ppackages}
import anndata as ad
import scanpy as sc
import pandas as pd
import numpy as np
import os
```

```{r config}

# --------------------------------------------------------------------------------
# This chunk specifies input paths and variables in R
# --------------------------------------------------------------------------------

# Paths to input datasets
input_path <- "../../input/Schulmann_visium/Visium_thal_seurat.rds"
h5ad_list <- list(
    InNeuSOX14neg="../../input/thalamus_neurons_reannotated/InNeuSOX14neg_adata.h5ad",
    InNeuSOX14pos="../../input/thalamus_neurons_reannotated/InNeuSOX14pos_adata.h5ad",
    ExNeuRNF220neg="../../input/thalamus_neurons_reannotated/ExNeuRNF220neg_adata.h5ad",
    ExNeuRNF220pos="../../input/thalamus_neurons_reannotated/ExNeuRNF220pos_adata.h5ad"
)

# Path to output directory
out_dir <- "visium-thalamus"

# Specify genes of interest
gene_list <- list(
    set1=c('SYT1', 'GAD2', 'SOX14'),
    set2=c('SYT1', 'SLC17A6', 'RNF220')
)

# Number of HVGs
hvgs <- 3000

# Annotation column name
celltype_col <- "annotation_04142026"

# Create directories if missing
for (p in c(out_dir)) {
    if (! dir.exists(p)) { dir.create(p, recursive=TRUE) }
}
```

# Loading input data {.tabset}

```{r import_dataset}

# --------------------------------------------------------------------------------
# This chunk imports the input dataset
# --------------------------------------------------------------------------------

# Import visium dataset
seurat <- readRDS(input_path)

print(paste("Input file path:", input_path))
print(seurat)
```

# Genes of interest {.tabset}

The following genes of interest are explored across the image set. Gene not present
in the current dataset are labeled as 'absent genes'.

```{r genes_of_interest}

# --------------------------------------------------------------------------------
# This chunk cleans genes of interest
# --------------------------------------------------------------------------------

# Retrieve the normalized count matrix
sct_mat <- GetAssayData(seurat, assay="SCT", layer="data")

# Retrieve all gene symbols
all_genes <- rownames(sct_mat)

# Remove genes of interest if absent and print
print("Genes of interest:")
for (name in names(gene_list)) {
    genes_set <- gene_list[[name]]
    genes_absent <- genes_set[!genes_set %in% all_genes]
    genes_present <- setdiff(genes_set, genes_absent)
    gene_list[[name]] <- genes_present
    print(paste0("- ", name, ": ", paste0(genes_present, collapse=", ")))
    print(paste("      (absent genes: ", paste0(genes_absent, collapse=", "), ")"))
}
```

Following the methodology of [Schulmann et al.](https://www.biorxiv.org/content/10.1101/2024.09.03.611112v1),
we utilize normalized expression data generated via the SCTransform method 
([Hafemeister et al.](https://link.springer.com/article/10.1186/s13059-019-1874-1)).

```{r spatial_featureplot, results='asis'}

# --------------------------------------------------------------------------------
# This chunk explores the spatial expression of genes of interest
# --------------------------------------------------------------------------------

for (name in names(gene_list)) {
    cat('##', name, '{.tabset}\n\n')
    for (gene in gene_list[[name]]) {
        cat('###', gene, '\n\n')
        p <- SpatialFeaturePlot(
            seurat,
            features=gene,
            pt.size.factor=3.5)
        print(p)
        plot_path <- file.path(
            out_dir,
            paste0(name, '_', gene, '_spatialfeature.pdf'))
        ggsave(plot_path,
               p,
               device="pdf")
        link_plot(plot_path)
        cat('\n\n')
    }
}
```

# Deconvolution

We deconvolve spot-level cell subtypes using 
the [Robust Cell Type Decomposition (RCTD, Cable et al.)](https://www.nature.com/articles/s41587-021-00830-w).
This analysis is performed using raw counts from annotated in-house datasets listed below:

## Load in-house single-cell datasets

```{python load_inhouse}

# --------------------------------------------------------------------------------
# This chunk loads in-house AnnData obj
# --------------------------------------------------------------------------------

# Import input in-house AnnData obj
adatas = {name: sc.read_h5ad(path) for name, path in r.h5ad_list.items()}

print("Input in-house dataset loaded and concatenated:")
for name, path in r.h5ad_list.items():
    print(name, "-", path)

# Concatenate all obj
adata = ad.concat(
    list(adatas.values()),
    keys=list(adatas.keys()),
    index_unique="_",
    label="AnnData",
    join="outer",
    merge="same")

print(adata)

# Prep features and barcodes as a list for the next step
all_features = list(adata.var_names)
all_barcodes = list(adata.obs.index)
```

To transfer annotations from the in-house dataset to the Visium dataset, the in-house dataset is converted
into a Seurat obj.

```{r adata_to_seurat, cache=TRUE}

# --------------------------------------------------------------------------------
# This chunk converts the AnnData obj to Seurat and normalize
# --------------------------------------------------------------------------------

# Prep an input count matrix by extracting from the concatenated AnnData
ad_mat <- t(py$adata[['layers']][['counts']])
rownames(ad_mat) <- py$all_features
colnames(ad_mat) <- py$all_barcodes

# Build a new Seurat obj using raw counts
ad_seurat <- CreateSeuratObject(
    counts=ad_mat,
    meta.data=py$adata[['obs']]
) %>%
# Filter nonzero-count cells
subset(nCount_RNA > 0)

print(ad_seurat)
```

```{r prep_annotated_input}

# Specify ID col
Idents(ad_seurat) <- celltype_col

# Prep raw counts
counts <- ad_seurat[['RNA']]$counts

# Prep annotation
cluster <- as.factor(ad_seurat@meta.data[[celltype_col]])
names(cluster) <- colnames(ad_seurat)

# Prep nUMI
nUMI <-ad_seurat@meta.data[['nCount_RNA']]
names(nUMI) <- colnames(ad_seurat)

reference <- Reference(counts, cluster, nUMI)
```

```{r dotplots, results='asis', eval=FALSE}

# Prep a list of normalized in-house seurat obj
ad_seurat_list <- list(
    # SCTransform
    SCT=SCTransform(ad_seurat, variable.features.n=hvgs),
    # CPM-normalize
    CPM=LogNormalize(ad_seurat, normalization.method="RC", scale.factor=1e6)
)

print(ad_seurat_list)

for (name in names(ad_seurat_list)) {
    cat('###', name, 'normalization\n\n')
    p <- DotPlot(
        ad_seurat_list[[name]],
        group.by=celltype_col,
        features=unique(unlist(gene_list))) +
        theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
    primt(p)
    cat('\n\n')
}
```


# Session info

```{r session_info, collapse=FALSE}
sessionInfo()
```



