---
title: "Visium spatial transcriptomics"
author: "Mira Sohn"
output:
    html_document:
        code_folding: hide
        df_print: paged
        toc: true
        toc_float: true
        toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE,
                      message=FALSE,
                      echo=FALSE,
                      cache.lazy=FALSE)
```

Last run: `r date()`

This workflow is designed for exploratory analysis on the Visium dataset from the human thalamus 
([Schulmann et al.](https://www.biorxiv.org/content/10.1101/2024.09.03.611112v1)).

```{r libraries}
# Libraries
library(parallel)
library(tidyverse)
library(ggplot2)
library(plotly)
library(Seurat)
library(patchwork)
library(sceasy)
library(reticulate)
library(spacexr)
library(SummarizedExperiment)
library(SpatialExperiment)
```

```{r additional_options}
# Additional options
set.seed(2570)
source('helpers.R')
use_condaenv("../../menv")

# Specify the number of available cores
n_cores <- future::availableCores()
options(future.globals.maxSize=500000 * 1024^2,   # 500G (1000 * 1024^2 = 1G)
        mc.cores=n_cores,
        future.rng.onMisuse="ignore")

# Import scanpy in R
sc <- import("scanpy")
```

```{python ppackages}
import anndata as ad
import scanpy as sc
import pandas as pd
import numpy as np
import random
import os

random.seed(1516)
```

```{r config}

# --------------------------------------------------------------------------------
# This chunk specifies input paths and variables in R
# --------------------------------------------------------------------------------

# Paths to input datasets
input_path <- "../../input/Schulmann_visium/Visium_thal_seurat.rds"
input_h5ad <- "../../input/Schulmann_visium/combined_reclustered.h5ad"

# Path to output directory
out_dir <- "visium-thalamus"

# Column names for cell subtypes
celltype_col <- "annotation_06012026"

# Filter neurons?
filter_neurons <- TRUE

# Column name for disease conditions
selection_col <- "general_disease"

# Filter specific barcodes based on the `selection_col` column in the metadata?
# Set to NULL to disable
barcode_selection <- NULL

# Upper bound of cell number. Disable by setting it to NULL
ncell_upper_bound <- 12000

# Specify genes of interest
gene_set <- c(
    "SYT1", "GAD2", "SOX14", "SLC17A6", "RNF220", "LAMP5", 
    "BCL11B", "TUBB3", "GRN", "ATP1B1", "HSP90AA1", "HSP90AB1",
    "NEFL", "CLU", "GAPDH", "PSAP", "SERINC1", "TUBA1B", 
    "PRNP", "MEG3", "NRG3", "MALAT1", "KCNIP4", "NALF1", 
    "PTPRD", "ROBO1", "MAGI2", "FTX", "LRP1B", "CALB1", 
    "NR2F2", "ERBB4", "PVALB", "CALB2", "MEIS2"
)

# Specify colors for plots
plot_colors <- c('#F4F1DE', '#E07A5F', '#BDB76B', '#4ecdc4')

# Create directories if missing
for (p in c(out_dir)) {
    if (! dir.exists(p)) { dir.create(p, recursive=TRUE) }
}
```

# Loading input data

## Single-cell obj

```{r import_dataset}

# --------------------------------------------------------------------------------
# This chunk imports the input dataset
# --------------------------------------------------------------------------------

# Import visium dataset
seurat <- readRDS(input_path)

print(paste("Input file path:", input_path))
print(seurat)
```

## Staining images

```{r hne_images, results='asis'}
p <- SpatialFeaturePlot(
    seurat,
    features="nCount_Spatial",
    alpha=c(0, 0)) + 
    theme(legend.position="none")
hne_svg <- file.path(out_dir, "HnE_image.svg")
ggsave(hne_svg, p, device="svg")
print(p)
link_plot(hne_svg)
```

# Genes of interest {.tabset}

The following genes of interest are explored across the image set. Gene not present
in the current dataset are labeled as 'absent genes'.

```{r genes_of_interest}

# --------------------------------------------------------------------------------
# This chunk cleans genes of interest
# --------------------------------------------------------------------------------

# Retrieve the normalized count matrix
sct_mat <- GetAssayData(seurat, assay="SCT", layer="data")

# Retrieve all gene symbols
all_genes <- rownames(sct_mat)

# Remove genes of interest if absent and print
genes_absent <- gene_set[!gene_set %in% all_genes]
genes_present <- setdiff(gene_set, genes_absent)
print(paste0("- Genes of interest: ", paste0(gene_set, collapse=", ")))
print(paste0("- Genes present: ", paste0(genes_present, collapse=", ")))
print(paste0("- Genes absent: ", paste0(genes_absent, collapse=", ")))

# Clean gene_set to exclude absent genes
gene_set <- intersect(gene_set, genes_present)
```

Following the methodology of [Schulmann et al.](https://www.biorxiv.org/content/10.1101/2024.09.03.611112v1),
we utilize normalized expression data generated via the SCTransform method 
([Hafemeister et al.](https://link.springer.com/article/10.1186/s13059-019-1874-1)).

```{r spatial_featureplot, results='asis'}

# --------------------------------------------------------------------------------
# This chunk explores the spatial expression of genes of interest
# --------------------------------------------------------------------------------

for (gene in gene_set) {
    cat('##', gene, '\n\n')
    p <- SpatialFeaturePlot(
        seurat,
        features=gene,
        pt.size.factor=3.5)
    print(p)
    plot_path <- file.path(
        out_dir,
        paste0(gene, '_spatialfeature.svg'))
    ggsave(plot_path,
           p,
           device="svg")
    link_plot(plot_path)
    cat('\n\n')
}
```

# Deconvolution

We deconvolve spot-level cell subtypes using 
the [Robust Cell Type Decomposition (RCTD, Cable et al.)](https://www.nature.com/articles/s41587-021-00830-w).
This analysis is performed using raw counts from annotated datasets listed below:

## Load annotated single-cell datasets

```{python load_inhouse}

# --------------------------------------------------------------------------------
# This chunk loads annotated AnnData obj
# --------------------------------------------------------------------------------

# Import input annotated AnnData obj
adata = sc.read_h5ad(r.input_h5ad)

print("Input annotated dataset loaded and concatenated:", r.input_h5ad)
print(adata)

# Prep features and barcodes as a list for the next step
all_features = list(adata.var_names)
all_barcodes = list(adata.obs.index)
```

All input cells:

```{r input_cell_counts, results='asis'}
py$adata[['obs']][[celltype_col]] %>%
    table %>%
    as.data.frame() %>%
    set_names(c('celltype', 'nCells')) %>%
    knitr::kable()
```

```{python filter_neurons}

# --------------------------------------------------------------------------------
# This chunk optionally filter neurons
# --------------------------------------------------------------------------------

# Filter neurons if specified
if r.filter_neurons:
    neuron_mask = adata.obs[r.celltype_col].str.contains('Neu', na=False)
    adata = adata[neuron_mask, :]
    print("Input cells are narrowed to neurons.")

print(adata)
```

```{python apply_upperbound}
# reticulate::repl_python()

# Retain randomly selected N cells the most, if specified
if r.ncell_upper_bound not None:
    max_ncells = int(r.ncell_upper_bound)
    print("In each cell subtype, cells are randomly selected to retain only", max_cells, "cells.")
    obs = adata.obs.copy()
    rng = np.random.default_rng(42)
    retained_cells = []
    for celltype, idx in obs.groupby(celltype_col).groups.items():
        idx = np.array(list(idx))
        if len(idx) > max_ncells:
            keep = rng.choice(idx, size=max_ncells, replace=False)
        else:
            keep = idx
        print(celltype, "---- from", len(idx), "to", len(keep))
        retained_cells.extend(keep)

    adata = adata[retained_cells]

```

```{r filtered_cell_counts, results='asis'}
if (filter_neurons) {
    py$adata[['obs']][[celltype_col]] %>%
        table %>%
        as.data.frame() %>%
        set_names(c('celltype', 'nCells')) %>%
        knitr::kable() %>%
        print()
}
```

To transfer annotations from the annotated dataset to the Visium dataset, the annotated dataset is converted
into a Seurat obj.

