---
title: "Visium spatial transcriptomics"
author: "Mira Sohn"
output:
    html_document:
        code_folding: hide
        df_print: paged
        toc: true
        toc_float: true
        toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE,
                      message=FALSE,
                      echo=FALSE,
                      cache.lazy=FALSE)
```

Last run: `r date()`

This workflow is designed for exploratory analysis on the Visium dataset from the human thalamus 
([Schulmann et al.](https://www.biorxiv.org/content/10.1101/2024.09.03.611112v1)).

```{r libraries}
# Libraries
library(parallel)
library(tidyverse)
library(ggplot2)
library(plotly)
library(Seurat)
library(patchwork)
library(sceasy)
library(reticulate)
library(spacexr)
library(SummarizedExperiment)
library(SpatialExperiment)
```

```{r additional_options}
# Additional options
set.seed(2570)
source('helpers.R')
use_condaenv("../../menv")

# Specify the number of available cores
n_cores <- future::availableCores()
options(future.globals.maxSize=500000 * 1024^2,   # 500G (1000 * 1024^2 = 1G)
        mc.cores=n_cores,
        future.rng.onMisuse="ignore")

# Import scanpy in R
sc <- import("scanpy")
```

```{python ppackages}
import anndata as ad
import scanpy as sc
import pandas as pd
import numpy as np
import os
```

```{r config}

# --------------------------------------------------------------------------------
# This chunk specifies input paths and variables in R
# --------------------------------------------------------------------------------

# Paths to input datasets
input_path <- "../../input/Schulmann_visium/Visium_thal_seurat.rds"
h5ad_list <- list(
    InNeuSOX14neg="../../input/thalamus_neurons_reannotated/InNeuSOX14neg_adata.h5ad",
    InNeuSOX14pos="../../input/thalamus_neurons_reannotated/InNeuSOX14pos_adata.h5ad",
    ExNeuRNF220neg="../../input/thalamus_neurons_reannotated/ExNeuRNF220neg_adata.h5ad",
    ExNeuRNF220pos="../../input/thalamus_neurons_reannotated/ExNeuRNF220pos_adata.h5ad"
)

# Path to output directory
out_dir <- "visium-thalamus"

# Specify genes of interest
gene_list <- list(
    set1=c('SYT1', 'GAD2', 'SOX14'),
    set2=c('SYT1', 'SLC17A6', 'RNF220')
)

# Number of HVGs
hvgs <- 3000

# Annotation column name
celltype_col <- "annotation_04142026"

# Create directories if missing
for (p in c(out_dir)) {
    if (! dir.exists(p)) { dir.create(p, recursive=TRUE) }
}
```

# Loading input data {.tabset}

```{r import_dataset}

# --------------------------------------------------------------------------------
# This chunk imports the input dataset
# --------------------------------------------------------------------------------

# Import visium dataset
seurat <- readRDS(input_path)

print(paste("Input file path:", input_path))
print(seurat)
```

# Genes of interest {.tabset}

The following genes of interest are explored across the image set. Gene not present
in the current dataset are labeled as 'absent genes'.

```{r genes_of_interest}

# --------------------------------------------------------------------------------
# This chunk cleans genes of interest
# --------------------------------------------------------------------------------

# Retrieve the normalized count matrix
sct_mat <- GetAssayData(seurat, assay="SCT", layer="data")

# Retrieve all gene symbols
all_genes <- rownames(sct_mat)

# Remove genes of interest if absent and print
gene_set <- unique(unlist(gene_list))
genes_absent <- gene_set[!gene_set %in% all_genes]
genes_present <- setdiff(gene_set, genes_absent)
print(paste0("- Genes of interest: ", paste0(gene_set, collapse=", ")))
print(paste0("- Genes present: ", paste0(genes_present, collapse=", ")))
print(paste0("- Genes absent: ", paste0(genes_absent, collapse=", ")))
```

Following the methodology of [Schulmann et al.](https://www.biorxiv.org/content/10.1101/2024.09.03.611112v1),
we utilize normalized expression data generated via the SCTransform method 
([Hafemeister et al.](https://link.springer.com/article/10.1186/s13059-019-1874-1)).

```{r spatial_featureplot, results='asis'}

# --------------------------------------------------------------------------------
# This chunk explores the spatial expression of genes of interest
# --------------------------------------------------------------------------------

for (gene in gene_set) {
    cat('##', gene, '\n\n')
    p <- SpatialFeaturePlot(
        seurat,
        features=gene,
        pt.size.factor=3.5)
    print(p)
    plot_path <- file.path(
        out_dir,
        paste0(gene, '_spatialfeature.pdf'))
    ggsave(plot_path,
           p,
           device="pdf")
    link_plot(plot_path)
    cat('\n\n')
}
```

# Deconvolution

We deconvolve spot-level cell subtypes using 
the [Robust Cell Type Decomposition (RCTD, Cable et al.)](https://www.nature.com/articles/s41587-021-00830-w).
This analysis is performed using raw counts from annotated datasets listed below:

## Load annotated single-cell datasets

```{python load_inhouse}

# --------------------------------------------------------------------------------
# This chunk loads annotated AnnData obj
# --------------------------------------------------------------------------------

# Import input annotated AnnData obj
adatas = {name: sc.read_h5ad(path) for name, path in r.h5ad_list.items()}

print("Input annotated dataset loaded and concatenated:")
for name, path in r.h5ad_list.items():
    print(name, "-", path)

# Concatenate all obj
adata = ad.concat(
    list(adatas.values()),
    keys=list(adatas.keys()),
    index_unique="_",
    label="AnnData",
    join="outer",
    merge="same")

print(adata)

# Prep features and barcodes as a list for the next step
all_features = list(adata.var_names)
all_barcodes = list(adata.obs.index)
```

To transfer annotations from the annotated dataset to the Visium dataset, the annotated dataset is converted
into a Seurat obj.

## Annotation transfer - doublet mode {.tabset}

Annotation transfer with the doublet mode option computes in the following two cell subtypes:

- **first_type**: first predicted celltype
- **second_type**: second predicted celltype, if exists

```{r adata_to_seurat, cache=TRUE}

# --------------------------------------------------------------------------------
# This chunk converts the AnnData obj to Seurat and normalize
# --------------------------------------------------------------------------------

# Prep an input count matrix by extracting from the concatenated AnnData
ad_mat <- t(py$adata[['layers']][['counts']])
rownames(ad_mat) <- py$all_features
colnames(ad_mat) <- py$all_barcodes

# Build a new Seurat obj using raw counts
ad_seurat <- CreateSeuratObject(
    counts=ad_mat,
    meta.data=py$adata[['obs']]
) %>%
# Filter nonzero-count cells
subset(nCount_RNA > 0)

print(ad_seurat)
```

```{r prep_rctd, cache=TRUE, dependson='adata_to_seurat'}

# --------------------------------------------------------------------------------
# This chunk preps input for RCTD
# --------------------------------------------------------------------------------

# Specify ID col
Idents(ad_seurat) <- celltype_col

# Prep raw counts from the annotated dataset
counts <- ad_seurat[['RNA']]$counts

# Prep annotation from the annotated dataset
cluster <- as.factor(ad_seurat@meta.data[[celltype_col]])
names(cluster) <- colnames(ad_seurat)

# Prep nUMI from the annotated dataset
nUMI <-ad_seurat@meta.data[['nCount_RNA']]
names(nUMI) <- colnames(ad_seurat)

# Create a new `Reference`
reference <- Reference(counts, cluster, nUMI)

# Prep a list of query across images
query_list <- mclapply(1:length(Images(seurat)), function(i) {
    # Extract raw counts per image
    ct <- GetAssayData(
        seurat,
        assay='Spatial',
        layer=Layers(seurat[['Spatial']])[i])
    # Extract coordinates per image
    cd <- GetTissueCoordinates(
        seurat,
        image=Images(seurat)[i])
    # Clean column names for the coordinate table
    colnames(cd) <- c('x', 'y')
    cd[is.na(colnames(cd))] <- NULL
    # Create a new spatial obj
    obj <- SpatialRNA(
        coords=cd,
        counts=ct,
        colSums(ct))
    return(obj)
}) %>%
# Name the list
set_names(Images(seurat))
```

```{r run_rctd, cache=TRUE, dependson='prep_rctd'}

# --------------------------------------------------------------------------------
# This chunk runs RCTD
# --------------------------------------------------------------------------------

# Create a list of RCTD obj
rctd_list <- lapply(
    names(query_list), function(name) {
        # Create a new RCTD obj
        create.RCTD(query_list[[name]], reference, max_cores=n_cores) %>%
            # Run RCTD
            run.RCTD(doublet_mode="doublet")
    }) %>%
    set_names(names(query_list))
```

```{r vis_rctd_doublet, results='asis'}

# --------------------------------------------------------------------------------
# This chunk visualizes RCTD results from the doublet mode
# --------------------------------------------------------------------------------

# Add RCTD results to the metadata of Visium obj
rctd_df <- map_df(rctd_list, ~.x@results$results_df)

# If uncertain, replace the annotation of the second type with "Undetermined"
anno_cols <- c('first_type', 'second_type')
anno_levels <- c(unique(ad_seurat@meta.data[[celltype_col]]), "Undetermined")
for (anno in anno_cols) {
    rctd_df[[anno]] <- factor(rctd_df[[anno]], levels=anno_levels)
}
rctd_df[rctd_df[['spot_class']] == "doublet_uncertain", 'second_type'] <- "Undetermined"

# Add the transferred annotation to the metadata of Visium seurat obj
seurat <- AddMetaData(seurat, metadata=rctd_df)

# Plot and save
for (name in Images(seurat)) {
    cat('###', name, '\n\n')
    p_list <- mclapply(c('first_type', 'second_type'), function(anno) {
        SpatialDimPlot(seurat,
                       group.by=anno,
                       pt.size.factor=3.5,
                       images=name) +
        theme(legend.margin=margin(0, 0, 0, 0), 
              legend.title=element_text(size=8),
              legend.text=element_text(size=8))
    })
    p <- p_list[[1]] | p_list[[2]]
    print(p)
    p_path <- file.path(out_dir, paste0(name, '_annotated.pdf'))
    ggsave(p_path, p, device="pdf")
    link_plot(p_path)
}
```

## Annotation transfer - celltype proportions {.tabset}

Since two cell subtypes do not adequately represent all transferred celltypes, we assess the proportion of 
cell subtypes across the annotated celltypes.

```{r run_rctd_multi, cache=TRUE, dependson='run_rctd'}

# --------------------------------------------------------------------------------
# This chunk runs multi-mode RCTD
# --------------------------------------------------------------------------------

# Create a list of RCTD obj
multi_list <- lapply(
    names(query_list), function(name) {
        # Create a new RCTD obj
        create.RCTD(query_list[[name]], reference, max_cores=n_cores) %>%
            # Run RCTD
            run.RCTD(doublet_mode="multi")
    }) %>%
    set_names(names(query_list))
```

```{r save, eval=FALSE}

# --------------------------------------------------------------------------------
# This chunk saves the Visium Seurat obj with RCTD-transferred annotation
# --------------------------------------------------------------------------------

out_rds <- file.path(out_dir, "updated_seurat_visium.rds")
saveRDS(seurat, out_rds)

print(paste("Spatial obj saved:", out_rds))
```


# Session info

```{r session_info, collapse=FALSE}
sessionInfo()
```

