---
title: "Visium spatial transcriptomics"
author: "Mira Sohn"
output:
    html_document:
        code_folding: hide
        df_print: paged
        toc: true
        toc_float: true
        toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE,
                      message=FALSE,
                      cache.lazy=FALSE)
```

Last run: `r date()`

This workflow is designed for exploratory analysis on the Visium dataset from the human thalamus 
([Schulmann et al.](https://www.biorxiv.org/content/10.1101/2024.09.03.611112v1)).

```{r libraries}
# Libraries
library(parallel)
library(tidyverse)
library(ggplot2)
library(plotly)
library(Seurat)
library(patchwork)

# Additional options
set.seed(2570)
source('helpers.R')

# Specify the number of available cores
n_cores <- future::availableCores()
options(mc.cores=n_cores)
```

```{r config}

# --------------------------------------------------------------------------------
# This chunk specifies input paths and variables in R
# --------------------------------------------------------------------------------

# Paths to input datasets
input_path <- "../../input/Schulmann_visium/Visium_thal_seurat.rds"
ref_path <- ""

# Path to output directory
out_dir <- "visium-thalamus"

# Specify genes of interest
gene_list <- list(
    set1=c('SYT1', 'GAD2', 'SOX14'),
    set2=c('SYT1', 'SLC17A6', 'RNF220')
)

# Create directories if missing
for (p in c(out_dir)) {
    if (! dir.exists(p)) { dir.create(p, recursive=TRUE) }
}
```

# Loading input data {.tabset}

```{r import_dataset}

# --------------------------------------------------------------------------------
# This chunk imports the input dataset
# --------------------------------------------------------------------------------

# Import visium dataset
seurat <- readRDS(input_path)

print(paste("Input file path:", input_path))
print(seurat)
```

# Genes of interest {.tabset}

The following genes of interest are explored across the image set. Gene not present
in the current dataset are labeled as 'absent genes'.

```{r genes_of_interest}

# --------------------------------------------------------------------------------
# This chunk cleans genes of interest
# --------------------------------------------------------------------------------

# Retrieve the normalized count matrix
sct_mat <- GetAssayData(seurat, assay="SCT", layer="data")

# Retrieve all gene symbols
all_genes <- rownames(sct_mat)

# Remove genes of interest if absent and print
print("Genes of interest:")
for (name in names(gene_list)) {
    genes_set <- gene_list[[name]]
    genes_absent <- genes_set[!genes_set %in% all_genes]
    genes_present <- setdiff(genes_set, genes_absent)
    gene_list[[name]] <- genes_present
    print(paste0("- ", name, ": ", paste0(genes_present, collapse=", ")))
    print(paste("      (absent genes: ", paste0(genes_absent, collapse=", "), ")"))
}
```

Following the methodology of [Schulmann et al.](https://www.biorxiv.org/content/10.1101/2024.09.03.611112v1),
we utilize normalized expression data generated via the SCTransform method 
([Hafemeister et al.](https://link.springer.com/article/10.1186/s13059-019-1874-1)).

```{r spatial_featureplot, results='asis'}

# --------------------------------------------------------------------------------
# This chunk explores the spatial expression of genes of interest
# --------------------------------------------------------------------------------

for (name in names(gene_list)) {
    cat('##', name, '{.tabset}\n\n')
    for (gene in gene_list[[name]]) {
        cat('###', gene, '\n\n')
        p <- SpatialFeaturePlot(
            seurat,
            features=gene,
            pt.size.factor=3.5)
        print(p)
        plot_path <- file.path(
            out_dir,
            paste0(name, '_', gene, '_spatialfeature.pdf'))
        ggsave(plot_path,
               p,
               device="pdf")
        link_plot(plot_path)
        cat('\n\n')
    }
}
```

# Deconvolution

```{r}

```

# Session info

```{r session_info, collapse=FALSE}
sessionInfo()
```



