---
title: "Pseudobulk differential analysis on cryptic exons (CEs)"
author: "Mira Sohn"
output:
    html_document:
        code_folding: hide
        df_print: paged
        toc: true
        toc_float: true
        toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE,
                      message=FALSE,
                      cache.lazy=FALSE)
```

Last run: `r date()`

This workflow is designed to conduct pseudobulk differential analysis
on splice junctions with and without CEs.

```{r libraries}
# Libraries
library(tidyverse)
library(ggplot2)
library(ggbeeswarm)
library(parallel)
library(BiocParallel)
library(yaml)
library(RColorBrewer)
library(plotly)
library(parallel)
library(rtracklayer)
library(GenomicRanges)
library(GenomicFeatures)
library(zinbwave)
library(scran)

# Additional options
set.seed(2570)
source('../config/helpers.R')

# Specify the number of available cores
n_cores <- future::availableCores()
options(mc.cores=n_cores)
```

```{r config}

# --------------------------------------------------------------------------------
# This chunk specifies input paths and variables in R
# --------------------------------------------------------------------------------

# Path to input matrices
in_rds <- "sc-exploratory/aggr_matrices.rds"

# Path to Snakemake config
snakemake_config <- "../config/config.yaml"

# Paths to yaml config
config_yaml <- read_yaml(snakemake_config)

# Path to output directory
out_dir <- "sc-pseudobulk"

# Path to GTF
gtf_path <- "../../../input/thalamus_excitatory/genes.gtf.gz"

# Create directories if missing
for (p in c(out_dir)) {
    if (! dir.exists(p)) { dir.create(p, recursive=TRUE) }
}

# Specify metadata columns for factors and factor levels
factor_groups <- list(
    general_disease=c('Control', 'FTD', 'AD'),
    annotation_04142026=c( 
        'ExNeu RNF220+',
        'ExNeu RNF220-',
        'InNeu SOX14+',
        'InNeu SOX14-'),
    study=c('Marsan', 'Biogen', 'Mathys'),
    study_specific_disease_specific=c(
        'Control-Marsan',
        'Control-Biogen',
        'Control-Mathys',
        'FTD-Marsan',
        'FTD-Biogen',
        'AD-Mathys')
)

# Specify HEX color code
color_map <- list(
    general_disease=c(
        # `Control-FTD`="#9999FF",
        # `Control-AD`="#6666FF",
        Control="#6666FF",
        FTD="#669900",
        AD="#FF6600"),
    study=c(
        Marsan="#FF00FF",
        Biogen="#00FFFF",
        Mathys="#BFBFBF"),
    cell_type=c(
        Astrocyte="#ABD158",
        BVC="#74858C",
        CP="#D5C18E",
        ExNeu1="#90C6E2",
        ExNeu2="#547CB2",
        InNeu1="#21409A",
        InNeu2="#262262",
        Lymphocyte="#68151F",
        Microglia="#D24A75",
        OL="#FDB515",
        OPC="#ED7020",
        PVM="#231F20")
)

# Specify column names to subset barcodes
# NOTE: Set to NULL if unnecessary
id_col <- 'SampleID'
subset_col <- 'study_celltype'
sample_col <- 'SampleID_celltype'
celltype_col <- 'annotation_04142026'
disease_col <- "general_disease"
study_col <- 'study'

# Specify CE target genes
ce_targets <- config_yaml[['genes']]

# Pseudocount factor
# NOTE: this number is multiplied by the minimum number of non-zero 
#       junction counts when calculating ratios between cryptic and non-cryptic,
#       as well as between cryptic and all junctions
pseudo_p <- 0.1

# Specify the proportion of y-value to be added to the max data point
ylim_factor <- 0.3
```

# Loading input data {.tabset}

The number of splicing junctions was counted based on cellranger-aligned reads mapped to the following
genes:

```{r genes_tested}
print(ce_targets)
```

Raw junction counts were compiled into a junction-by-barcode matrix for differential testing in the
previous analysis, [`sc-exploratory.html`](sc-exploratory.html). This matrix was preprocessed
to aggregate the counts per sample per celltype, along with their associated metadata table.
These preprocessed matrices and metadata are imported from the following location:

```{r infile_info}
print(in_rds)
```

```{r import_adata}

# --------------------------------------------------------------------------------
# This chunk imports input data saved in the previous single-cell 
# exploratory analysis. Manually edit the following code if necessary.
# --------------------------------------------------------------------------------

# Import input RDS file
obj <- readRDS(in_rds)

# Extract raw count matrices for subsets of interest
mat_list <-  map(obj[['bulk']], ~.x[['raw']])

# Extract the entire sampletable (Edit manually if necessary)
sampletable <- obj[['meta']]
sampletable[[sample_col]] <- paste0(
    sampletable[[id_col]],
    "_",
    sampletable[[celltype_col]])
sampletable[[subset_col]] <- paste0(
    sampletable[[study_col]],
    "_",
    sampletable[[celltype_col]]
)

# Ensure to have correct levels for each factor columns
for (f in names(factor_groups)) {
    sampletable[[f]] <- factor(sampletable[[f]], levels=factor_groups[[f]])
    # Break if any of the conversion introduced missing values
    if (any(is.na(sampletable[[f]]))) {
        stop(paste0("NAs introduced to the following column:", f))
    }
}

# Prep a list of metadata tables for each subset
coldata_list <- mclapply(mat_list, function(m) { 
    # Splice the metadata table based on samples in the matrix
    cdata <- sampletable[sampletable[[sample_col]] %in% colnames(m),
                         c(sample_col, subset_col, names(factor_groups))] %>%
        unique()
    # Break if columns of the matrix and rows of the metadata don't match
    if (nrow(cdata) != ncol(m)) {
        stop("Rownames of coldata and colnames of matrix should be the same!")
    }
    # Reorder rows of the subsetted metadata table
    rownames(cdata) <- cdata[[sample_col]]
    cdata <- cdata[colnames(m),]
    return(cdata)
})

```

# Subsets {.tabset}

The subset-wise aggregation was performed as summarized below:

```{r subset_overview, results='asis'}

# --------------------------------------------------------------------------------
# This chunk explores the number of samples and metadata info across the subsets
# --------------------------------------------------------------------------------

# Function to summarize the number of samples (N)
summarize_N <- function(meta_df) {

    # Count the number of samples
    n_df <- meta_df %>%
        group_by_at(vars(subset_col, disease_col)) %>%
        summarize(N=n()) %>%
        spread(disease_col, N)

    # Replace missing values with zero
    n_df[is.na(n_df)] <- 0
    # Add a new column for all N
    n_df[['all_N']] <- rowSums(n_df[, colnames(n_df)[-1]])

    return(n_df)
}

for (name in names(coldata_list)) {

    df <- coldata_list[[name]] %>%
        remove_rownames()
    cat("##", name, "{.tabset}\n\n")
    cat("### N\n\n")
    # Prep the summary table for N
    n_df <- summarize_N(df)

    # Print tables
    print(knitr::kable(n_df))
    cat('\n\n')

    cat("### Full table\n\n")
    subchunkify(paste0(name, "_subset_N_unfiltered"))
    csv_path <- file.path(out_dir,
                          paste0(name, '_unfiltered_sampletable.csv'))
    write.csv(df,
              csv_path,
              row.names=FALSE,
              quote=FALSE)

    link_table(csv_path)
    cat('\n\n')
}
```

# Pre-filtering nonzero samples {.tabset}

Samples with zero junction counts are removed from each subset. The number of samples before
and after filtering is summarized below.


```{r prefilter_samples, results='asis'}

# --------------------------------------------------------------------------------
# This chunk filters nonzero-count samples
# --------------------------------------------------------------------------------

for (name in names(mat_list)) {
    cat("##", name, "{.tabset}\n\n")
    # Remove zero-count samples from the count matrix
    m <- mat_list[[name]]
    nonzero_samples <- colSums(m) > 0
    m <- m[, nonzero_samples]
    mat_list[[name]] <- m
    # Remove zero-count samples from the metadata table
    cdata_old <- coldata_list[[name]]
    cdata_new <- cdata_old[colnames(m),]
    coldata_list[[name]] <- cdata_new
    
    cat("### N\n\n")
    n_df <- summarize_N(cdata_new)
    cat("A total of",
        sum(!nonzero_samples),
        "samples are removed from the current subset.",
        "Refer to the following table for remaining samples:")
    print(knitr::kable(n_df))
    cat("\n\n")

    cat("### Full table\n\n")
    subchunkify(paste0(name, "_subset_N_filtered"))
    csv_path <- file.path(out_dir,
                          paste0(name, '_filtered_sampletable.csv'))
    write.csv(df,
              csv_path,
              row.names=FALSE,
              quote=FALSE)

    link_table(csv_path)
    cat('\n\n')
}

```

```{r setup_obj}

# --------------------------------------------------------------------------------
# This chunk creates dds objects per subset by modeling zero counts with zinbwave
# (https://github.com/mikelove/zinbwave-deseq2/blob/master/zinbwave-deseq2.Rmd)
# --------------------------------------------------------------------------------

# Remove subsets with zero remaining samples
coldata_list <- coldata_list[map_lgl(coldata_list, function(x) nrow(x) > 0)]
mat_list <- mat_list[names(coldata_list)]

# Convert the count matrix into the SummarizedExperiment obj across the subsets
# NOTE: Features that were undetected in any of the samples are removed here.
#       This is required by zinbwave
se_list <- map(mat_list, ~.x[rowSums(.x) > 0,]) %>% 
    map2(coldata_list,
         ~SummarizedExperiment(assays=list(counts=.x), colData=.y))

# Specify a design formula
design_formula <- as.formula(paste("~", disease_col))

# Run log2(raw_counts + 1) transformation
l2t_list <- lapply(names(se_list), function(name) {
    se <- se_list[[name]]
    # varianceStabilizingTransformation(dds, blind=TRUE)
    m_raw <- assay(se, "counts")
    # Specify a pseudocount to the non-zero min count
    pseudo_c <- min(m_raw[m_raw > 0])
    log2(m_raw + pseudo_c)
}) %>%
set_names(names(se_list))
```

# Quality control (QC)

## Sample similarity heatmap {.tabset}

Sample-to-sample similarity in junction profiling is explored using a heatmap of normalized counts 
(log2 scale), with *Euclidean distance* represented by the color scale. Darker blue indicates
less distance (i.e., greater similarity) in splice junction profiling. Note that this analysis 
excluded junctions that were not detected in any of the samples.

```{r qc_heatmap, results='asis', fig.height=8, fig.width=8}

# --------------------------------------------------------------------------------
# This chunk explores sample similarity using heatmap
# --------------------------------------------------------------------------------

for (name in names(l2t_list)) {
    cat('###', name, '{.tabset}\n\n')
    # Prep a matrix
    scmat <- l2t_list[[name]] %>%
        t() %>%
        dist() %>%
        as.matrix()

    # Set color to be displayed
    colors <- colorRampPalette(rev(RColorBrewer::brewer.pal(9, 'Blues')))(255)

    # Prep metadata for heatmap
    h_meta <- coldata_list[[name]]
    h_meta <- h_meta[, c(disease_col, subset_col)]
    h_meta <- h_meta[, colnames(h_meta) != sample_col]

    # Print heatmap
    print(pheatmap::pheatmap(
        scmat,
        scale='none',
        color=colors,
        annotation_col=h_meta,
        annotation_colors=color_map,
        show_rownames=FALSE,
        show_colnames=FALSE,
        cluster_cols=TRUE,
        border_color=NA
    )
    )
    cat('\n\n')
}

```

